# Introduction {#intro}

```{r, echo=FALSE}
library(png)
library(grid)
```


## Who this book is for and how to use it

This book is for anyone who wants to learn how to *do* spatial microsimulation.
By that we mean taking individual level and geographical datasets and
processing them to solve real-world problems.

This book is a general-purpose introduction
to the field and its implementation in a modern programming language. As such
there is no single target audience, although
the book was written with the following people in mind:

- Practitioners: because of the applied nature of the method and its utility for
modelling people, it is useful in a wide range of sectors. Planning for the
future of educational, health and transport services at the local level,
for example, could benefit from spatial microdata on the inhabitants of the study area.
The future is of course uncertain and spatial microsimulation can provide tools
for scenario-based planning to explore various visions of the future.
In this context spatial microsimulation can be especially useful for designing
local policies to assist with the global transition away from
fossil fuels [@Lovelace2014-vul].

- Academics: spatial microsimulation originated in academic research, building on
the work of early pioneers such as @Deming1940 and @Holm1987. With recent advances in
computer power, open source software and data accessibility, spatial microsimulation
can now answer a greater range of questions than ever before. Because spatial
microsimulation and associated methods are still in their infancy, there are
still many areas of research where the approach has yet to be used. This
makes spatial microsimulation an ideal method for doing new research,
for example as part of a PhD.

- Educators: although spatial microsimulation has become more accessible over the
last few years, it is still out of reach for the majority of people.
Yet the potential benefits are large. This book provides example data and
code that can be used as part of a course involving the analysis of spatial
microdata, for example in a module contributing to an undergraduate or MSc
course on Transport Modelling, Spatial Economics or Quantitative Geography.

The book has a definite progression from easier content to more advanced
topics, especially those in Chapters 11 and 12, which deal with spatial microsimulation
for transport modelling and agent-based modelling (ABM), respectively.
The book can be read in order from front to back for a detailed overview
of the field, its applications and its concepts (Part I);
the practicalities and software decisions involved in generating
spatial microdata (Part II); and issues around the modelling of spatial microdata
(Part III).

Equally, the book can be used as a reference volume, to be dipped into as and when
particular topics arise. There is a path dependency in the book however: Part II
assumes that the reader is familiar with the concepts introduced in Part I, and
the two chapters in
Part III assume that the reader is competent with generating and handling spatial
microdata, as introduced in Chapters 4 and 5. These are the central chapters in
the book, in terms of generating spatial microdata with R. Chapters 4 and 5 are also the
most important for people who simply want to generate spatial microdata rapidly.
Chapter 6 provides insight into more experimental approaches for generating spatial
microdata, while Chapter 7 provides a comprehensive worked example of the methods
used 'in the wild'. A more detailed overview is provided in [](#overview).

If you can't wait to get your hands dirty with code and example data for
doing spatial microsimulation, please skip to Chapter 3.
For other readers who want a little more background on this book and
spatial microsimulation, bear with us.
The remainder of this chapter explains the thinking behind the book,
including the reasons for focussing on spatial microsimulation in R
rather than in another language and the motivations for doing spatial
microsimulation in the first place. The next chapter is practical.
Chapter 3, by contrast, brings the
reader up-to-date on what spatial microsimulation is and how it is currently
used.

\pagebreak

For the structure, each chapter begins with an introduction and ends with a
summary of what has been done in the chapter. The last section
of this chapter contains an overview of the book.


## Motivations

Imagine a world in which 
data on companies, households and governments
were widely available. Imagine, further, that researchers
and decision-makers acting in the public interest had
tools enabling them to test and *model* such data to explore different
scenarios of the future.
People would be able to
make more informed decisions, based on the best available evidence.
In this technocratic dreamland pressing
problems such as climate change, inequality and poor human
health could be solved. 

These are the types of real-world issues that we hope the methods in this
book will help to address. Spatial microsimulation can provide new
insights into complex problems and, ultimately, lead to better
decision-making. By shedding new light on existing information,
the methods can help shift decision-making processes away from
ideological bias and towards evidence-based policy.

The 'open data' movement has made many datasets more widely available.
However, the dream sketched in the opening paragraph is still far from reality.
Researchers typically must work with data that
is incomplete or inaccessible.
Available datasets often
lack the spatial or temporal resolution required to understand complex processes.
Publicly available datasets frequently miss key attributes, such as income.
Even when high quality data is made available, it can be very difficult
for others to check or *reproduce* results based on them.
Strict conditions inhibiting data access and use are aimed at protecting citizen
privacy but can also serve to block democratic and enlightened decision making.

The empowering potential of new information is encapsulated in the
saying that 'knowledge is power'.
This helps explain why methods such as spatial microsimulation,
that help represent the full complexity of reality,
are in high demand.

Spatial microsimulation is a growing approach to studying complex issues
in the social sciences.
It has been used extensively in fields as diverse
as transport, health and education (see Chapter \ref{what-is}), and
many more applications are possible.
Fundamental to the approach are approximations of
individual level data at high spatial resolution: people allocated to places.
This *spatial microdata*, in one form or another,
provides the basis for all spatial microsimulation research.

\pagebreak
The purpose of this book is to teach methods for *doing* (not reading about!)
spatial microsimulation. This involves techniques for generating and analysing
spatial microdata to get the 'best of both worlds'
from real individual and geographically-aggregated data.
*Population synthesis* is therefore a key stage in spatial microsimulation: 
generally real spatial microdata are unavailable due to concerns over data
privacy. Typically, synthetic spatial microdatasets are generated by
combining aggregated outputs from Census results with individual level data
(with little or no geographical information) 
from surveys that are representative of the population of interest. 

The resulting *spatial microdata* are useful in many
situations where individual level and geographically specific
processes are in operation. Spatial microsimulation
enables modelling and analysis on multiple levels.
Spatial microsimulation also overlaps with (and provides useful
initial conditions for) agent-based models (see Chapter 12).

Despite its utility, spatial microsimulation
is little known outside the fields of human geography and regional
science. The methods taught in this book
have the potential to be useful in a wide
range of applications. Spatial microsimulation
has great potential to be applied to new areas
for informing public policy. Work of great potential social benefit
is already being done using spatial microsimulation in
housing, transport and
sustainable urban planning. Detailed modelling
will clearly be of use for planning for a *post-carbon future*, one in
which we stop burning fossil fuels. 

For these reasons there 
is growing interest in spatial microsimulation. This is due largely to its
practical utility in an era of 'evidence-based policy' but is also driven by
changes in the wider research environment inside and outside of academia.
Continued improvements in computers, software and data availability mean
the methods are more accessible than ever. It is now possible to
simulate the populations of small administrative areas at the individual level
almost anywhere in the world. This opens new possibilities for a range of
applications, not least policy evaluation.

Still, the meaning of spatial microsimulation is ambiguous for many. 
This book also aims to clarify what the method entails in practice.
Ambiguity surrounding the term seems to arise
partly because the methods are inherently complex, operating at multiple
levels, and partly
due to researchers themselves. Some uses of the term 'spatial microsimulation'
in the academic literature
are unclear as to its meaning; there is much
inconsistency about what it means.  Worse is work that
treats spatial microsimulation as a magical black box that just 'works'
without any need to describe, or more importantly make reproducible, the
methods underlying the black box. This book is therefore also about
demystifying spatial microsimulation.

\pagebreak

## A definition of spatial microsimulation

At this early stage, it is worth considering how spatial microsimulation
has been interpreted in past work and how we define the term for the book.
Depending on the available data, the aim of the research, and the
interpretation of the researcher using the techniques,
spatial microsimulation has two broad meanings. Spatial microsimulation can be 
understood either as a technique or an approach:

1. A *method* for generating spatial microdata --- individuals allocated to
zones (see Figure 1.1) --- by combining individual and geographically aggregated
datasets. In this interpretation, 'spatial microsimulation' is roughly synonymous with
'population synthesis'.
2. An *approach* to understanding multi level phenomena based on
 spatial microdata --- simulated or real.
 
Figure 1.1 is a simplified flow diagram showing the difference between
spatial microdata (bottom) and more commonly found official data types (top).
There are clear disadvantages with both geographically aggregated and
individual level (non-geographical) sources. Spatial microsimulation
seeks to overcome some of these limitations, creating new outputs that
are useful for a number of research applications and as inputs to
more sophisticated models.

```{r, fig.cap="Schematic diagram contrasting conventional official data (above) against spatial microdata produced during *population synthesis* (below). Note that for brevity the geographically aggregated data are called *constraints* and the individual level survey data are called *microdata* in this book.", fig.scap="Schematic diagram of spatial microdata", fig.height=6, echo=FALSE}
img <- readPNG("figures/msim-flow.png")
grid.raster(img)
```

Throughout this book we use a broad definition of the term spatial microsimulation:
\index{spatial microsimulation!definition of}

> **The creation, analysis and modelling of individual level data allocated to geographic zones.**

To further clarify the distinction between the methodology and
the approach, we use term *population synthesis* to describe the narrower
process of generating the spatial microdata (see
Chapter \ref{smsimr}).
The term *spatial microsimulation* is reserved to describe the overall
process, which usually involves population synthesis. Unless you are lucky
enough to have access to real spatial microdata, a scarce but increasingly
available commodity, spatial microdata must be generated by combining
zone level and individual level data.
Throughout the course of the book the emphasis shifts:
the focus in Chapters 4 to 6 is on generating spatial microdata
while Chapters 7 onwards focus on how to use this rich data type.

## Learning by doing

Another issue tackled in this book is reproducibility, which is intimately
linked to the 'learning by doing' approach.  Notwithstanding a few exceptions
[e.g. @Williamson2007], most findings in the
field cannot be replicated: there is no way of independently
verifying the results of spatial microsimulation research.
Some model-based decisions, such as where to build smoking cessation clinics
[@Tomintz2008],
are a matter of life or death.
In such cases model transparency becomes vital.

Today fast internet connections, open access
datasets and free software make it easier than ever to create reproducible research.
This movement is growing in many fields, from Regional Science to Epidemiology
[@Ince2012;@Peng2006a;@McNutt2014;@Rey2014].

This book encourages 'Open Science' throughout.
\index{reproducibility}
Reproducibility is encouraged through the provision of code.
You will be expected to *do* spatial microsimulation rather than simply *watch*
the results that others have produced! Small datasets are
provided on which these 'code chunks' (such as that presented below) can run.
All the findings presented in this
book can therefore be reproduced using code and data in the book's GitHub
[repository](https://github.com/Robinlovelace/spatial-microsim-book).

```{r}
# This is a code chunk. Below is the code!
ind <- read.csv("data/SimpleWorld/ind-full.csv") # read-in data
print(ind[1,]) # print the 1st line of the data
```

Notice that in addition to showing what to type into R, we show what the output should be.
In fact, using the 'RMarkdown' file format, the code runs every time the book is
compiled, ensuring the code's long-term viability.
There are good reasons to ensure such a reproducible work-flow.
First, reproducibility can save time.
Reproducible code that can be re-used multiple times, meaning you only have to write it once,
reducing the need to 're-invent the wheel' for every new analysis.
Second, reproducibility can increase
research impact: by enabling and encouraging others to use your work,
it can gain more credit and, if you are in academia, citations. 
The third reason is more profound:
reproducibility is a prerequisite of falsifiability,
the backbone of science [@Popper1959].  The results of
non-reproducible research cannot be verified, reducing its
scientific credibility.
All these reasons inform the book's practical nature.

```{r, echo=FALSE}
# [@Popper1959]
# Why spend time and effort on reproducibility? The first answer is that
# reproducibility actually saves time in the long-run, by ensuring more
# readable code and allowing your results to be easily re-run at a later data.
# The second reason is more profound. Reproducibility is a prerequisite
# of falsifiability and falsifiability is the backbone of science
# (Popper, 1959).
# The results on non-reproducible research cannot be verified, reducing scientific
# credibility. These observations inform the book's practical nature.
# The aim is simple: to provide a foundation in spatial microsimulation.
# http://en.wikipedia.org/wiki/Experiential_learning
# Poper's link also here
# [2011](http://www.manning.com/kabacoff/)
```

\pagebreak
This book presents spatial microsimulation as a living, evolving set of
techniques rather than a prescriptive formula for arriving at the 'right'
answer.  Spatial microsimulation is largely defined by its user-community, made
up of a growing number of people worldwide. This book
aims to contribute to the community by encouraging collaboration, innovation and rigour. It also encourages playing with the methods and
'getting your hands dirty' with the code. As
@kabacoff2011r [p. xxii]
put it regarding R: "The best way to learn is to
experiment". 

## Why spatial microsimulation with R? {#whyR}

```{r, echo=FALSE}
# @Kabakoff [p. xxii]
# The book aims to make spatial microsimulation accessible to more people,
# with a practical approach that encourages playing with the data and code.
# expressing oneself.^[This video introduces the idea of
# expressing oneself in [R](http://youtu.be/wki0BqlztCo)].
# [Hölm (1987, p.
# 153)](http://www.jstor.org/stable/10.2307/490448)
# @Holm1987 [p. 153]
# @Holm1987
```

Software decisions have a major impact on the flexibility, efficiency and
reproducibility of research.  Nearly three decades ago
@Holm1987
observed that "little
attention is paid to the choice of programming language used" for
microsimulation. This appears to be as true now as it was then. Software is
rarely discussed in papers on the subject and there are few mature spatial
microsimulation packages.^[The Flexible Modelling Framework
([FMF](https://github.com/MassAtLeeds/FMF)) is a notable exception written in
Java that can perform various modelling tasks.] 
There is a strong and diverse software community in the ABM field,
and many of these bring insights that could be of use to researchers using
spatial microsimulation. @Mannion2012, for example, present JAMSIM.
This is a new software framework for combining the powerful Repast ABM 
application with R for graphing and analysis of the results. JAMSIM
and other projects in the field (including the development of NetLogo and
MASON ABM systems and associated add-ons) should be seen as developing in
parallel to the R approach developed in this field, not competitors. There
is clearly much potential for mutual benefit by software for ABM and
spatial microsimulation interacting more closely together. 

Factors that should influence software selection include cost, maturity,
features and performance. Perhaps most important for busy researchers are the
ease and speed of learning,
writing, adapting and communicating the analysis. R excels in
each of these areas.

```{r, echo=FALSE}
# Yet the software used has a lasting
# impact, including what can and cannot be done
# and opportunities for collaboration.  explains the choice of R.
# In my own research, for example, a conscious decision was made early on to use
# R. This had subsequent knock-on impacts on
# the features, analysis and even design of my simulations.
# There are hundreds computer programming languages and many of these
# are general purpose and 'Turing complete', meaning they could, with
# sufficient effort, perform spatial microsimulation (or any other
# numerical operation). So why choose R?
# ^[Speed
# of execution is an arguable exception, an issue that can be tackled
```

R is a *low level* language compared with statistical programs based on a strong
graphical user interface (GUI) such as Microsoft Excel and SPSS.  R offers great
flexibility for analysing and modelling data and enables easy creation of
user-defined functions. These are all desirable attributes of
software for undertaking spatial microsimulation.
On the other hand, R is *high level* compared with
general purpose languages such as C and Python.  Instead of writing code to
perform statistical operations 'from scratch', R users generally use pre-made
functions. To calculate the mean value of variable `x`, for example, one would
need to type 20 characters in Python: `float(sum(x))/len(x)`.^[The `float`
function is needed in case integers are used. This can be reduced to 13
characters with the excellent **NumPy** package: `import numpy; x = [1,3,9];
numpy.mean(x)` would generate the desired result. The R equivalent is `x =
c(1,3,9); mean(x)`.] In pure R just 7 characters are sufficient: `mean(x)`. This
terseness and range of pre-made functions is useful for ease of
reading and writing spatial microsimulation models and analysing the results.

```{r, echo=FALSE}
# One may argue that saving a few keystrokes while writing
# code is not a priority but it is certain
# that the time savings of being concise can be vast.
```

The example of calculating the mean in R and Python
may be trite but illustrates a wider point: R
was *designed* to work with statistical data, so many functions in the default R
installation (e.g. `lm()`, to create a linear regression model) perform
statistical analysis 'out of the box'.  In agent-based modelling, the
statistical analysis of results often occupies more time than running the model
itself
[@Thiele2014].
The same
applies to spatial microsimulation, making R an ideal choice due to its
statistical analysis capabilities.

R has an active and  growing user community and is easy to extend.
R's extreme flexibility allows it to call code written in other programming
languages. This means that 'glass ceilings' encountered in other environments
are not an issue in R: there is a wide range of algorithms that can be used
from within R. This ability has been used to dramatically speed up R code.
The new R package **dplyr**, for example, uses C++ code to do the hard work
of data manipulation tasks.

As a result of R's flexibility, open source ethic and strong user community, there are
thousands of packages that extend R's capabilities with new functions.
Improvements are being added to the 'R ecosystem' all the time. This book provides an
overview of the functionality of key packages for spatial microsimulation.
Software development is a fast moving game, however, so keep a
look out for updates.
New software can save both research and computer time
so it's worth staying up-to-date with the latest developments.

The **ipfp** and **mipfp** packages, for example,
can greatly reduce the number of lines of code *and* the computational time
needed for population synthesis compared with 'hard-coding' the method
in R from scratch. (For the curious reader, the 'IPF' letters in the package
names stand for 'Iterative Proportional Fitting', one of a number of
technical terms we will use frequently
in subsequent chapters. See the Glossary for a definition.)

These time-saving add-ons to R are described in Chapter 5.
A clear advantage of R is that anyone can write a package.
This is also potentially a disadvantage: it has been argued there 
are too many packages, making it difficult for new users to identify which
packages are most reliable and which are best suited to different situations
[@Hornik2012]. However, this problem is mitigated by the open-source
nature of R: users can see precisely how any particular function
works (providing they are willing to learn some programming), rather than
relying on a "black box".

A recent development that has made R far more accessible is
RStudio, an Integrated Development Environment (IDE) for R that helps
new users become familiar with the language (see Figure 2.2). 
Further information about why R is a good choice for spatial microsimulation
is provided in the Appendix, a tutorial introduction to R for spatial
microsimulation applications. The next section describes approaches
to learning R in general terms.

```{r, echo=FALSE}
# For speed-critical applications,
# R provides access to lower level languages. It
# is possible to say a lot in R in few lines of code,
# but it is also possible for users to create their own
# commands, allowing users complete control. 
# The reasons for using R for spatial
# microsimulation can be summarised by modifying
# the arguments put forward by Norman Matloff (2001)
# for using R in general. R is:
# 
# -  "the de facto standard among
#     professional statisticians", meaning that the spatial microsimulation
#     code can easily be modified to perform a variety of statistical operations.
#  
# 
# -   "a general
#     programming language, so that you can automate your analyses and
#     create new functions." This is particularly useful if you need to run the same
#     code in many different ways for many locations. In R, the computer
#     can be asked to iterate over as many combinations of model runs as desired.
# 
# -   open source, meaning its easy to share your code and reproduce your
#     findings anywhere in the world, without the worry of infringing copyright
#     licences. In work funded by the public, this also has a large benefit
#     in terms of education and the democratisation of research.
```

## Learning the R language {#learningR}

Having learned a little about *why* R is a good tool for the job, it is worth
considering at this stage *how* R should be used.  It is useful to think of R
not as a series of isolated commands, but as an interconnected *language*.
The code is used not only for the computer to crunch numbers,
but also to communicate ideas, from one person to another.
In other words, this book teaches spatial microsimulation in the language of R.
Of course, English is more appropriate than R for *explaining* rather than
merely describing the method and the language of mathematics is ideal
for describing quantitative relationships conceptually. However,
because the practical components of this book are implemented in R, you will gain
more from it if you are fluent in R. To this end the book aims to improve your
R skills as well as your ability to perform spatial microsimulation, especially
in the earlier practical chapters. Some prior knowledge of R will make
reading this book easier, but R novices should be able to follow the worked
examples, with reference to appropriate supplementary material.

The references listed in the Appendix
([](#further))
provide a recommended starting point for R novices.
For complete beginners, we recommend the introductory 'Owen R Guide' [@owen2006r].
As with learning Spanish or Chinese, frequent practice, persistence and
experimentation will ensure deep learning.

A more practical piece of advice is to organise your workflow.  Each
project should have its own self-contained folder containing all that is needed
to replicate the analysis, except perhaps large input datasets. This could
include the raw (unchanged) input data^[Raw data should be kept safely on an
external hard disk or a server if it is large or sensitive.], R code for analysis,
the graphical outputs and files containing data outputs.  To avoid clutter,
it is sensible to arrange this content into folders, as illustrated below
(thanks to Colin Gillespie for this tip): 

```
|-- book.Rmd
|-- data
|-- figures
|-- output
|-- R
|   |-- load.R
|   `-- parallel-ipfp.R
`-- spatial-microsim-book.Rproj
```

The example directory structure above is taken from an early version of this
book.  It contains the document for the write-up (`book.Rmd` --- this could
equally be a `.docx` or `.tex` file) and RStudio's `.Rproj` file in the *source
directory*.  The rest of the entities are folders: one for the input data, one
for figures generated, one for data outputs and one for R scripts. The R scripts
should have meaningful names and contain only code that works and is commented.
An additional backup directory could be used to store experimental code. There
is no need to be prescriptive in following this structure, but projects using
spatial microdata tend to be complex, so imposing order over your workflow early
will likely yield dividends in the long run. If you work in a team, this kind of
structure helps each member to rapidly understand and continue your work.

The same applies to learning the R language.  Fluency allows complex numerical
ideas to be described with a few keystrokes.  If you are a new R
user it is therefore worth spending some time learning the R language. To this
end the Appendix provides a primer on R from the perspective of spatial
microsimulation.

```{r, echo=FALSE}
# Consider the following expression in the language of mathematics:
# 
# 
# 
# It is easy for experienced R users to translate this into R:
# 
# 
# 
# Note that although the R language is not quite as concise or elegant as
# mathematics, it is certainly faster at conveying the meaning of numerical
# operations than plain English and, in many cases, other programming languages.
# 
# 
# 
# The unusually concise nature of R code is not an accident. It was
# planned to be this way from the outset by its early developers, Robert
# Gentleman and Ross Ihaka, who thought carefully about syntax from the
# outset: "the syntax of a language is important because it determines the
# way that users of the language express themselves" (Ihaka and Gentleman, 2014, p. 300).
# 
# If you are new to R but have some experience with data analysis and
# microsimulation, do not feel intimidated that R is a foreign language.
# As with a spoken language, often the best way to learn is to
# 'jump in the deep end' by living abroad, so learning R through the course
# of this book is certainly an option. However, a deep understanding of R
# will greatly assist understanding the practical elements of the book which
# to focus primarily on the methods of spatial microsimulation and not the
# language in which they are implemented.
```

## Typographic conventions {#typographic}

The following typographic conventions are followed to make the practical
examples easier to follow:

- In-line code is provided in `monospace` font to show it's something the
computer understands.
- Larger blocks of code, referred to as *listings*, are provided on separate lines
and have coloured *syntax highlighting* to distinguish between values, names and functions:

```{r}
x <- c(1, 2, 5, 10) # create a vector
sqrt(x) # find the square root of x
```
 - Output from the *R console* is preceded by the `##` symbol, as illustrated above.
 - Comments are preceded by a single `#` symbol to explain specific lines.
 - Often, reference will be made to files contained within the book's project
 folder. The notation used to refer to the location of these files follows
 the way we refer to files and folders on Linux computers. Thus 'code/CakeMap.R'
 refers to a file titled 'CakeMap.R', within the 'R' directory of the project's
 folder. 
 
There are many ways to write R code that will generate the same results.
However, to ensure clarity and consistency, a single style, advocated in
a [chapter](http://r-pkgs.had.co.nz/style.html) in Hadley
Wickham's *Advanced R*, is followed
throughout
[@wickham2014adv]. 
Consistent style and plentiful comments will make your code
readable by yourself and others for decades to come.

## An overview of the book {#overview}

The book's chapters progress in a logical
order that corresponds to the steps typically taken during a spatial microsimulation
research project that generates, analyses and models spatial microdata. The
next two chapters are primarily conceptual, introducing the concept and applications
of spatial microsimulation in more detail whilst the remaining chapters are
practical. Because we'll be making heavy use of terminology specific to
spatial microsimulation, it is recommended that you read these introductory
chapters before tackling the practical section that follow although this is
not essential. If any of the words do not make sense, the
Glossary at the end of the book may help to clarify what they mean.

In Part II, Chapters
4, 5, 6 and 7 cover in detail the process of *population synthesis* whereby
spatial microdata is generated. This process is
key to much spatial microsimulation work. These chapters
should ideally be read together in that order, although Chapter 5 is the
most important and provides the basics needed to create spatial microdata.
Chapters 8 introduces the important step of model validation.


In Part III, Chapters 9 and 10 cover more
advanced methods for generating spatial microdata when
no individual level data is available and when household level data are required.
These chapters are more
self-standing and can be read in any order. The chapter titles and brief
descriptions of their contents are as follows:

```{r, echo=FALSE}
# Idea: the book could be easily split into 3 parts. Worth it?
```

- [*SimpleWorld*](http://spatial-microsim-book.robinlovelace.net/simpleworld)
is a simple and reproducible
explanation of spatial microsimulation with reference to an imaginary planet.
It's also a chance to set up your system and RStudio settings to
get the most out of the subsequent chapters.

- [*What is spatial microsimulation?*](http://spatial-microsim-book.robinlovelace.net/what-is.html)
introduces the concepts and applications of spatial 
microsimulation not only as a narrow methodology
but as an *approach* to understanding real-world phenomena.

- [*Data preparation*](http://spatial-microsim-book.robinlovelace.net/data-prep.html)
is dedicated to the boring but vital task
of loading and 'cleaning' the input data, ready for spatial microsimulation.

- [*Population synthesis*](http://spatial-microsim-book.robinlovelace.net/smsimr.html)
uses the input data created in the previous chapter to perform population
synthesis, the process of creating synthetic spatial microdata. 
The chapter describes the main functions
and techniques that are used for undertaking this key aspect of spatial
microsimulation, with a focus
on the **ipfp** and **mipfp** packages, which perform Iterative Proportion
Fitting (IPF). 

- [*Alternative approaches to population synthesis*](http://spatial-microsim-book.robinlovelace.net/alternative-reweighting.html) describes some alternative methods, beyond IPF, for population
synthesis. These include an R implementation the
Generalized Regression Weighting procedure (GREGWT), a description
of 'simulated annealing'
and a demonstration that population synthesis can be interpreted as a
constrained optimisation problem. This final insight has the consequence
that many previously developed optimisation algorithms can be used to
perform the 'heavy lifting' of population synthesis. A few of these,
including R's `optim` and `GenSA` functions, are tested, illustrating the
importance of performance testing.

- [*Spatial microsimulation in the wild*](http://spatial-microsim-book.robinlovelace.net/cakemap)
is based on a larger
and more complex example than SimpleWorld called CakeMap, that uses real data.
This chapter combines geographic constraints with non-geographical survey
data to
estimate the geographical variation in cake consumption.

- [*Model checking and evaluation*](spatial-microsim-book.robinlovelace.net/svalidation.html)
tackles the understudied but important questions of 'is the model working?' and 'do the results coincide with reality?'. Here you will learn appropriate tests and checks to verify spatial microsimulation outputs and think about its underlying assumptions.

- [*Population synthesis without microdata*](http://spatial-microsim-book.robinlovelace.net/nomicrodata.html)
gives a way to produce a spatial microsimulation without having individual level data.

- [*Household allocation*](http://spatial-microsim-book.robinlovelace.net/ha.html)
explains techniques to group the synthetic individuals into households.

- [*The TRESIS approach to spatial microsimulation*](http://spatial-microsim-book.robinlovelace.net/tresis.html)
describes an alternative approach to spatial microsimulation
that operates at the household level. This contributed chapter, by
Richard Ellison and David Hensher, demonstrates the population synthesis
stage of the wider Transport and Environment Strategy Impact Simulator
(TRESIS) modelling system. 

- [*Spatial microdata in agent-based models*](http://spatial-microsim-book.robinlovelace.net/abm)
describes how the output
from spatial microsimulation can be used as an input into more complex models.

The final thing to say in this opening chapter relates to the future
of this book as an up-to-date and accessible teaching resource.
*Spatial Microsimulation with R*
is both a book and an open source, open access project that is designed to be improved
on and updated by the authors and the community using spatial microsimulation.
Any comments relating to code, content or
clarity of explanation will therefore be gratefully
received.^[Feedback can be left via email to
r.lovelace@leeds.ac.uk or via the project's GitHub page
(http://github.com/Robinlovelace/spatial-microsim-book).]
