---
title: "SimpleWorld"
layout: default
output: pdf_document
bibliography: bibliography.bib
---

# SimpleWorld: A worked example of spatial microsimulation {#SimpleWorld}

This chapter focuses on the minimum input datasets needed for the classical type
of microsimulation. We will use input data on
the inhabitants of an imaginary world
(geographical individual-level) called SimpleWorld to demonstrate
the basic concepts and techniques to perform spatial microsimulation with R.

This is the
first practical chapter, so it also aims get a good 'work-flow'.
By work-flow you up-to-speed with
using R via RStudio. The aim is twofold: 1) to explain key
concepts in spatial microsimulation with reference to a practical example and 2) to
demonstrate the software and a workflow
that will make developing spatial microsimulation models as easy as possible.

```{r, echo=FALSE}
# TODO: do we talk enough about RStudio projects?
```

The chapter also serves to highlight the links between the methodology
presented in-depth later in the book (see Chapter 5)
and the various real-world applications presented in Chapter 3.
For R 'newbies' it also provides a chance
for the reader to do some basic programming in R:
the unwritten
subtitle of this book is "a *practical* introduction" for a reason!

```{r, echo=FALSE}
# TODO (RL): Update the above if we get a subtitle, mention in preface 
```

This chapter is ordered as follows:

- *Getting set-up with the RStudio environment* ([](#rstudioUpSpeed)) contains
the basic explanation of RStudio to enable you to code the examples of this 
book on your own computer.
- *SimpleWorld data* ([](#SimpleWorldData)) describes the context and the
data of this little illustrative example.
-*Generating a weight matrix* ([](#weight-matrix)) contains the code and the description of 
the role of a weight matrix in a spatial microsimulation.
-*Spatial microdata* ([](#spatial-microdata)) shows the typically useful output 
of a spatial microsimulation.
-*SimpleWorld in context* ([](#SimpleWorldContext)) mentions the kind of
context where the output could be use.

## Getting set-up with the RStudio environment {#rstudioUpSpeed}

Before progressing further, it is important to ensure that R is set-up correctly
and working on your computer, so you can follow the practical examples.
This section will not go into much detail. Further resources are provided in
the Appendix at the end of the book to provide insight into how R
works as a programming language.

The majority of the help that you need to install and set-up R on your computer
can be found online, so it is recommended that you consult this section with
reference to online searches and help pages. As with all open source software
projects R evolves over time so it is important to keep up-to-date with
any changes, which may not be accounted for in this book.

### Installing R

This section assumes that you already have R installed on your computer.
If you have not, please refer to the documentation on the Comprehensive R Archive Network
(CRAN):

- For Windows, see https://cran.r-project.org/bin/windows/base/
- For Mac, see https://cran.r-project.org/bin/macosx/
- For Linux, see https://cran.r-project.org/doc/manuals/r-release/R-admin.html.
On Debian-based systems such as Ubuntu, the following bash command should be sufficient to install R.

```
sudo apt-get install r-base
```

Once R is installed, new packages are easy to install from within R using
`install.packages("package_name")`. To install the **mipfp** package, for example,
we would use:

```{r, eval=FALSE}
install.packages("mipfp")
```


### RStudio

RStudio is an Integrated Development Environment (IDE) for R.
It provides an advance Graphical User Interface (GUI) that makes it easier not only to write R code, but to install packages, manage R's graphical outputs and work effectively on mulitple projects.
If you are an R beginner, it is highly recommended that you use RStudio for the work presented in this book because it will enable you to spend less time worrying about how to write R code and set-up R correctly and more time focussed on spatial microsimulation.
One example of how RStudio saves time is its tab autocompletion functionality.
Type the beginning of any object or function stored in R's environment,
hit the `Tab` button, and the name should auto-complete, as illustrated below.

```{r rstudio,  fig.cap="Autocompletion in RStudio", fig.scap="The RStudio interface", fig.align='center', echo=FALSE, message=FALSE}
# todo: create tutorial on installing R and RStudio
grid.raster(readPNG("figures/rstudio-autocomplete.png"))
```

To test this functionality, download and install RStudio if you have not already.
The installation process is simple on Linux, Mac and Windows platforms:
simply follow the instructions provided here at rstudio.com/products/rstudio/download/.

There are many other advantages of using
RStudio.^[For
more information about RStudio, please see the RStudio website:
https://www.rstudio.com/products/rstudio/features/ and search for
other on-line resources.
There is even an entire book dedicated to the subject
[@verzani2011getting].
]
The next section focuses on one particular advantage that we recommend you
make use of whilst working through the book: RStudio projects.

### Projects

RStudio provides a neat system for creating, managing and even sharing your projects.
This functionality merits a section of its own, because spatial microsimulation
projects are likely to be large and complex, requiring collaboration with
other people. If you are not careful the projects can become over-complex
and un-manageable. RStudio projects can prevent this from happening and
lead to improved workflow.

To set-up a new project click on the drop-down menu in the top-right
section of RStudio. If you are work in a project the name of the
project will appear in this dropdown menu (see Fig. 2.3).
When a project is loaded, the following things happen:

- The files open in the script panel (top left) last time your worked on the project will open.
- R's working directory (set and checked with `setwd()` and `getwd()` respectively) will change.
- Any R objects saved in the `.RData` file in the project will be automatically loaded, saving your working.
- If you are using `git` to manage your project's development and sharing, you'll have options to commit code and 'push' new work, e.g. to GitHub.

In fact, this entire book was written as an RStudio project.
This allowed us to work from the same system and share code easily, by pushing our
code frequently to the book's online repository: https://github.com/Robinlovelace/spatial-microsim-book .
This online repository will allow you to access all of the example code and data for the book.
It will also allow you to contribute to the book as an evolving project.
The next section explains how to download and use the data stored in the `spatial-microsim-book`
GitHub repository to access the code and data resources associated with the book.

### Downloading data for the book

There are two main ways to download the code and data associated with this book onto your computer:
by 'cloning' the repository and by downloading and extracting its contents in a .zip file.
We will only cover the second option as it's simpler. For details on how to fork, clone
and potentially contribute to the book project, see the GitHub website. There is also a
growing literature on GitHub [@lima_coding_2014].

To simply download the code and data, first navigate to
https://github.com/Robinlovelace/spatial-microsim-book in your browser. From this page, click on the
'Download ZIP' button to the right and extract the folder into a
sensible place on your computer, such as the Desktop.  From there, you will want
to run R from the project's root directory: open the folder in a file browser
and double click on `spatial-microsim-book.Rproj`.
Alternatively, start RStudio and click on 'Open Project' from the dropdown menu above the top-right panel
(you can also open the project file with File > Open).
This should cause RStudio to
be opened at this location, with all the input data files easily accessible to R
through *relative file paths* (you must have installed the RStudio software for this stage to work --- see the Appendix).

\index{RStudio}

```{r studio,  fig.cap="The RStudio interface with the spatial-microsim-book project loaded and objects loaded in the Environment after running SimpleWorld.R",fig.scap="The RStudio interface", fig.align='center', echo=FALSE, message=FALSE}
# todo: create tutorial on installing R and RStudio
grid.raster(readPNG("figures/rstudio-environment.png"))
```

To check if the project has been downloaded and opened correctly, try running the following code.

```{r, message=FALSE}
source("R/SimpleWorld.R")
```

You should see some output in red. Further, some new objects should appear in the Environment tab in the top-right panel (Figure 2.3).
This is the result of running the code located in a script file called `SimpleWorld.R`, using the `source()` function.

Now we have an understanding of RStudio and the project loaded, it's time to 'get our hands dirty'
with the SimpleWorld example, and run some of the contents of `SimpleWorld.R` line by line.

## SimpleWorld data {#SimpleWorldData}

SimpleWorld is a small world, consisting of 33 persons split across 3 zones.
We have two sources of information about these people available to us:

1. aggregate counts of persons by age and sex in each zone (from the SimpleWorld Census)
2. survey microdata recording more detailed information (age, sex and income), for five of the world’s residents.

Unfortunately the survey data lack geographical information, and include only 
a small subset of the population (5 out of 33). To infer further information
about SimpleWorld --- and more importantly to be able to model its
inhabitants --- we need a methodology. This is precisely the kind of situation
where spatial microsimulation is useful. After an explanation of this
'starting simple' approach, we describe the input data in detail, demonstrate
what spatial microdata look like and, finally, place the example in its wider
context.

This SimpleWorld example is analogous to the
'cartoon world' used by professor @mackay2008sustainable to simplify the
complexities of sustainable energy. The same principle works here:
spatial microsimulation will be used to simulate
differences in income and age distributions
between the three zones SimpleWorld, illustrating
the kinds of situation where spatial microsimulation can be useful.

The SimpleWorld planet,
a 2 dimensional
plane split into 3 zones, is illustrated in figure 3.1. 

```{r simpleworld,  fig.cap="The SimpleWorld environment, consisting of individuals living in 3 zones.", fig.scap="The SimpleWorld environment", fig.align='center', echo=FALSE, message=FALSE}
# Code to create SimpleWorld
# Builds on this vignette: http://cran.r-project.org/web/packages/sp/vignettes/over.pdf
library(sp)
library(ggplot2)
xpol <- c(-180, -60, -60, -180, -180)
ypol <- c(-70, -70, 70, 70, -70)
pol = SpatialPolygons(list(
  Polygons(list(Polygon(cbind(xpol, ypol))), ID="x1"),
  Polygons(list(Polygon(cbind(xpol + 120, ypol))), ID="x2"),
  Polygons(list(Polygon(cbind(xpol + 240, ypol))), ID="x3")
  ))
# plot(pol)
proj4string(pol) <- CRS("+init=epsg:4326")
pol1 <- fortify(pol)

theme_space_map <- theme_bw() +
  theme(
#     rect = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major = element_line(size = 3)
  )
# 
# ggplot(pol1) + geom_path(aes(long, lat, group, fill = group)) +
#     coord_map("ortho", orientation=c(41, -74, 52)) + 
#   theme_space_map
b <- matrix(c(-180, 180, 180, -180, -180, 90, 90, -90, -90, 90), ncol = 2)
library(sp)
l <- Line(coords = b)
l <- Lines(list(l), ID = "a")
l <- SpatialLines(list(l))
library(rgeos)
p <- gPolygonize(l)
plot(p)
l1 <- Line(matrix(c(-60, -60, 90, -90), ncol = 2))
l1 <- Lines(list(l1), "a")
l1 <- SpatialLines(list(l1))
lines(l1)
l2 <- Line(matrix(c(60, 60, 90, -90), ncol = 2))
l2 <- Lines(list(l2), "a")
l2 <- SpatialLines(list(l2))
lines(l2)
zt <- paste0("Zone ", 1:3)
text(seq(-120, 120, length.out = 3), 0, zt)
```


SimpleWorld is inhabited by 12, 10 and 11 individuals of its alien inhabitants
in zones 1 to 3, respectively: a planetary population of 33.
From the SimpleWorld Census, we know
how many young (strictly under 50 space years old) and old (50 and over)
residents live in each
zone, as well their genders: male and female.
This information is displayed in the tables below.

Table: Aggregate-level age counts for SimpleWorld.

|zone   | 0-49 yrs| 50 + yrs|
|:--|-----:|-----:|
|1  |     8|     4|
|2  |     2|     8|
|3  |     7|     4|

Table: Aggregate sex counts for SimpleWorld.

|Zone   |  m|  f|
|:--|--:|--:|
|1  |  6|  6|
|2  |  4|  6|
|3  |  3|  8|

\newpage

The following commands load-in this data.
Play with the R objects you have loaded: try printing
them to screen, plotting them and even-subsetting them.
If you cannot, it may be worth consulting the Appendix
on using R.

```{r}
# Load the constraint data
con_age <- read.csv("data/SimpleWorld/age.csv")
con_sex <- read.csv("data/SimpleWorld/sex.csv")
cons <- cbind(con_age, con_sex)
```

```{r fig = "Mercator maps of the zones in SimpleWorld", echo=FALSE, message=FALSE}

# library(knitr)
# kable(con_age, row.names = T)
# kable(con_sex, row.names = T)

pol <- SpatialPolygonsDataFrame(pol, cons, match.ID = F)

# pol@data
pol$p_young <- pol$a0.49 / (pol$a.50. + pol$a0.49) * 100
pol$p_male <- pol$m / (pol$f + pol$m) * 100

pol$id <- c("x1", "x2", "x3")
library(dplyr)
pol1 <- inner_join(pol1, pol@data)
pol1$Name <- paste("Zone", 1:3, sep = " ")
pol1$xpos <- seq(-120, 120, length.out = 3)
pol1$ypos <- 0

# ggplot(pol1) + 
#   geom_polygon(aes(long, lat, group, fill = p_young)) +
#   geom_path(aes(long, lat, group, fill = p_young)) +
#   geom_text(aes(xpos, ypos, label = Name)) +
#   theme_bw() +
#   scale_fill_continuous(low = "black", high = "white", limits = c(0, 100),
#     name = "% Young") +
#   coord_map() 
# 
# ggplot(pol1) + 
#   geom_polygon(aes(long, lat, group, fill = p_male)) +
#   geom_path(aes(long, lat, group, fill = p_male)) +
#   geom_text(aes(xpos, ypos, label = Name)) +
#   theme_bw() +
#   scale_fill_continuous(low = "black", high = "white", limits = c(0, 100),
#     name = "% Male") +
#   coord_map() 
```

These geographically specific datasets
come from the SimpleWorld census. Responding to census questionnaires
is a legal requirement in SimpleWorld so we the planet's entire
population is represented in the counts.
From these *constraint tables* we know the marginal distributions
of the two categorical variables,
but not the contingency table (or cross-tabulation)
that links age and sex. This means that we have per-zone information on
the number of young and old people and the number of males and females.
But we currently lack information about the number
of young females, young males, and so on. Also note that we have no information
about other important variables such as income.

Spatial microsimulation can be used to better understand the
population of SimpleWorld. To do this, we need some additional information:
an individual-level *microdataset*.

This is provided in an individuals-level dataset on 5 of SimpleWorld's
inhabitants. Note that the individual-level data has different dimensions than the
aggregate data presented above.
The individual-level *microdata* survey has one row per individual whereas
the *constraints* have one row per zone.
This individual-level data includes two variables that *link* it to the
aggregate-level data described above: age and sex. The individual-level
data also provides information about a *target variable* not included
in the geographical constraints: income. To load the individual-level data,
enter the following:

```{r}
ind <- read.csv("data/SimpleWorld/ind-full.csv")
```


```{r, echo=FALSE}
# ind <- read.csv("data/SimpleWorld/ind.csv")
# ind$income <- round(rnorm(n = nrow(ind), mean = 1000, sd = 100))
# ind$income <- ind$income + 30 * ind$age
# ind$income[ ind$age == "f"] <- ind$income + 1000
# write.csv(ind, "data/SimpleWorld/ind-full.csv", row.names = F)
# kable(ind)
```

Table: Individual-level survey data from SimpleWorld.

| id| age|sex | income|
|--:|---:|:---|------:|
|  1|  59|m   |   2868|
|  2|  54|m   |   2474|
|  3|  35|m   |   2231|
|  4|  73|f   |   3152|
|  5|  49|f   |   2473|

Note that although the microdataset contains additional information about the
inhabitants of SimpleWorld, it lacks geographical information about where each
inhabitant lives or even which zone they are from.  This is typical of
individual-level survey data.  Spatial microsimulation tackles this issue by
allocating individuals from a non-geographical dataset to geographical zones in
another.

## Generating a weight matrix {#weight-matrix}

The process to generate spatial microdata is allocating *weights*
to each individual for each zone. The higher the weight for a particular
individual-zone combination, the more representative the individual is of that
zone. This information can be represented as a *weight matrix*.

The subsequent code block uses the **mipfp** package to convert these
inputs into a matrix of weights. Just type it in and observe the
result. If it works, congratulations! You have generated your first
weight matrix using Iterative Proportional Fitting (IPF). The
details of this process are described in subsequent chapters.

```{r, message=FALSE}
target <- list(as.matrix(con_age[1,]), as.matrix(con_sex[1,]))
descript <- list(1, 2)
ind$age_cat <- cut(ind$age, breaks = c(0, 50, 100))
seed <- table(ind[c("age_cat", "sex")])

library(mipfp) # install.packages("mipfp") is a prerequisite
res <- Ipfp(seed, descript, target)
res$x.hat # the result for the first zone
```

Note that the output of the above command is only for the first zone.
It suggests that in zone 1, younger females are far more (nearly 4.5 times more)
numerous than would be expected based on the individual level data alone.
This makes sense because `con_age[1,]` contains many young people and `ind`
contains only one young female. At this stage you may be wondering: what just
happened? This is explained in detail in the subsequent chapters.

An alternative approach to arriving at the same result, for all zones,
is demonstrated in the code chunk below using another package: **ipfp**.
Again, the reader is not expected to understand what has just happened.
This will be explained in subsequent chapters.

```{r}
A <- t(cbind(model.matrix(~ ind$age_cat - 1),
           model.matrix(~ ind$sex - 1)[, c(2, 1)]))
cons <- matrix(c(8,2,7,4,8,4,6,4,3,6,6,8), nrow = 3)
library(ipfp) # install.packages("ipfp") is a prerequisite
weights <- apply(cons, 1, function(x) ipfp(x, A, x0 = rep(1, 5)))
weights[,1] # result for the first zone
```


Table: A 'weight matrix' linking the microdata (rows) to the zones (columns)
of SimpleWorld.

| Individual| Zone 1  | Zone 2  | Zone 3  |
|---:|----:|----:|-----:|
|   1| 1.23| 1.73| 0.725|
|   2| 1.23| 1.73| 0.725|
|   3| 3.54| 0.55| 1.550|
|   4| 1.54| 4.55| 2.550|
|   5| 4.46| 1.45| 5.450|

The highest value (5.450) is located, to use R's notation, in cell
`weights[5,3]`, the 5th row and 3rd column in the matrix `weights`. This means
that individual number 5 is considered to be highly representative of Zone 3,
given the input data in SimpleWorld.  This makes sense because there are many
(7) young people and many (8) females in Zone 3, relative to the input
microdataset (which contains only 1 young female). The lowest value (0.55) is
found in cell `[3,2]`. Again this makes sense: individual 3 from the
microdataset is a young male yet there are only 2 young people and 4 males in
zone 2. A special feature of the weight matrix above is that each of the column
sums is equal to the total population in each zone.

We will discover different ways of generating such weight matrices
in subsequent chapters.
For now it is sufficient to know
that they matrices link individual-level data to geographically aggregated
data and that there are multiple techniques to generate them. The techniques are sometimes
referred to as *reweighting algorithms* in the literature (Tanton et al. 2011).
These include deterministic methods such as Iterative Proportional
Fitting (IPF) and
probabilistic methods that rely on a pseudo-random number generator such as
simulated annealing (SA). These and other reweighting algorithms are described
in detail in Chapter 5.

## Spatial microdata {#spatial-microdata}

A useful output from spatial microsimulation is what we refer to as
*spatial microdata*.  This is a dataset that contains a single row per individual
(as with the input microdata) but also an additional variable on geographical location.
The ideal spatial microdataset
selects individuals representative of the aggregate constraints for each zone,
while containing the diversity of information present in the individual-level
non-spatial input population. In Chapter 4 we will explore all the steps needed
to produce a spatial microdataset for SimpleWold. A subset of such spatial microdataset
is presented in Table 2.5 below.

\pagebreak

| id| zone| age|sex | income|
|--:|----:|---:|:---|------:|
|  1|    2|  59|m   |   2868|
|  2|    2|  54|m   |   2474|
|  4|    2|  73|f   |   3152|
|  4|    2|  73|f   |   3152|
|  4|    2|  73|f   |   3152|
|  4|    2|  73|f   |   3152|
|  5|    2|  49|f   |   2473|
|  4|    2|  73|f   |   3152|
|  5|    2|  49|f   |   2473|
|  2|    2|  54|m   |   2474|

Table: Spatial microdata generated for SimpleWorld zone 2. Each row represents an individual taken from the individual-level sample and the 'zone' column represents in where in geographical space they reside.

The Table 2.5 is a reasonable approximation of the inhabitants of zone 2: older
females dominate in both the aggregate (which contains 8 older people and 6
females) and the simulated spatial microdata (which contains 8 older people and
6 females). Note that in addition the constraint variables, we also have
an estimate the income distribution in SimpleWorld's second zone.

Towards the end of Chapter 5 we will learn how to generate
this table. The remainder of this section
considers how the outputs of spatial microsimulation, in the context of
SimpleWorld, can be useful before progressing to the practicalities.

```{r, echo=FALSE}
# Add section link here!
```

## SimpleWorld in context {#SimpleWorldContext}

Even though the datasets are tiny in SimpleWorld, we have already generated some
useful output. We can estimate, for example, the average income in each zone.
Furthermore, we could create an estimate of the *distribution* of income in each
area. Although these estimates are unlikely to be very accurate due to the
paucity of data, the methods could be very useful if performed on larger
datasets from 'RealWorld' (planet Earth).  The spatial microdata
presented in the above table could be used as an input into an agent-based model
(ABM). Assuming the inhabitants of SimpleWorld are more predictable than those
of RealWorld, the outputs from such a model could be very useful indeed, for
example for predicting future outcomes of current patterns of behaviour.

In addition to clarifying the advantages of spatial microsimulation, the above
example also flags some limitations of the methodology. Spatial microsimulation
will only yield useful results if the input datasets are representative of
the population as a whole, and for each region. If the relationship between age and
sex is markedly different in one zone compared with what we assume to be the
global averages of the input data, for example, our estimates could be way out.
Using such a small sample, one could rightly argue, how could the diversity of
33 inhabitants of SimpleWorld be represented by our simulated spatial microdata?
This question is equally applicable to larger simulations. These issues are
important and will be tackled in
[](#svalidation).

## Chapter summary

In this chapter we have learned how to set-up the R/RStudio environment for running
the example code provided in this book. This involved installing the software,
understanding projects in RStudio and setting up the `spatial-microsim-book`
project so that data can be accessed easily.
This was the first practical chapter and it involved basic commands for loading
data and creating a weight matrix. The process of running and playing with the
code should have helped get acquainted with the 'workflow' associated with
developing and running spatial microsimulation models in R.
Discussion of the results, which represent the inhabitants of the imaginary
system *SimpleWorld*, create the backdrop for the rest of the book which
we will return to,
in [](#ABM).

# References
