---
title: "SimpleWorld"
layout: default
output: pdf_document
bibliography: bibliography.bib
---

# SimpleWorld: A worked example of spatial microsimulation {#SimpleWorld}

```{r, echo=FALSE}
# TODO: add more about using the spatial microdata
```

This chapter focuses on the minimum input datasets needed for the classical type
of spatial microsimulation. We will use input data on
the inhabitants of an imaginary world
(geographical individual-level) called SimpleWorld to demonstrate
the basic concepts and techniqes to perform spatial micrsomulation with R.

This is the
first practical chapter, so it also aims get a good 'work-flow'.
By work-flow you up-to-speed with
using R via RStudio. The aim is twofold: 1) to explain key
concepts in spatial microsimulation with reference to a practical example and 2) to
demonstrate the software and a workflow
that will make developing spatial microsimulation models as easy as possible.

```{r, echo=FALSE}
# TODO: do we talk enough about RStudio projects?
```

The chapter also serves to highlight the links between the methodology
presented in-depth later in the book (see Chapter 5)
and the various real-world applications presented in Chapter 3.
For R 'newbies' it also provides a chance
for the reader to do some basic programming in R:
the unwritten
subtitle of this book is "a *practical* introduction" for a reason!

```{r, echo=FALSE}
# TODO (RL): Update the above if we get a subtitle, mention in preface 
```

SimpleWorld is a small world, consisting of 33 persons split across 3 zones.
We have two sources of information about these people available to us:

1. aggregate counts of persons by age and sex in each zone (from the SimpleWorld Census)
2. survey microdata recording more detailed information (age, sex and income), for five of the world’s residents.

Unfortunately the survey data lack any geography and include only 
a little part of the population (5 out of 33). To infer further information
about SimpleWorld --- and more importantly to be able to model its
inhabitants --- we need a methodology. This is precisely the kind of situation
where spatial microsimulation is useful. After an explanation of this
'starting simple' approach, we describe the input data in detail, demonstrate
what spatial microdata look like and, finally, place the example in its wider
context.

This SimpleWorld example is analogous to the
'cartoon world' used by professor David MacKay (2009) to simplify the
complexities of sustainable energy. The same principle works here:
spatial microsimulation will be used to simulate
differences in income and age distributions
between the three zones SimpleWorld, illustrating
the kinds of situation where spatial microsimulation can be useful.

The SimpleWorld planet,
a 2 dimensional
plane split into 3 zones, is illustrated in figure 3.1. 

```{r simpleworld,  fig.cap="The SimpleWorld environment, consisting of individuals living in 3 zones.", fig.scap="The SimpleWorld environment", fig.align='center', echo=FALSE, message=FALSE}
# Code to create SimpleWorld
# Builds on this vignette: http://cran.r-project.org/web/packages/sp/vignettes/over.pdf
library(sp)
library(ggplot2)
xpol <- c(-180, -60, -60, -180, -180)
ypol <- c(-70, -70, 70, 70, -70)
pol = SpatialPolygons(list(
  Polygons(list(Polygon(cbind(xpol, ypol))), ID="x1"),
  Polygons(list(Polygon(cbind(xpol + 120, ypol))), ID="x2"),
  Polygons(list(Polygon(cbind(xpol + 240, ypol))), ID="x3")
  ))
# plot(pol)
proj4string(pol) <- CRS("+init=epsg:4326")
pol1 <- fortify(pol)

theme_space_map <- theme_bw() +
  theme(
#     rect = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major = element_line(size = 3)
  )
# 
# ggplot(pol1) + geom_path(aes(long, lat, group, fill = group)) +
#     coord_map("ortho", orientation=c(41, -74, 52)) + 
#   theme_space_map
b <- matrix(c(-180, 180, 180, -180, -180, 90, 90, -90, -90, 90), ncol = 2)
library(sp)
l <- Line(coords = b)
l <- Lines(list(l), ID = "a")
l <- SpatialLines(list(l))
library(rgeos)
p <- gPolygonize(l)
plot(p)
l1 <- Line(matrix(c(-60, -60, 90, -90), ncol = 2))
l1 <- Lines(list(l1), "a")
l1 <- SpatialLines(list(l1))
lines(l1)
l2 <- Line(matrix(c(60, 60, 90, -90), ncol = 2))
l2 <- Lines(list(l2), "a")
l2 <- SpatialLines(list(l2))
lines(l2)
zt <- paste0("Zone ", 1:3)
text(seq(-120, 120, length.out = 3), 0, zt)
```

## Getting set-up with the RStudio environment



## SimpleWorld data {#SimpleWorldData}

SimpleWorld is inhabited by 12, 10 and 11 individuals of its alien inhabitants
in zones 1 to 3, respectively: a planetary population of 33.
From the SimpleWorld Census, we know
how many young (strictly under 50 space years old) and old (50 and over)
residents live in each
zone, as well their genders: male and female.
This information is displayed in the tables below.

\newpage

Table: Aggregate-level age counts for SimpleWorld.

|zone   | 0-49 yrs| 50 + yrs|
|:--|-----:|-----:|
|1  |     8|     4|
|2  |     2|     8|
|3  |     7|     4|

Table: Aggregate sex counts for SimpleWorld.

|Zone   |  m|  f|
|:--|--:|--:|
|1  |  6|  6|
|2  |  4|  6|
|3  |  3|  8|


```{r fig = "Mercator maps of the zones in SimpleWorld", echo=FALSE, message=FALSE}
con_age <- read.csv("data/SimpleWorld/age.csv")
con_sex <- read.csv("data/SimpleWorld/sex.csv")
cons <- cbind(con_age, con_sex)

# library(knitr)
# kable(con_age, row.names = T)
# kable(con_sex, row.names = T)

pol <- SpatialPolygonsDataFrame(pol, cons, match.ID = F)

# pol@data
pol$p_young <- pol$a0.49 / (pol$a.50. + pol$a0.49) * 100
pol$p_male <- pol$m / (pol$f + pol$m) * 100

pol$id <- c("x1", "x2", "x3")
library(dplyr)
pol1 <- inner_join(pol1, pol@data)
pol1$Name <- paste("Zone", 1:3, sep = " ")
pol1$xpos <- seq(-120, 120, length.out = 3)
pol1$ypos <- 0

# ggplot(pol1) + 
#   geom_polygon(aes(long, lat, group, fill = p_young)) +
#   geom_path(aes(long, lat, group, fill = p_young)) +
#   geom_text(aes(xpos, ypos, label = Name)) +
#   theme_bw() +
#   scale_fill_continuous(low = "black", high = "white", limits = c(0, 100),
#     name = "% Young") +
#   coord_map() 
# 
# ggplot(pol1) + 
#   geom_polygon(aes(long, lat, group, fill = p_male)) +
#   geom_path(aes(long, lat, group, fill = p_male)) +
#   geom_text(aes(xpos, ypos, label = Name)) +
#   theme_bw() +
#   scale_fill_continuous(low = "black", high = "white", limits = c(0, 100),
#     name = "% Male") +
#   coord_map() 
```

These geographically specific datasets
come from the SimpleWorld census. Responding to census questionnaires
is a legal requirement in SimpleWorld so we the planet's entire
population is represented in the counts.
From these *constraint tables* we know the marginal distributions
of the two categorical variables,
but not the contingency table (or cross-tabulation)
that links age and sex. This means that we have per-zone information on
the number of young and old people and the number of males and females.
But we currently lack information about the number
of young females, young males, and so on. Also note that we have no information
about other important variables such as income.

Clearly, this geographic count data lacks detail on SimpleWorld's
inhabitants and is therefore unsuitable for many purposes. 
It is precisely such situations, where we have access to geographical count data
(usually from a census of population) on the one hand and a more
detailed and a non-geographical individual-level dataset on the other,
that spatial microsimulation is most useful.
It is in this context, where we would like to impute missing information
about the individuals in specific geographic zones,
that we turn to spatial microsimulation. In this 'cartoon' case study we
will see how spatial microsimulation can be used to better understand the
population of SimpleWorld. To do this, we need some additional information:
an individual-level *microdataset*.

This is provided in an individuals-level dataset on 5 of SimpleWorld's
inhabitants, recorded from a survey. Note that this data has a different
form from the aggregate-level data presented in the above tables.
The individual-level data has, for now, different dimensions than the
aggregate data presented above.
This *microdata* survey contains one row per individual, in contrast
to the *aggregate constraints*, which have one row per zone.
This individual-level data includes two variables that *link* it to the
aggregate-level data described above: age
(as an integer variable, rather than the
broad categories used in the zone data) and sex. The individual-level
data also provides information about a *target variable* not included
in the geographical constraints: income.

```{r, echo=FALSE}
# ind <- read.csv("data/SimpleWorld/ind.csv")
# ind$income <- round(rnorm(n = nrow(ind), mean = 1000, sd = 100))
# ind$income <- ind$income + 30 * ind$age
# ind$income[ ind$age == "f"] <- ind$income + 1000
# write.csv(ind, "data/SimpleWorld/ind-full.csv", row.names = F)
ind <- read.csv("data/SimpleWorld/ind-full.csv")
# kable(ind)
```

Table x: Individual-level survey data from SimpleWorld.

| id| age|sex | income|
|--:|---:|:---|------:|
|  1|  59|m   |   2868|
|  2|  54|m   |   2474|
|  3|  35|m   |   2231|
|  4|  73|f   |   3152|
|  5|  49|f   |   2473|

Note that although the microdataset contains additional information about the
inhabitants of SimpleWorld, it lacks geographical information about where each
inhabitant lives or even which zone they are from.  This is typical of
individual-level survey data.  Spatial microsimulation tackles this issue by
allocating individuals from a non-geographical dataset to geographical zones in
another.

## Practical: generating a weight matrix {#weight-matrix}

The procedures we will learn to use in this book do this by allocating *weights*
to each individual for each zone. The higher the weight for a particular
individual-zone combination, the more representative that individual is of that
zone. This information can be represented as a *weight matrix*.

The first practical challenge in this book is to generate a weight matrix.
Usually, the input data for such procedures are read in from a file.
However, to get acquainted with typing R code and to 'get your hands dirty'
we type in the data. Then the **ipfp** package is used to convert these
inputs into a matrix of weights. Don't worry if this makes little sense:
it looks like gobbledygook to most people! Just type it in and observe the
result. If it works, congratulations! You have generated your first
weight matrix using Iterative Proportional Fitting (IPF). The nuts and
bolts of this process are described in the next chapter.

```{r, eval=FALSE}
cons <- matrix(c(8,2,7,4,8,4,6,4,3,6,6,8), nrow = 3)

A <- matrix(c(0,1,1,0,0,1,1,0,1,0,1,0,0,1,0,1,1,0,0,1), nrow = 4)

library(ipfp) # install.packages("ipfp") is a prerequisite
weights <- apply(cons, 1, function(x) ipfp(x, A, x0 = rep(1, 5)))
weights
```

Table: A 'weight matrix' linking the microdata (rows) to the zones (columns)
of SimpleWorld.

| Individual| Zone 1  | Zone 2  | Zone 3  |
|---:|----:|----:|-----:|
|   1| 1.23| 1.73| 0.725|
|   2| 1.23| 1.73| 0.725|
|   3| 3.54| 0.55| 1.550|
|   4| 1.54| 4.55| 2.550|
|   5| 4.46| 1.45| 5.450|

The highest value (5.450) is located, to use R's notation, in cell
`weights[5,3]`, the 5th row and 3rd column in the matrix `weights`. This means
that individual number 5 is considered to be highly representative of Zone 3,
given the input data in SimpleWorld.  This makes sense because there are many
(7) young people and many (8) females in Zone 3, relative to the input
microdataset (which contains only 1 young female). The lowest value (0.55) is
found in cell `[3,2]`. Again this makes sense: individual 3 from the
microdataset is a young male yet there are only 2 young people and 4 males in
zone 2. A special feature of the weight matrix above is that each of the column
sums is equal to the total population in each zone.

We will discover different ways of generating such weight matrices
in subsequent chapters.
For now it is sufficient to know
that they matrices link individual-level data to geographically aggregated
data and that there are multiple techniques to generate them. The techniques are sometimes
referred to as *reweighting algorithms* in the literature (Tanton et al. 2011).
These include deterministic methods such as Iterative Proportional
Fitting (IPF) and
probabilistic methods that rely on a pseudo-random number generator such as
simulated annealing (SA). These and other reweighting algorithms are described
in detail in Chapter 5.

## Spatial microdata {#spatial-microdata}

A useful output from spatial microsimulation is what we refer to as
*spatial microdata*.  This is a dataset that contains a single row per individual
(as with the input microdata) but also an additional variable on geographical location.
The ideal spatial microdataset
selects individuals representative of the aggregate constraints for each zone,
while containing the diversity of information present in the individual-level
non-spatial input population. In Chapter 4 we will explore all the steps needed
to produce a spatial microdataset for SimpleWold. A subset of such spatial microdataset
is presented in Table 3.5 below.

The complete spatial microdataset allocates whole individuals to
provides an insight into the nature and spatial distribution of the
inhabitants of SimpleWorld. This spatial microdataset is also useful
for the purposes of modelling, as described in Chapter 12.

\pagebreak

| id| zone| age|sex | income|
|--:|----:|---:|:---|------:|
|  1|    2|  59|m   |   2868|
|  2|    2|  54|m   |   2474|
|  4|    2|  73|f   |   3152|
|  4|    2|  73|f   |   3152|
|  4|    2|  73|f   |   3152|
|  4|    2|  73|f   |   3152|
|  5|    2|  49|f   |   2473|
|  4|    2|  73|f   |   3152|
|  5|    2|  49|f   |   2473|
|  2|    2|  54|m   |   2474|

Table: Spatial microdata generated for SimpleWorld zone 2. Each row represents an individual taken from the individual-level sample and the 'zone' column represents in where in geographical space they reside.

The Table 3.5 is a reasonable approximation of the inhabitants of zone 2: older
females dominate in both the aggregate (which contains 8 older people and 6
females) and the simulated spatial microdata (which contains 8 older people and
6 females). Note that in addition the constraint variables, we also have
an estimate the income distribution in SimpleWorld's second zone.

Towards the end of Chapter 4 we will learn how to generate
this table from first principles. The remainder of this section
considers how the outputs of spatial microsimulation, in the context of
SimpleWorld, can be useful before progressing to the practicalities.

```{r, echo=FALSE}
# Add section link here!
```

## SimpleWorld in context {#SimpleWorldContext}

Even though the datasets are tiny in SimpleWorld, we have already generated some
useful output. We can estimate, for example, the average income in each zone.
Furthermore, we could create an estimate of the *distribution* of income in each
area. Although these estimates are unlikely to be very accurate due to the
paucity of data, the methods could be very useful if performed on larger
datasets from 'RealWorld' (planet Earth).  Finally, the spatial microdata
presented in the above table can be used as an input into an agent-based model
(ABM). Assuming the inhabitants of SimpleWorld are more predictable than those
of RealWorld, the outputs from such a model could be very useful indeed, for
example for predicting future outcomes of current patterns of behaviour.

In addition to clarifying the advantages of spatial microsimulation, the above
example also flags some limitations of the methodology. Spatial microsimulation
will only yield useful results if the input datasets are representative of
the population as a whole, and for each region. If the relationship between age and
sex is markedly different in one zone compared with what we assume to be the
global averages of the input data, for example, our estimates could be way out.
Using such a small sample, one could rightly argue, how could the diversity of
33 inhabitants of SimpleWorld be represented by our simulated spatial microdata?
This question is equally applicable to larger simulations. These issues are
important and will be tackled in a subsequent Section (\ref{svalidation}) on
validation.
