# What is spatial microsimulation? {#what-is}

The purpose of this chapter is to demystify
the various interpretations and uses of the term 'spatial microsimulation'
and to define clearly what we mean by it in this book.
Following the brief introduction to the field in this first section,
the chapter is ordered as follows:

- *Terminology* (\@ref(terminology)) explains the basis of the concept (\@ref(SimCity))
and shows how spatial microsimulation
can be understood as either a method or an approach
(\@ref(meth-approach)).
- *What spatial microsimulation is not* (\@ref(is-not)) describes the differences between 
spatial microsimulation and other fields that could seem similar.
- *Applications* (\@ref(applications)) explores various academic and real-world
applications.
- *Assumptions* (\@ref(assumptions)) addresses
the often unspoken assumptions underlying spatial microsimulation results.

For the purposes of this book spatial microsimulation was defined in Chapter&nbsp;1 as:

> **The creation, analysis and modelling of individual level data allocated to geographic zones.**

Spatial microsimulation is well-suited to the analysis of complex phenomena which happen over
geographical space, such as transport systems and housing markets.
Because it includes the creation of synthetic data, the method is well-suited
to situations in which available data are limited. Figure 3.1 illustrates how
the process of *population synthesis* can be used to impute missing data,
by approximating the original geo-referenced individual level data
[@Lovelace-ipfinr]. It is important to note that the process does not stop with the
generation of spatial microdata: it involves *doing stuff* with the spatial microdata
to better understand the world.

```{r, fig.cap="Schematic of population synthesis, a critical element in spatial microsimulation", fig.scap="Schematic of population synthesis", fig.pos='!h', echo=FALSE}
img <- readPNG("figures/msim-schema.png")
grid.raster(img)
```

Spatial microsimulation is a way to combine the advantages of individual level
data with the geographical specificity of geographical data. If used correctly,
it can be used to provide 'the best of both worlds' of available data by
combining them into a single analysis. Spatial microsimulation should therefore
be used in situations where you have access to individual and area level data but
no means of combining them. Spatial microsimulation can also be used in
contexts where no individual level data is available, as described in Chapter&nbsp;9.

Typical use cases include modelling the spatial distribution of population
growth and changes in demographics; high level transport-modelling; scenario-based
modelling of behavioural change; and as an input into agent-based models.
Because large individual and area level datasets have only recently become available
and because computers have only recently become sufficiently powerful to run
large-scale models on thousands of individuals and zones, spatial microsimulation
is still a field in its infancy. There are many areas of application where
the method has considerable benefits but where the method has not yet been
used. It is hoped that this book provides sufficient guidance to enable the
reader to use spatial microsimulation on new problems.

## Terminology {#terminology}

Like many phrases, the meaning of *spatial microsimulation*
varies depending on context. The term is more ambiguous than most,
however, because it is a relatively new method and because different
people have used the term in different ways.
Of course, the label attributed to the method is less important than the
method itself. As eminent physicist Richard Feynman
put it, "the difference between knowing the name of something and
knowing something" is vital to understand the world. The
methods used in this book could equally be "multi level modelling" or 
"real-world SimCity",
but this would not change how its methods work or what they do. However,
it is important that the terminology we use is at least *internally consistent*
to avoid confusion. Furthermore, it is important to understanding something about
how others have used 'spatial microsimulation' to avoid
misinterpretation of literature that employs the term.

There are some issues with term *spatial microsimulation*, as
described below
(see \@ref(is-not)).
However, *spatial microsimulation* is an appropriate label for the material covered in this
book because it is
already widely used and because it
succinctly conveys the main elements of the approach:

- **Spatial** microsimulation is inherently concerned with how things vary
over space, not just between individuals, groups or periods of time: this is
what distinguishes *spatial* microsimulation from the wider field of microsimulation.
- Spatial **micro**simulation explores issues at the individual level, as implied
by the word *micro*.
- Spatial micro**simulation** involves the creation of fictitious data for
modelling purposes, captured by the word *simulation*.

From this breakdown of spatial microsimulation into its component parts its meaning
may seem obvious.
However, it is important to define the term precisely at an early stage
and understand how other people have used the term
to avoid confusion.

### Spatial microsimulation as SimCity {#SimCity}

SimCity is
a popular computer game series in which the player constructs urban infrastructure
and observes his God-like influence on the virtual citizens. 
The analogy of SimCity helps to describe spatial microsimulation.
In practice, however, the underlying aims of SimCity (entertainment,
education and profit for its publisher) are quite different from those of
spatial microsimulation research.
But, in some ways the comparison is appropriate:
SimCity creates virtual individuals allocated to geographic space and provides
a framework for *model experiments* in the same way that spatial microsimulation
does. SimCity can be used for teaching urban planning
[@Gaber2007] and illustrates how complex computer simulations of urban
systems can become. A number of open-source versions of
the SimCity concept are now available (e.g. LinCity-NG, Micropolis and Simutrans).

### Spatial microsimulation: method or approach? {#meth-approach}

```{r, echo=FALSE}
# The aim of this section is to show how the term is used in the acadmic
# literature, implicitly and via explicit definitions
```

The most common confusion about what spatial microsimulation is arises because
the term has been used to refer to both a narrow
methodology and a broader research approach. From
Feynman's distinction between a thing's label and what
it actually is, it is clear that both interpretations are acceptable.
The critical step is to use the term in a way that others can understand,
remembering that the audience may not have heard of spatial microsimulation before.
As stated in the introduction, we acknowledge both uses of the term in the
literature but advocate that 'spatial microsimulation' is used to describe
the overall modelling approach. The term *population synthesis*, used in
transport modelling, is used to describe the methods for generating
the spatial microdata on which the spatial microsimulation approach depends.

A subsequent section
(\@ref(applications))
outlines some real-world problems spatial microsimulation has
been used to tackle. First, we consider how spatial
microsimulation is understood in the literature and how this relates to
the definitions used in this book.

**Population synthesis**
is a set of techniques for generating individual level data
allocated to geographical zones. Population synthesis is an important
(and often crucial) component of the spatial microsimulation approach,
the aim of which is to
generate a realistic sample for each area that is as similar as possible to
aggregate level constraints. Population synthesis usually involves the allocation of individuals from a survey dataset
to administrative zones, based on shared variables between the areal (where each unit is a zone) and
individual level data. When
additional *target variables* exist in the microdata inputs for population synthesis (which are not present in the aggregate level data), the process
can be used to simulate information that is not otherwise available at the local
level. Population synthesis in this context can be seen as part of the
long-established field of *small area estimation* [@Rao2003].

**Microsimulation** (of which spatial microsimulation is a subset)
is an approach that was first conceived by
Guy Orcutt. This can be defined in general terms as
"a methodology ...
to simulate the states and behaviours of different
units --- e.g. individuals, households, firms --- as
they evolve in a given environment" 
[@Baroni2007]. The defining feature of *spatial microsimulation* is that
the 'environment' is defined in predominantly geographical terms: the
individuals are allocated to small parcels of land which affect their
characteristics and inferred behaviour.
This wider perspective helps explain why,
despite not using the term 'spatial microsimulation',
@Orcutt1957-new-type is frequently cited as one of the founding fathers of
the field.

As with many new and infrequently used words, the term
spatial microsimulation is a source of confusion, and its meaning can vary
depending on context and who you ask. To an
economist, spatial microsimulation is likely to imply
modelling a temporal process such as how individual agents in
different areas respond to changes in prices or policies. To a transport
planner, the term may imply simulating the precise movements of vehicles on
the transport network. To your next door neighbour it may mean you have
started speaking gobbledygook! Hence the need to consider what spatial
microsimulation is, and what it is not, at the outset. However, in every case, 
the term involves the creation of individual level data that is grouped by 
geographic zone via some kind of approximation method.



```{r, echo=FALSE}
#### Communicating spatial microsimulation {#comms}
# TODO add something about the distinction between spatial microsimulation 
# and small area estimation - e.g.:
# "There is much overlap between spatial microsimulation and small area estimation and the terms are sometimes used
# interchangeably. It is useful to make the distinction, however, as small area estimation refers to methods to estimate summary statistics for geographical units—sometimes via a spatial microdataset. With spatial microsimulation, on the other hand, the emphasis is on generating and analysing spatial microdata"

# 
# Delving a little into the etymology and history of spatial microsimulation
# reveals the reasons behind the various meanings. Rarely will you be understood saying
# "I use *spatial microsimulation*" in everyday life. Usually it
# is important to add context. Below are a few hypothetical situations and
# suggested responses.
# 
# -   When talking to a colleague, a transport modeller: spatial
#     microsimulation, also known as population synthesis..."
# 
# -   Speaking to agent-based modellers: "we use spatial microsimulation
#     to simulate the characteristics of geo-referenced agents..."
# 
# -   Communicating with undergraduates who are unlikely to have come
#     across the term or its analogies. "I do spatial microsimulation, a
#     way of generating individual level data for small areas..."
# 
# -   Chatting casually in the pub or coffee shop: "I'm using a technique
#     called spatial microsimulation to model people...".
# 
# The above examples illustrate potential for
# confusion: care needs to be taken to use terminology each audience understands. The transport modeller, for example, may
# already know that the term 'population synthesis' means creating an individual level
# dataset of real areas, whereas more basic terms need to be used when communicating
# the method to policy makers. All this links back to the importance
# of transparency and reproducibility of method discussed in the previous Chapter:
# avoid implying spatial microsimulation is something it is not.
# 
# Faced with uncomprehending stares when describing the method, some may
# be tempted to 'blind them with science', relying on
# sophisticated-sounding jargon, for example by saying: "we use simulated
# annealing in our integerised spatial microsimulation model". Such
# wording obscures meaning (how many people in the room will understand
# 'integerised', let alone 'simulated annealing') and makes the process
# inaccessible. Although much jargon is used in the spatial
# microsimulation literature and in this book, care must be taken to ensure
# that people understand what you are saying.
# 
# This raises the question, why use the term spatial microsimulation at
# all, if it is understood by so few people? The answer to this is that
# spatial microsimulation, defined clearly at the outset and used
# correctly, can concisely describe a technique that would otherwise need
# many more words on each use. Try replacing 'spatial microsimulation'
# with 'a statistical technique to allocate individuals from a survey
# dataset to administrative zones' each time it appears in this book and
# the advantages of a simple term should become clear! 'Population
# synthesis' is perhaps a more accurate term but, transport modelling
# aside, the literature already uses 'spatial microsimulation'. Rather
# than create more complexity with *another* piece of jargon, we proceed
# with the term favoured by the majority of practitioners.
# 
# Why has this situation, in which practitioners of a statistical method
# must tread carefully to avoid confusing their audience, come about?
# First it's worth stating that the problem is by no means unique to this
# field: imagine the difficulties that Bayesian statisticians must
# encounter when speaking of prior and posterior probability distributions
# to an uninitiated audience. Let alone describing Gibb's sampling. Describing complex methods is 
# a constant challenge for the scientist. Using jargon when the audience 
# is experienced and simplifying when they are not is the art of science communication.
# 
# As outlined above, disassembling the term provides
# detect three pieces of information about the approach.
# First, it is a "simulation", meaning that
# we can not reach the real needed data, so we make an approximation helped 
# with available data. Second, "micro" reflects the desire to simulate
# the data at the individual level. This captures the ability of spatial microsimulation to 'zoom in' from aggregated data to a higher resolution data where all the characteristics of all individuals are simulated.
# Finally, "spatial" indicates that
# each individual is allocated to a zone, that depending on the context
# can correspond to a postcode, a municipality, a district, a country etc. Usually these zones are geographical.
```



```{r, echo=FALSE}
# # Commented out for now as it risks confusing the reader
# In each application of spatial microsimulation we refer to a population to be simulated. For example, a population can be a set
# of bulbs, if we analyse the quality of a new production process of bulbs. 
# And their characteristics could be their life duration and their power. 
# Moreover, we need the data in an individual level and the location of each
# of these entities. In our example, we want to know that the first bulb
# has a power of 60Watt, a life duration of 200 hours and was created in 
# the firm of Texas.
# 
# ### Spatial microsimulation in this book {#this-book}
# 
# As we have seen, the term "spatial microsimulation" has a number of meanings incorporating a range of methods and applications. In this book, the focus is on the
# most used part of this branch, called reweighting: the process of generating the spatial microdata. From 
# [](#CakeMap)
# onward, we also explain what can be done with the spatial microdata that we have generated in the preceding chapters. The technique described in this book will be most useful in contexts where you
# have categorical geographically aggregated count data about the entire target population (these are the *constraint variables*, often taken from a national census) and an individual level survey based on a sample of the target population (the non-spatial *microdata*).
```

To avoid confusion regarding the terminology used in this book,
a glossary defining much of the jargon relating to spatial microsimulation
is provided at the end. For now, to
help answer what spatial microsimulation is we will look at 
its applications and then at what it is not.

## What spatial microsimulation is not {#is-not}

Having seen contemporary definitions of spatial microsimulation and what it *is*,
it is also useful to define spatial microsimulation negatively, in terms of
what it is not. This is partly due to the close association between spatial
microsimulation and other methods, but also because there is a tendency
for people to think that spatial microsimulation is more complicated than
it is.

**Spatial microsimulation is not small area estimation**

Small area estimation consists in estimating aggregate counts for
a small area. For example, in this field, we can forecast 
the total population of a zone for a future year. However, 
we have no information about each individual, it is restricted
to statistics on the area. On the other hand, spatial microsimulation
really focuses on the *micro* level. Thus, we estimate the population
individual per individual. 

Note that, thanks to spatial microsimulation, 
we are able to aggregate counts and deduce the *macro* level
per municipality, which corresponds to small area estimation. 


**Spatial microsimulation is not (quite) agent-based modelling**

Spatial microsimulation does involve the creation and analysis of individuals 
and their allocation to families and zone.
But it does not necessarily imply interaction between these individuals.
For this,
agent-based model (ABM) is needed. One could assume that because
the method contains the word 'simulation', it includes detailed
modelling of individual
behaviours in which individuals interact with each other and the environment
over time and space. This is not always the case: spatial microsimulation
is generally a more 'top down' approach to modelling, in which the results
can be broadly predicted. ABM, by contrast, is bottom-up and can result in
highly non-linear and chaotic states [@Batty2005].

There is, however, no clearly defined boundary stating where
spatial microsimulation ends and ABM begins and the two approaches are closely
linked. The synthetic populations produced as part of spatial microsimulation
can form an excellent starting point for ABM.
ABM can be seen as an extension of spatial microsimulation.
While spatial microsimulation *produces* individuals and assigns their
characteristics over space (and en masse via various 'what-if' scenarios),
ABMs tend to have higher spatial and temporal resolution,
allowing individuals to interact through space
and time, with each other and with their environment. This increased
level of detail and complexity means that ABM tends to have higher
computational needs per individual. As a result, spatial microsimulation
models tend to be much larger, encapsulating up to millions of individuals.
As computing power continues to increase the potential for adding ABM
capabilities to such models is only set to grow [@Wegener2011].

The above discussion illustrates that
microsimulation can be seen as a subset of advanced ABM.
The results of spatial microsimulation models tend to apply to only specific
snapshots in time and individuals tend to be fixed to a single area.
ABM can thus add great additional value to spatial microsimulation models,
by providing a framework for more complex interactions.
Chapter 12
illustrates how the outputs of spatial 
microsimulation can form an empirical basis for ABM.

In summary, ABM and spatial microsimulation are closely related, overlapping and
complementary approaches to the analysis of individual level processes
operating over geographical space and time. A main conceptual difference is
that the spans of space and time tend to be larger in spatial microsimulation
work, for reasons relating to computing power and model complexity.


```{r, echo=FALSE}
# TODO: (MD)
# add a references to chapter about households when created.
```

```{r, echo=FALSE}
# (MD) For the moment, I put this in comment, because it does not sound so clear for me (and reviewer 2)
# Spatial microsimulation as an approach to modelling involves
#the allocation of individuals to specific zones, what-if scenarios, and, in *dynamic spatial microsimula#tion*, a time
#step to represent the individuals' evolution over time. In *static spatial microsimulation* studies that #rely on *what-if* scenarios, time is not 
#really included. 

# # removed as this section was getting long! (RL)
# 
# The aims of the two methods are complementary.
# ABM requires an initial population which spatial microsimulation can provide and adds more complexity and interactivity. The rules of the ABM make each agent evolve through time and space 
# depending in a bottom-up process. Spatial microsimulation is, by contrast, more of a *top down* process. Spatial microsimulation can be closely linked to ABM. 
```



**Spatial microsimulation does not really generate new data**

During spatial microsimulation, apparently new individuals are created and
placed into zones. It would be tempting to think that new information
about the world is somehow
being created. This is not the case: the 'new' individuals are simply repeats
of individuals we already knew about from the individual level data, albeit
in a different order and in different combinations. 
Thus we are not increasing the diversity of the dataset,
simply changing its aggregate level characteristics. 
Spatial microsimulation creates a complete data that take into
accounts all other data you included in the process. It
is just a way to put all information together to approximate
the whole real population, but it does not really generate new data.

**Spatial microsimulation is often not strictly spatial**

The most surprising feature of spatial microsimulation
is that the method is not strictly *spatial*. The only
reason why the method has developed this name (as opposed to 'small
area population synthesis', for example) is that practitioners tend
to use administrative zones, which represent geographical areas, as the
grouping variable. However, any mutually exclusive grouping variable,
such as age band or number of bedrooms in your house, could
be used. Likewise, geographical location can be used as a *constraint variable*.
In most spatial microsimulation models, the spatial variable is a mutually
exclusive grouping, interchangeable with any such group. Of course, the spatial
microdata, maps and analysis that result from spatial microsimulation are spatial;
it's just that there is nothing inherently spatial about the method used to
generate the spatial microdata.

To be more precise, spatial microsimulation is not *inherently spatial*.
Spatial attributes such as the geographic coordinates of the zones
to which individuals have been allocated and home and work
locations can easily be added to the spatial microdata after they have been
generated. It is the use of geographical variables as the grouping variable that is
critical here and which distinguishes spatial microsimulation from other types
of microsimulation.

```{r, echo=FALSE}
# **Spatial microsimulation is not always simulation**
```

A common use of spatial microsimulation (at least the population synthesis aspect)
is simply to create model estimates of data which does not exist. This 
usage case is represented in Figure 3.1, whereby individual level data from
a survey is 'scaled down' to the local level using population synthesis algorithms.
As illustrated in Figure 3.1, the process of population synthesis can be seen
as an attempt to reproduce the real spatial microdata collected during the
census but which is unavailable for confidentiality reasons.

Input *microdata* and *constraints* ensure the simulated results match reality
(at least at the aggregate level for the constraint variables --- see
\@ref(assumptions)). The resulting synthetic spatial microdata is extremely useful
for estimating missing data at the local level. If *target variables* contained
in the output were not present in the constraints (income is a common example),
estimates of income variability over space can be extracted from the spatial microdata.
In addition, the estimated spatial microdata represented in Figure 3.1 will contain
estimates of *cross-tabulations* (contingency tables) between different variables
and estimates of the distribution of continuous variables such as age and income.
These estimates are useful in many applications (\@ref(applications)).

## Applications {#applications}

Spatial microsimulation has a wide variety of applications and there
are many areas where the technique has been used.
The main areas of application have
been health, economic policy evaluation and transport. Rather than attempt
to provide a comprehensive account of the range of current and possible
applications, this section describes a single study in each area to exemplify
how spatial microsimulation is used.

### Health applications

A classic example of the potential practical utility of spatial microsimulation
is a study which estimated the rate of smoking at the
small area level in the city of Leeds UK
[@Tomintz2008].
Smoking is a classic 'target variable' in spatial microsimulation:
it is reported in a number of individual level surveys but there is surprisingly
little information about how smoking rates vary from place to place.
Thus it is difficult to determine where to locate services that depend on the rate of smoking.
The synthetic spatial microdata could thus be used to help identify new clinics
to help people stop smoking. (Alternatively, the spatial microdata could be used
by a tobacco chain to help decide where to invest in a new shop, highlighting
the potential misuse of the technique by unscrupulous analysts.) The authors
found that actual anti-smoking clinics were not located optimally. Furthermore,
the results pointed to optimal locations for new clinics, potentially improving
the cost-effectiveness of public health campaigns.

This research has since been 'scaled-up' to estimate smoking rates across the whole
of Austria. The [simSALUD](http://www.simsalud.org/) portal provides users with
access to the resulting spatial microdata and an on-line interface to allow
for the selection of constraint variables and other options to customise the
model for the specific purposes. This portal-based system and the provision
of synthetic spatial microdata to researchers illustrates one possible direction that
spatial microsimulation research could go in, where the synthetic data produced from
a large model is the main output of the research, to be used by others for a variety of
applications.

The example of smoking demonstrates the increase of spatial resolution
that spatial microsimulation can bring to bear on under-studied areas in public
health. Where the prevalence of unhealthy activities is closely related 
to socio-demographic variables, a synthetic microdataset can lead to decision
making tools that would be difficult to implement with non-spatial surveys alone.
**Simobesity** is another research project and spatial microsimulation software tool that
estimates the prevalence of obesity at the local levels depending on demographic
constraint variables
[@Edwards2013].
Recent evidence has emerged on the impact of
car-dependent urban environments on inactive lifestyles and resulting poor health
(these areas have been labelled 'obesogenic'). In this context, there is
great potential for combining socio-demographic and environmental-geographic
variables in a spatial microsimulation model. Using the same principles
described by Tomintz, Clarke and Rigby (2008), the outputs of such a model
could help target local interventions
to tackle physical inactivity, maximising the benefits of public health
initiatives.

```{r, echo=FALSE}
# # TODO: link the reader here to a chapter where smsim models are linked to geo-data
# Another health-related example in which spatial microsimulation serves to link
# different secondary data sources to generate new insight for health applications
# is Deetjen and Powell (forthcoming).
# This model combines individual level survey data on Internet use with census
# data and hospital episode statistics for all of England.
# The aim was to explore the relationship between Internet use, perceived health
# and health service provision.
# 
# The simulated dataset was analysed using hierarchical linear modelling,
# structural equation modelling and cluster analysis. 
# Internet use is often assumed to
# reduce inequalities in access to health services and reduce costs.
# However, quantitative evidence to support this hypothesis is been lacking,
# leading to the adoption of a spatial microsimulation approach.
#  
# Deetjen and Powell (forthcoming) is notable for its use of
# external validation and qualitative data alongside spatial
# microsimulation. The results were compared with the results
# from 3 other data sources (see Chapter 9 of this book).
# Interviews helped
# 'ground-truth' the quantitative findings against real-world experience.
```

### Economic policy evaluation

The social-demographic distribution of impacts arising from
economic policy evaluation is one of the most common applications of
microsimulation (although the analysis in this area is often non-spatial).
'Social impact evaluation', where the
impact of policy changes on different income and socio-demographic groups is
explored, is a classic example of applied microsimulation research. Frequently
these simulations are undertaken by government departments
and focus on overall shifts in the population rather than
spatial variability in the impacts. The EU–funded EUROMOD project,
and software package of the same name,
is the largest of these 
initiatives. The EUROMOD software is used by government analysts and
research agencies in many countries to estimate the distributional impacts
of policy reforms (Figure 3.2).
The resulting research demonstrates that modelling work based on
microsimulation can provide important new lines of evidence
to inform national level policies [@avram2012distributional].

```{r, fig.cap="Output from the EUROMOD economic microsimulation model (Avram et al. 2014). Along the x axes is income group rising to the right. This means, for example, that Latvia (LV) has implemented progressive policies whereas Portugal (PT) has implemented regressive policies. Country acronyms from left to right stand for Estonia (EE), Greece (EL), Spain (ES), Italy (IT), Latvia (LV), Lithuania (LT), Portugal (PT), Romania (RO) and the UK.", fig.scap="Output from the EUROMOD model", fig.width=5, fig.height=5, echo=FALSE}
# A typical output from a microsimulation model for economic policy evaluation, from De Agostini et al. (2014). Each vertical bar represents the relative gains or losses of each income 'vingtile' (there are 20 equally sized groups, by income, in the model) and the black line shows that the austerity budget has been most detrimental to low income groups and most beneficial to the upper-middle classes.
library(png)
library(grid)
img <- readPNG("figures/austerity.png")
grid.raster(img)
```

Spatial microsimulation uses very similar techniques as those employed by
EUROMOD and other economic microsimulation models,
including probability-weighted random sampling of individual level
data and aggregate level scenario development
[@sutherland2013euromod]. 

\pagebreak
The majority of microsimulation
research for economic policy evaluation does not
disaggregate the impacts over space, however. The estimation of variability
at the *local level* is what differentiates spatial microsimulation models
from economic microsimulation models, although the underlying methods are
very similar.
This book does not cover EUROMOD, focussing instead on
spatially disaggregated microdata. However, there is great potential for
future work to make EUROMOD more spatially enabled and to bring elements
of EUROMOD into the methods outlined in the following chapters.

### Transport

Transport modelling is a mature field that increasingly uses individual level
data as the basis of analysis. Large scale models
such as [MATSim](http://www.matsim.org/) rely on spatial microdata to provide
demand for travel and individual characteristics for origins and destinations.
The same techniques are used in spatial microsimulation and transport modelling
 generating spatial microdata although in the transport literature,
 the process is referred to as *population synthesis* [@Muller2010]. 

Generally, little attention is paid to this process of
synthetic population generation in transport modelling
because the focus is on movement of individuals rather
than their characteristics. Distributional impacts
are often overlooked in transport models [@Lucas2012]
and there is much potential to integrate spatial microsimulation
with existing transport modelling methods.

An example of the potential uses of spatial microsimulation
in transport models is illustrated in Figure 3.3. This
shows the simulated commuting behaviour of 20 randomly
selected individuals from a large scale spatial microdataset
of Sheffield. Because the constraints used in this model
included socio-demographic variables, each individual
represented in the figure has a rich profile of characteristics
associated with them. This analysis can provide new evidence
about the likely winners and losers from very specific
interventions such as a new bicycle path or bus route
[@Lovelace2014-jtg]. As a result of the increased detail
allowed by such methods there is much interest in spatial
microsimulation for transport applications. Figure 3.3 also
illustrates the potential for the output of 
spatial microsimulation to be used as an input into agent-based
models (ABM).

```{r, fig.cap="An illustration of spatial microdata in transport modelling. 20 people are illustrated on the map as travelling to a range of destinations, specified based on probability-weighted sampling from origin-destination tables (Lovelace et al. 2014).", fig.scap="An illustration of spatial microdata in transport modelling", fig.width=4, fig.height=4, echo=FALSE}
library(png)
library(grid)
img <- readPNG("figures/jtg.png")
grid.raster(img)
```

A larger and more advanced illustration of the potential for spatial microsimulation
in transport modelling work is described by @Barthelemy2015a. This paper
describes a stochastic model to allocate travel behaviours to a geo-located
synthetic population of 10 million people, representing the entirety of
the transport system in Belgium. By combining the synthetic microdata
with an agent–based modelling approach (described in Chapter 12),
@Barthelemy2015a are able to characterise a very large transport system
in great detail.


## Assumptions {#assumptions}

As with any simulation technique, spatial microsimulation is based on
assumptions, some of which are unlikely to hold in all cases. This
does not preclude spatial microsimulation in cases where the assumptions
do not hold: "It is far better to foresee even without
certainty than not to foresee at all", as Henri Poincaré put it
[@barthelemy2014parallelized].


It is vital, however, that users of spatial microsimulation and
'consumers' of the resulting research understand that the results of spatial
microsimulation are not *real* but a best estimate of the population in a
given area. The danger is that if the assumptions are not understood, incorrect
conclusions will result. It is therefore the duty of researchers using spatial
microsimulation (and other techniques) to clearly state the assumptions on which
the results depend on and the extent to which these assumptions can be expected
to hold in practice. Roughly speaking there are four main assumptions underlying
all spatial microsimulation models:

1. The individual level microdata are representative of the study area.
2. The target variable is dependent on the constraint variables and their
interactions in a way that is relatively constant over space and time.
3. The relationships between the constraint variables are not spatially dependent.
4. The input microdataset and constraints are sufficiently rich and detailed
to reproduce the full diversity of individuals and areas in the study region.

Obviously the real world is complex and many processes are spatially dependent,
invalidating assumptions 2 and 3. The extent to which the relationships between
variables can be deemed to be constant over space is often
unknown. However, there are ways of
checking the spatial dependency of
relationships between multiply variables, not least Geographically Weighted Regression
(GWR). 

These limitations should be discussed at the outset of spatial
microsimulation research, with reference to the input data. To see how spatial
microsimulation simplifies the real world, the next chapter describes
a hypothetical scenario where 33 inhabitants of an imaginary land
are simulated and allocated to three zones
based on a microdataset of only five individuals and two constraint variables.

## Chapter summary

In this chapter we have defined what spatial microsimulation is
and what it is not. Some research and real-world applications were
described, with comments on areas for further work.
The final section on assumptions underlying spatial microsimulation
is in some ways the most important: It
shows the importance of understanding the limitations associated with the method and
the dangers of drawing conclusions from simulated data.
