# Spatial microsimulation in the wild {#CakeMap}

So far the book has explained what spatial microsimulation
is, described its applications and demonstrated how it works.
We have seen something of its
underlying theory and its implementation in R. But how can
the method be applied 'in the wild', on real datasets?

The purpose of this chapter is to answer this question
using real data to estimate cake consumption in
different parts of Leeds, UK. The example is deliberately rather absurd to make
it more memorable. The steps are presented in a generalisable way, to be
applicable to a wide range of datasets.

The input microdataset is a randomized ('jumbled') subset of the
2009 [Dental Health Survey](http://data.gov.uk/dataset/adult_dental_health_survey),
(DHS) which covers England, Northern Ireland and Wales. 1173 variables are available
in the DHS, many of which are potentially interesting
target variables not available at the local level.
These include weekly income, plaque build-up, and
oral health behaviours.
Potential linking variables include socio-economic
classification, and dozens of variables related to oral health.

```{r, echo=FALSE}
# (MD): Could be good to say the proportion of the population 
# included in this survey
```

In terms of constraint variables, we are more
limited: the Census is the only survey that provides
count data at the small area
level in the UK. Thus the 'domain' of available input data,
related to our research question involves two sources:

1. Non-geographical
individual level survey, DHS --- the *microdata*.
2. Geographically
aggregated categorical count data from the census --- the *constraint tables*.

This chapter is structured as follow: 

- *Selection of constraint variables* (\@ref(CakeMapConVar)) describes and justifies
the choice of the constraint variables on this specific example.
- *Preparing the input data* (\@ref(CakePrep)) contains the pieces of code
necessary to load and prepare the data, before performing the spatial microsimulation.
- *Using the `ipfp` package* (\@ref(CakeMapIPFP)) develops the whole process
to create a synthetic population thanks to the R package **ipfp**.
- *Using the `mipfp` package* (\@ref(CakeMIPF)) also includes the whole
process, but with the R package **mipfp**.
- *Comparing methods of reweighting large datasets* (\@ref(CompareMeth)) compares
the different results and describes briefly how to transform one type of result
to the other type.


## Selection of constraint variables {#CakeMapConVar}

As discussed in Chapter 4, we must first decide which variables 
should be used to link the two. We must
select the constraints from available linking variables.

The selection of linking variables should not be arbitrarily preordained
by preconceptions. The decision of which constraints to use to
allocate individuals to zones should be context dependent.
If the research is on social exclusion, for example,
many variables could potentially be of interest: car ownership,
house tenancy, age, gender and religion could all affect the dependent variable.
Often constraint variables must be decided not based on what 
would be ideal, but which datasets are available. The selection criteria
will vary from one project to the next, but there are some overriding
principles that apply to most projects:

1. **More the merrier**: each additional constraint used
will further differentiate the spatial microdata from the input microdata.
If gender is the only constraint used, for example, the spatial microdata will
simply be a repetition of the input microdata but with small differences in
the gender ratio from one zone to the next. If five constraints are used
(e.g. age, gender, car ownership, tenancy and religion), the differences
between the spatial microdata from one zone to the next will be much more
pronounced and probably useful.

2. **Relevance to the target variable**: often spatial microsimulation is
used to generate local estimates of variables about which little geographically
disaggregated information is available. Income is a common example: we have
much information about income distributions, but little information about how
average values (let alone the distribution) of income varies from one small area
to the next. In this case income is the target variable. Therefore constraints
must be selected which are closely related to income for the output to resemble
reality. This is analogous to multiple regression (which can also be used
to estimate average income at the local level), where the correct
*explanatory variables* (i.e. constraint variables in spatial microsimulation)
must be selected to effectively predict the *dependent variable*. As with
regression models, there are techniques which can be used to identify the most
suitable constraint variables for a given target variable.

```{r, echo=FALSE}
# TODO: See section x
# may need to spell-out what constraint categories are for below
```

3. **Simplicity**: this criterion to some extent contradicts the first. Sometimes
more constraints do not result in better spatial microdata and problems associated
with 'over-fitting' can emerge. Spatial microsimulation models based on many
tens of constraint categories will take longer to run and require more time
to develop and modify. In addition, the chances of an error being
introduced during every phase of the project is increased with each additional
constraint. The extent to which increasing the number of constraint categories
improves the results of spatial microsimulation, either with additional
variables or by using cross-tabulated constraints (e.g. age/sex)
instead of single-variable constraints, has yet to be explored. It is therefore
difficult to provide general rules of thumb regarding simplicity other than
'do not over-complicate the model with excessive constraint variables and
constraints'.

So, we always need to reach an equilibrium between these principles. 
Indeed, we have to take into account *enough* and *pertinent* variables,
without making the population synthesis process too complex
(see \@ref(Selecting)).

To exemplify these principles, let us consider the constraint variables available
in the CakeMap datasets. Clearly only variables available both in the individual level and 
aggregate level datasets can be chosen. Suppose our aim is to analyse the consumption 
of cakes depending on socio-demographic variables. Five interesting variables assigned to each
of the 916 individuals are available from the individual level data:

\pagebreak

- 'Car': The number of working cars in the person's household.
- 'Sex' and 'ageband4': Gender and age group, in two separate variables. 
Age is divided into 6 groups ranging from '16--24' to '65--74'.^[R tip:
This information can be seen, once the dataset is loaded, by entering
`unique(ind$ageband4)` or, to see the counts in each category,
`summary(ind$ageband4)`. Because the variable is of type `factor`,
`levels(ind$ageband4)` will also provide this information.]
- 'NSSEC': National Statistics Socio-economic Classification: a categorical
variable classifying the individual's work into one of 10 groups including '97',
which means 'no answer' (`NA`).
- 'NCakes': the target variable, reported number of times that the respondent
consumes cake each week.

All of these variables, except for 'NCakes', have a corresponding constraint
variable to be loaded for the 124 Wards that constitute the Leeds Local Authority
in the UK. In real datasets it is rarely the case that the categories of the
individual and aggregate level data match perfectly from the outset. This
is the first problem we must overcome before running a spatial microsimulation
model of cake consumption in Leeds.

The code needed to run the main part of the example is contained within
'CakeMap.R'.
Note that this script makes frequent reference to files contained
in the folder 'data/CakeMap', where input data and processing scripts
for the project are stored.

## Preparing the input data {#CakePrep}

Often spatial microsimulation is presented in a way that suggests the data
arrived in a near perfect state, ready to be inserted directly into the model.
This is rarely the case; usually, one must spend time 
loading the data into R, dealing with missing values,
re-coding categorical variables and column
names, binning continuous variables and subsetting from the microdataset. In a
typical project, data preparation can take as long as the analysis
stage.  This section builds on Chapter 2 to illustrate strategies for data
cleaning on a complex project. To learn about the data cleaning steps that may
be useful to your data, we start from the beginning in this section, with a real
(anonymised) dataset that was downloaded from the internet.

The raw constraint variables for CakeMap were downloaded from
the Infuse website (http://infuse.mimas.ac.uk/).
These, logically enough, are stored in the 'cakeMap/data/' directory
as .csv files and contain the word 'raw' in the file name
to identify the original data. The file 'age-sex-raw.csv', for example, is the raw
age and sex data that was downloaded. As the screenshot in
Figure 7.1 illustrates, these datasets are rather verbose and
require pre-processing. The resulting 'clean' constraints are saved in files such as
'con1.csv', which stands for 'constraint 1'.

```{r, fig.cap="Example of raw aggregate level input data for CakeMap aggregate data, downloaded from http://infuse.mimas.ac.uk/.", fig.scap="Example of raw aggregate level input data", fig.pos='!h' , echo=FALSE}
library(jpeg)
library(grid)
library(png)
img <- readJPEG("figures/raw-data-screenshot.jpeg")
grid.raster(img)
# ![Example of raw aggregate level input data for CakeMap aggregate data, downloaded from
# http://infuse.mimas.ac.uk/.](figures/raw-data-screenshot.jpeg)
```

To ensure reproducibility in the process of converting the raw data into
a form ready for spatial microsimulation, all the steps have been saved.
Take a look at the R script files 'process-age.R', 'process-nssec.R' and
'process-car.R'. The contents of these scripts should provide insight
into methods for data preparation in R. Wickham
(2014b) provides a more
general introduction to data reformatting.
The most difficult input dataset to deal with, in this example,
is the age/sex constraint data.
The steps used to clean it are saved in 'process-age.R', in the `data/CakeMap/`
folder. Take a look through this
file
and try to work out what is going on: the critical stage is grouping single year
age bands into larger groups such as 16--24.

```{r, echo=FALSE}
# TODO: add more here...
```

The end result of 'process-age.R' is a 'clean' .csv file, ready to be loaded and
used as the input of our spatial microsimulation model. Note that the last
line of 'process-age.R' is `write.csv(con1, "con1.csv", row.names = F)`.  This
is the first constraint that we load into R to reweight the individual level
data in the next section.
The outputs from these data preparation steps are named
'con1.csv' to 'con3.csv'. For simplicity, all these were merged (by
'load-all.R')
into a single dataset called 'cons.csv'. All the input data for this section are
thus loaded with only two lines of code:

```{r}
ind <- read.csv("data/CakeMap/ind.csv")
cons <- read.csv("data/CakeMap/cons.csv")
```

\enlargethispage*{\baselineskip}

Take a look at these input data using the techniques learned in the previous
section. To test your understanding, try to answer the following questions:

- What are the constraint variables?
- How many individuals are in the survey microdataset?
- How many zones will we generate spatial microdata for?

For bonus points that will test your R skills as well as your practical knowledge
of spatial microsimulation, try constructing queries in R that will automatically
answer these questions.

It is vital to understand the input datasets before trying to model them, so
take some time exploring the input. Only when these
datasets make sense (a pen and paper can help here, as well as R!) is it time to
generate the spatial microdata.

As explained in the previous chapters, there are different methods to 
perform a spatial microsimulation. We will here use a reweighting method, 
since we have spatial aggregated data and non spatial individual data, both
containing some common variables. In this category, the more often used and intuitive
method is the IPF. 

We mentioned in the SimpleWorld example that there are two different points of view
of IPF. First, weights are assigned to each individual for each zone (`ipfp` package).
Second, weights are assigned to each possible category of individuals for each zone 
(`mipfp` package). The next sections develop the whole procedure with each package.

## Using the `ipfp` package {#CakeMapIPFP}
### Performing IPF on CakeMap data {#CakeIPF}

The `ipfp` reweighting strategy is concise, generalisable and computationally
efficient. On a modern laptop, the `ipfp` method was found to be *almost 40
times faster* than the 'IPFinR' method (section 4.1;
Lovelace, 2014) over 20
iterations on the CakeMap data,
completing in 2 seconds instead of over 1 minute.  This is a huge
time saving!^[These tests were conducted
using the `microbenchmark()` commands found
towards the end of the 'CakeMap.R' file.
The second of these benchmarks depends on files from
`smsim-course` (Lovelace, 2014), the repository of which can be downloaded from
(https://github.com/Robinlovelace/smsim-course).]

\pagebreak
Thanks to the preparatory steps described above,
the IPF stage can be run on a single line. After the datasets are loaded in
the first half of 'CakeMap.R', the following code creates the weight matrix:

```{r, echo=FALSE}
source("code/CakeMap.R")
```

```{r, eval=FALSE}
weights <- apply(cons, 1, function(x)
  ipfp(x, ind_catt, x0, maxit = 20))
```

As with the SimpleWorld example, the correlation^[The function "cor" of R calculates the 
correlation coefficient of Pearson that measures the force of a linear correlation
between the two variables. Included between -1 and 1, the best value when comparing a
simulation of a theoretical count (this is the case here) is 1.] between the
constraint table and the aggregated results of the spatial microsimulation
can be checked to ensure that the reweighting process has worked correctly.
This demonstrates that the process has worked with an *r* value above 0.99:

```{r}
cor(as.numeric(cons), as.numeric(ind_agg))
```

This value is very close to 1, so we can consider that there is a big linear
correlation between the results and the constraints. We can verify it by plotting
the graph on Figure 7.2.

```{r, echo=FALSE, fig.cap="Scatter plot of the relationship between observed cell counts for all categories and all zones from the census (x axis) and the simulated cell counts after IPF (y axis).", fig.scap="Scatter of simulated and desired counts with ipfp",fig.pos='!h'}
plot(ind_agg, cons, pch=20, xlab="Result of the simulation",
  ylab="Constraint",
  main="Correspondance between the simulation and the constraints")
# I don't think it's necessary to show the code that creates this plot - but happy
# to remove 'echo=FALSE' if need's be (RL)
```

With the best fit, we would have all constraints equal to the simulation, so a 
perfect line (this is why we consider a `linear` correlation). Only few points
are outside the area of the line. Note that we have 1,623,797 inhabitants^[
To obtain it, code `sum(ind_agg)/3`, because you have 3 constraints.] in the
simulation and in the constraints, `con1` and `con2` have 1,623,800, while `con3`
contains 1,623,797. As seen before, with IPF, the result depends on
the order of the constraints. Thus, it is logical to put as last constraint
one really reliable, also in terms of total number of people. 
An alternative is to rescale all constraints to be consistent. This will be done 
in the next section, with `mipfp` and in the comparison for `ipfp`.

For the comparison category per category, we can take the absolute value of 
the differences of the two tables. This gives a table of differences. The worst 
category and zone will be the maximum of this matrix.

```{r}
# Maximum error
max(abs(ind_agg-cons))

# Index of the maximum error
which(abs(ind_agg-cons) == max(abs(ind_agg-cons)), 
      arr.ind = TRUE)
```

The maximum error is now known and corresponds to the zone 84 and 13 
columns of the constraints, which is `Car`. 

This is for the basic validation; a more detailed analysis of the quality 
of the results is present in the next chapter.

### Integerisation {#CakeINT}

As before, weights of the IPF procedure are fractional, so must be *integerised*
to create whole individuals. The code presented in Chapter 4 requires little
modification to do this: it is your task to convert the weight matrix generated
by the above lines of code into a spatial microdataset called, as before,
`ints_df` (hint: the `int_trs` function in 'code/functions.R' file will help).
The spatial microdata generated in 'code/CakeMapInts.R' contains the same information
as the individual level dataset, but with the addition of the 'zone' variable,
which specifies which zone each individual inhabits.

The spatial microdata is thus *multilevel* data, operating at one level on
the level of individuals and at another on the level of zones. To generate
summary statistics about the individuals in each zone, functions must be
run on the data, one zone (group) at a time. `aggregate` provides one
way of doing this. After converting the 'NSSEC' socio-economic class variable
into a suitable numeric variable, `aggregate` can be used to identify the
variability in social class in each zone. The `by =` argument is used to
specify how the results are grouped depending on the zone each individual inhabits:

```{r, echo=TRUE, message=FALSE}
source("code/CakeMapInts.R")
```

```{r, eval=FALSE}
aggregate(ints_df$NSSEC, by = list(ints_df$zone), sd,
  na.rm = TRUE)
```

```
##    Group.1        x
## 1         1 1.970570
## 2         2 2.027638
## 3         3 2.019839
## ...
```

```{r, echo=FALSE}
#(MD) Interpret the sd and say why we calculate it
# MD I do not understand everything here..
```

In the above code the third argument refers to the function to be applied to the
input data. The fourth argument is simply an argument of this function, in this
case instructing the standard deviation function (`sd`) to ignore all `NA` values.
An alternative way to perform this operation, which is faster and more
concise, is using `tapply`:

```{r, eval=FALSE}
tapply(ints_df$NSSEC, ints_df$zone, sd, na.rm = TRUE)
```

Note that operations on `ints_df` can take a few seconds to complete.
This is because the object is large, taking up much RAM on
the computer. This can be seen by asking `object.size(ints_df)` or
`nrow(ints_df)`. The latter shows we have created a spatial microdataset
of 1.6 million individuals! Try comparing
this result with the size of the original survey dataset 'ind'. Keeping an eye
on such parameters will ensure that the model does not generate datasets too
large to handle.

To estimate average cake consumption per zone from the demographic variables we must first create a numeric variable that can be averaged.
Given the levels we can create a new variable as follows:

```{r}
levels(ints_df$NCakes)
cake_levels <- c(0.5, 1.5, 4, 8, 0.1)
ints_df$NCakes_num <- cake_levels[as.numeric(ints_df$NCakes)]
```

From here the average levels of cake consumption can be calculated as follows:

```{r, fig.cap="Average cake consumption per zone inferred from microsimulation model."}
library(tmap)
ints_agg <- group_by(ints_df, zone) %>% 
  summarise(av_cakes = mean(NCakes_num))
load("data/CakeMap/wards.RData")
wards@data <- cbind(wards@data, ints_agg)
qtm(wards, "av_cakes")
```


Next chapter, we move on to a vital
consideration in spatial microsimulation models such as CakeMap: model checking and validation.
First, we develop the spatial microsimulation using `mipfp`.

```{r, echo=FALSE}
save.image("cache-CakeMap.RData")
```

```{r, echo=FALSE}
#(MD) Make it with CO to compare the results?
# (MD) explain better the integerisation and the reason why you calculate the things here above
```


## Using the `mipfp` package {#CakeMIPF}

### Performing IPF on CakeMap data {#CakeIPF}

As in the previous section, the first stage is to load the data.
An additional step required by the `Ipfp` function is ensuring the categories
of `cons` and `ind` correspond. In the variable associated with car ownership,
`ind` contains "1" and "2", but `cons` contains "Car" and "NoCar".

To undertake this operation, we convert the category labels
in the individual level variable `ind` so that they conform to the
variable names of the count data contained in constraint data `cons`. 
Continuing with the car ownership example,
every "1" in `ind$Car` should be converted to "Car" and every "2" to "NoCar".
Likewise, the correct labels of "f" and "m" must be created for the
Sex variable and likewise for the other variables in `ind`.
The code that performs this operation is included in the R file
`CakeMapMipfpData.R` file. Take a look at this file (e.g. by typing
`file.edit("code/CakeMapMipfpData.R")` from RStudio) and note the use of the
`switch` function in combination with the recursive function
`sapply` to *switch* the values of the individual level variables.
The entire script is run below using the `source` command.

```{r}
source("code/CakeMapMipfpData.R")
```

The data are now loaded correctly. This can be verified by
printing the first lines of the individual level data and comparing
this with `colnames(con1)`. Note that this step was not 
necessary for the **ipfp** package. To use **ipfp** the user must encode
the constraints in the right order and only the order counts.
In practice this is a good idea. 
Ensuring same variable names match category labels will
help subsequent steps of validation and interpretation. Moreover, 
the structure of some constraints will have to change (**mipfp** takes
as input a table of $n_var$ dimensions, whereas **ipfp** takes as input
a 2-dimensional dimension, whose columns describe variables) and 
this process will be easier if the names are the same.
To reaffirm our starting point, a sample of the
new individual level data is displayed below.

```{r}
head(ind)
```

Defining the initial weight matrix is the next step. For each zone, we simply 
take the cross table (or contingency table)
of the individual data. It is important to be aware of
the order of the variables inside this table. We have in our case
(NCakes, Car, Sex, NSSEC8, ageband4). This information will be necessary 
for the definition of the target and for the description for `mipfp`.

```{r}
# Initial weight matrix for each zone
weight_init_1zone <- table(ind)

# Check order of the variables
dimnames(weight_init_1zone) 
```

To use the full power of the `mipfp` package, we
choose to add the spatial dimension into the weight matrix
instead of performing a `for` loop over the zones.
For this purpose, we repeat the table for each area.
We define the names of the new dimension and 
create an array with the correct cells and names.

```{r}
# Adding the spatial dimension
# Repeat the initial matrix n_zone times
init_cells <- rep(weight_init_1zone, each = nrow(cons))

# Define the names
names <- c(list(rownames(cons)),
           as.list(dimnames(weight_init_1zone)))

# Structure the data
weight_init <- array(init_cells, dim = 
                c(nrow(cons), dim(weight_init_1zone)),
                dimnames = names)
```

```{r, echo=FALSE}
library(mipfp)
```

An advantage of **mipfp** over **ipfp** is that the algorithm
first checks the consistency of the constraints before reweighting. If 
the total number of individuals in a zone is not exactly
the same from one constraint to another, `mipfp` will
return the following warning:
'Target not consistent - shifting to probabilities!'.
In this case, the cells of the resulting table become probabilities, 
such as the sum of all cells is 1. This means that the algorithm does not
know how many people per zone you want.
This is a very useful feature of **mipfp**: creating weights corresponding to
the correct total population is as simple as
multiplying the probabilities by this number.

An alternative to this is to verify by your own 
the totals per zone and re-scale the constraints that are 
different from your purpose population. In our case, 
for each constraint, we calculate the number of persons
in the zone. For each zone, we compare the totals of 
the different constraints. The result is `TRUE` if the totals
are the same and `FALSE` otherwise.
It gives `nrow(cons)` boolean values (TRUE or FALSE), of 
which we print the table.

```{r}
# Check the constraint's totals per zone
table(rowSums(con2) == rowSums(con1))
table(rowSums(con3) == rowSums(con1))
table(rowSums(con2) == rowSums(con3))
```

Constraints 1 and 2 have exactly the same marginals per zone.
Constraint 3 has 72 zones with different totals. Since NSSEC (`con3`) 
is anomalous in this respect 
we consider the totals of the 
two first constraints to be valid. It is always
important to find the reason of the differences in totals.
For `con3`, the individual could give no answer. This explains
why we have a smaller population in this variable.
To perform (non-probabilistic) reweighting with **mipfp**,
we must first rescale `con3` to keep its proportions but
with updated totals per zone. 

```{r}
# Re-scale the constraint 3
con3_prop <- con3 * rowSums(con2) / rowSums(con3)

# Check the new totals
table(rowSums(con2) == rowSums(con3_prop))
```

The rescaling operation above solved one problem, but there is another,
more difficult-to-solve issue with the input data.
The second and third constraints are simple marginals per zone with
a single, clearly-defined variable each. The first constraint (`con1`),
by contrast, is cross-tabulated, a combination of two variables
(sex and age^[or age $\times$ sex, as cross-tabulations are often represented.]).
We must change the structure of the first constraint
to execute `mipfp`. 

```{r}
# View the content of the first constraint
con1[1:3, 1:8]
```

The results of the above command show that `con1` is, like the other constraints,
a two–dimensional table. 
Because it actually contains 2 constraints per zone, the information contained within
needs to be represented as a three–dimensional array (a cube),
the dimensions of which are zone, sex and age.
This conversion is done by
taking the cells of `con1` and putting them
to a three–dimensional array in the correct order
(corresponding to the order of the 2–dimensional table). The code to
do this is included into the 
file `code/CakeMapMipfpCon1Convert.R` and the resulting array is called
`con1_convert`.

```{r, echo=FALSE}
# TODO: ascertain if this is the best way to do this (RL)
```

```{r}
source("code/CakeMapMipfpCon1Convert.R")
```

All constraints being converted and all the margins being
rescaled, the target and its 'description'
are ready to be used in `Ipfp()`.
The target is the set of constraints and the description 
is the figures of the corresponding dimensions into the weight matrix.
The order of the weight's variables is here important. Remember this order is
(Zone, NCakes, Car, Sex, NSSEC8, ageband4). For this reason, the constraint 1, 
containing (zone, Sex, ageband4) has the description `c(1,4,6)`.


```{r}
# Definition of the target and descript
# For the execution of the mipfp
target <- list(con1_convert,
               as.matrix(con2),
               as.matrix(con3_prop))

descript <- list(c(1,4,6), c(1,3), c(1,5))
```

The remaining step involves executing the `Ipfp` function 
on these data.

```{r}
weight_mipfp <- Ipfp(weight_init, descript, target)
```

Note that `Ipfp` does not
print the `tol` result by default. This was a wise decision
by the developer of **mipfp** in this case: 
more than 100 iterations were needed for the weights to converge.

```{r, echo=FALSE, fig.cap="Scatter plot of the relationship between observed cell counts for all categories and all zones from the census (x axis) and the simulated cell counts after IPF (y axis).", fig.scap="Scatter plot of simulated and desired counts with mipfp", fig.pos='!h'}
layout(matrix(c(1:3,3), 2, 2))
plot(con1_convert, apply(weight_mipfp$x.hat,c(1,4,6),sum), pch=20, xlab="Result of the simulation for constraint 1",
  ylab="Constraint 1",
  main="Simulation vs constraint 1")
plot(as.matrix(con2), apply(weight_mipfp$x.hat,c(1,3),sum), pch=20, xlab="Result of the simulation for constraint 2",
  ylab="Constraint 2",
  main="Simulation vs constraint 2")
plot(as.matrix(con3_prop), apply(weight_mipfp$x.hat,c(1,5),sum), pch=20, xlab="Result of the simulation for constraint 3",
  ylab="Constraint 3",
  main="Simulation vs constraint 3")
# I don't think it's necessary to show the code that creates this plot - but happy
# to remove 'echo=FALSE' if need's be (RL)
```

We can see on Figure 7.3 that the third constraint is perfectly fitted. The first and second 
constraints are globally well fitted, but some points, representing
a specific category in the specific zone, are not so good. Analysis of 
quality of fit is provided in the next chapter.
In the next section we focus on comparing **ipfp** and 
**mipfp** for real-world microsimulation applications involving
large datasets, such as CakeMap.

## Comparing methods of reweighting large datasets {#CompareMeth}

We mentioned in the **ipfp** part of this chapter that the three constraints
do not contain the same number of individuals. To compare 
both methods, we will run a second time `ipfp`, but 
with the rescaled constraints. Note that in case of 
non-consistent constraints, `mipfp` warns the user, 
but `ipfp` does not. Thus, with `ipfp`, it is 
recommended that the user checks constraint totals before reweighting.
Note that instead of the number of iterations, we 
chose to fix the tolerance threshold to $10^{-10}$,
the default tolerance argument used in the `mipfp` function
(remember, such details can be verified by typing `?mipfp`).

```{r}
# Re-running ipfp with the rescaled constraints
cons_prop <- cbind(con1, con2, con3_prop)
weights <- apply(cons_prop, 1,
                 function(x) ipfp(x, ind_catt, x0, tol = 1e-10))

# Convert back to aggregates
ind_agg <- t(apply(weights, 2, function(x) colSums(x * ind_cat)))
```

This section is divided in two parts, one that analyses 
the distance between the results generated by the two
packages and the other that focus on the
time differences.


### Comparison of results

Both results being based on the rescaled data, we can compare them.
The structure of both resulting tables are different, however.
To compare them easily, we transform the `ipfp` final
matrix into an array with the same structure 
as the one of the result of `mipfp`.

```{r}
source("code/ConvertIpfpWeights.R")
```

The total number of people per zone is exactly the same in both 
weights matrices (maximum of differences being of order $10^{-10}$).
The size of the population is in both cases
the one desired^[Note that to affirm this, we checked the total population
in the constraints.].

```{r}
# Comparison of the totals per zone
max( apply(weight_mipfp$x.hat, 1, sum) 
       - apply(weight_ipfp, 1, sum) )
```

To visualize all the weights, we plot 
the different counts per constraint at Figure 7.4. We observe
that the constraint 3 is very well fitted with 
both methods. For the two other constraints, 
some categories in some zones are more apart.
However, both methods seems to return very similar 
weights, since the crosses are included inside the 
circles.

```{r, echo=FALSE, fig.cap="Scatter plot of the relationship between observed cell counts for all categories and all zones from the census (x axis) and the simulated cell counts after IPF (y axis) (ipfp vs mipfp).", fig.scap="Scatter plot of simulated and desired counts (ipfp vs mipfp)", fig.pos='!h'}
layout(matrix(c(1:3,3), 2, 2))
plot(con1_convert, apply(weight_mipfp$x.hat,c(1,4,6),sum), pch=4, ylab="Simulated",
  xlab="Constraint 1", main="Constraint 1")
par(new=TRUE)
plot(as.matrix(cons_prop[,1:12]),as.matrix(ind_agg[,1:12]), pch=1, ann=FALSE,axes=FALSE,col='blue')
legend("topleft",c("ipfp","mipfp"),pch = c(1,4),col=c("blue","black"))
plot(as.matrix(con2), apply(weight_mipfp$x.hat,c(1,3),sum), pch=4, ylab="Simulated",
  xlab="Constraint 2", main="Constraint 2")
par(new=TRUE)
plot( as.matrix(cons_prop[,13:14]),as.matrix(ind_agg[,13:14]), pch=1, ann=FALSE,axes=FALSE,col='blue')
legend("topleft",c("ipfp","mipfp"),pch = c(1,4),col=c("blue","black"))
plot(as.matrix(con3_prop), apply(weight_mipfp$x.hat,c(1,5),sum), pch=4,ylab="Simulated",
  xlab="Constraint 3", main="Constraint 3")
par(new=TRUE)
plot(as.matrix(cons_prop[,15:24]),as.matrix(ind_agg[,15:24]), pch=1, axes=FALSE,ann=FALSE,col='blue')
legend("topleft",c("ipfp","mipfp"),pch = c(1,4),col=c("blue","black"))
# I don't think it's necessary to show the code that creates this plot - but happy
# to remove 'echo=FALSE' if need's be (RL)
```

By analysing the absolute differences between the two 
resulting weights, we discover that the largest
absolute distance is 3.45608e-11.

We can consider that both methods give very similar results, even if
there are both distant from the constraints
for few zones and categories. A longer analysis 
of this is done in the next chapter.

### Comparison of times

The weights produced by **ipfp** and **mipfp**
packages are the same. Thus, we must use other criteria to assess
which is more appropriate in different settings.
For applications involving very large microdatasets and numbers of zones,
the speed of execution may be an important criterion. For this reason, 
the time is an important characteristic to compare
the packages. However, other considerations such as flexibility and
user-friendliness should also be taken into account. The question of
'scalability' (how large can the simulation be before the software crashes)
is also important.

When reweighting individuals for the case
of CakeMap example, `mipfp` was slower than `ipfp`. However, 
when accounting for wider considerations (for example if we replicate the zones
two times) the speed advantage of `ipfp`
is reduced. Tests show that as the size of input microdata increases, any
performance advantage of `ipfp` declines.
This phenomenon should not appear in the `mipfp` package. To demonstrate this
we will run the CakeMap example with variable numbers of individuals 
in the microdata. To do this, `ind` objects with more than 916
persons in the data are created, by copying  the database a second time.
The code used to create new `ind` objects can be seen
by entering `file.edit("code/CakeMapTimeAnalysis.R")` in RStudio
from inside the 'spatial-microsimulation-book' project. 

```{r, fig.cap="Time necessary to perform the generation of the weight matrix depending on the number of individuals inside the microdata.", echo=FALSE}
img <- readPNG("figures/TimeCakeMap.png")
grid.raster(img)
```

Figure 7.5 confirms that for problems with few individuals in the microdata, 
`ipfp` is better. However, having few people means the increase
of the probability to have non–represented categories.
Of course, we can optimise `ipfp` to take only one 
time the individuals that appear several times and add a count 
in the `ind_cat` instead of a "1" (as made in the `indu` variable of SimpleWorld).
However, this process is 
possible only if the added persons are with the same characteristics
as a previous one.

In conclusion, `mipfp` is more flexible and could be used
without microdata, but `ipfp` is faster for problems with
less individuals in the microdata.

These results suggest that to perform population synthesis for spatial
microsimulation in contexts where the input microdataset is large,
one should use `mipfp`. Note that
when having very few microdata, it may be advantageous to 
consider a method without microdata. This way avoids biasing
the results with the (non-random) categories included in the microdata.

The extent to which the input microdata represent the study
area is critical to obtaining valid results:
the results of spatial microsimulation
that use input microdata are highly
dependent on the extent to which the microdata are representative of the 
study area
[@Lenormand2012].

If the used sample 
is too small, or if the sample, for example, contains 90%
female, whereas the whole population has only 51%, the
microdata will not be representative. Usually, to consider a 
sample as representative, each individual should
have exactly the same probability of being in the 
sample. For example, if you make your survey in a university, 
older people have less chance to be inside than younger. In 
this case, the sample is not representative. 

## Chapter summary

In summary, this chapter has coded the whole population synthesis for the
CakeMap example. This reasoning has been developed with two R packages: **ipfp** and
**mipfp**. This allowed us to observe the differences in the structure of the
input data:

1. **ipfp** takes as input the individual data of the sample and the counts
of the constraints included in a two–dimensional array (even if we have several 
constraint variables)
2. **mipfp** considers input in contingency tables. This means that if you have 
5 different table of constraints, each included cross–tabulated variables, **mipfp**
will take into account the 5 tables.

As illustrated in the chapter, we can transform the input of one algorithm into
the input of the other, but **mipfp** is really simpler to use when having several
and cross-tabulated constraints.

We have also seen that **ipfp** is faster for very small problems, but **mipfp**
performs better on large datasets.
