---
title: "Allocating individuals to households"
layout: default
output:
  pdf_document:
    fig_caption: yes
  word_document: default
bibliography: bibliography.bib
---

```{r, echo=FALSE}
library(png)
library(grid)
```


# Allocating individuals to households

So far the book has explored data on 2 levels: the individual level and
the level of administrative zones. The household is another
fundamental building block of human organisation around which key decision-making,
economic and data-collecting activities are centered.

This chapter explains how to take spatial microdata, of the type
we have generated in the previous chapters, and allocate the resulting
individuals into household units.

As for all spatial microsimulation, the method to use
depends on the data available. Having a sample of 
households for which you have information about 
each member is a possibility. Or, you can have
independently data about individuals and households.
In an other case, you have no aggregate data
for the households, but some individuals'
characteristics concern the household of
this person (number of people living with, 
type of households,...).

We will in this book develop a method
for each of these three cases. The first
possibility, having a sample of households
is the topic of next chapter, developed
thanks to a method called TRESIS. 

In this chapter, we focus on the two cases
where you have no microdata for the households.

## Independent data (individuals and households)

The case when you have completely independent data
is rather difficult. In most cases, the individuals 
and households data don't fit to each other in this 
configuration. Data coming from different sources
and with different cleaning processes causes this
inconsistency.

We explain in this section the method
proposed by Johan Barthélemy^[This is a contributed chapter by Johan Barthélemy, 
Associate Research Fellow (Agent-Based Modelling), SMART, University of Wollongong].

The idea is to proceed in three steps. 
First, we determine the individual distribution, 
thanks to for example the package `mipfp` as
explained before. Second, we make a similar
stage but for the household's data. This can
be done by a totally similar way, considering
the households instead of the individuals in
the previous chapters.

Then, individuals' and households' distributions having been 
estimated, the last step of 
the synthetic population generation consists of gathering 
individuals into households. This is done one household 
at a time by first selecting its type before randomly 
drawing its constituent members [@Barthelemy2012]. 

### Household type selection

The household type selection is performed in order to make 
the distribution of the generated synthetic households 
statistically similar to the estimated one, denoted $Hh$. 
This is achieved by choosing the type $hh*$ such that the 
distribution $Hh'$ of the already generated households 
(including the household being built) minimize the 
$\chi^2$ distance between $Hh'$ and $Hh$ i.e:

$$d_{\chi^2}=\sum_{i} \frac{(hh'_i-hh_i)^2}{hh_i^2} $$

where $hh_i$ and $hh_i'$ respectively denote the number 
of households of type $i$ in the estimated and generated 
synthetic population. Note that this optimization is simple 
as the number of household type is limited.

### Constituent members selection

Now that a household type has been determined, we can 
detail the members selection process. First a household 
head is draw from the pool of individual `IndPool` defined 
by the estimated individuals distribution `Ind`. Then, 
depending on the household type a partner, children 
and additional adults are also draw if necessary. 
This process is illustrated in Figure 10.1.


```{r, fig.cap="Constituent members selection process", fig.height=6, echo=FALSE}
img <- readPNG("figures/Jojo.png")
grid.raster(img)
```

Some attributes of the members can be directly 
obtained from their household type (for instance 
the gender of the head for an household of the type 
`Isolated Man`). The remaining missing attributes are then 

- either randomly draw according to some known distributions 
(e.g. the household type x head's gender x head's age x mate's age);
- or if different values are feasible and equally likely then, 
similarly to what is done for the household type selection, 
we retain the one minimizing the $\chi^2$ distance between the generated 
and the estimated individuals distribution.


Once an individual type has been determined, then the 
corresponding member is added to the household being generated:

- if the selected class is still populated in the `IndPool`,
we extract an individual from this class and add it to the household;
- else find a suitable member by searching in the members of 
the households already generated. This last individual is then replaced 
an appropriate one drawn in `IndPool`.


### End of the household generation process

The households generation process ends if all households have been 
constructed or if the pool of individual is empty or if it fails 
to find a member for a household in the previously generated ones.
Note that when the procedure stops, two types of inconsistencies 
may remain in the synthetic population: the final number of 
households and/or individuals may be smaller than estimated.

## References
