---
title: "Allocating individuals to households"
layout: default
output:
  pdf_document:
    fig_caption: yes
  word_document: default
bibliography: bibliography.bib
---

```{r, echo=FALSE}
library(png)
library(grid)
```


# Allocating individuals to households

So far the book has explored data on 2 levels: the individual level and
the level of administrative zones. The household is another
fundamental building block of human organisation around which key decision-making,
economic and data-collecting activities are centered.

This chapter explains how to take spatial microdata, of the type
we have generated in the previous chapters, and allocate the resulting
individuals into household units.

As with all spatial microsimulation work, the appropriate method
for household creation
depends on the data available. Data availability scenarios,
in descending order of detail, include:

- Access to a sample of 
households for which you have information about 
each member.
- Access to a separate datasets about individuals and households, stored in independent
data tables that are not linked by household ID.
-  No access to aggregate data relating to households, but access some
individual-level variables related to the household of which they are a part
(e.g. number of people living with, type of households).

This chapter explains methods
for household-level data generation for the latter two cases. The first
possibility, having a sample of households,
is the topic of next chapter (Chapter 11) on the TRESIS method.
In this chapter, we focus on the two cases
where you have no microdata for the households
(one row of data per household). We begin by considering the
case in which data on households and individuals remain completely separate.

## Independent data (individuals and households)

When the individual-level data are
independent from the household-level data, they can rarely be linked.
Data coming from different sources,
sometimes implying different total populations,
can cause this inconsistency.
This section describes the method
proposed by Johan Barthélemy^[This is a contributed chapter by Johan Barthélemy, 
Associate Research Fellow (Agent-Based Modelling), SMART, University of Wollongong]
for dealing with such situations.

The method is to proceed in three steps. 
First, we determine the individual distribution, 
for example by using the package `mipfp`, as
explained before. Second, determine the distribution
of characteristics for the household's data. This can
be done using the same technique as for the individual-level data, considering
the households instead of the individuals in
the previous chapters.

Third, after individual and household level distributions have been 
estimated, the individuals can be allocated
to households. This is done one household 
at a time by first selecting its type before randomly 
drawing its constituent members [@Barthelemy2012]. 

### Household type selection

The household type selection is performed to ensure
the distribution of the generated synthetic households 
is statistically similar to the estimated one, denoted $Hh$. 
This is achieved by choosing the type $hh*$ such that the 
distribution $Hh'$ of the already generated households 
(including the household being built) minimize the 
$\chi^2$ distance between $Hh'$ and $Hh$ i.e:

$$d_{\chi^2}=\sum_{i} \frac{(hh'_i-hh_i)^2}{hh_i^2} $$

where $hh_i$ and $hh_i'$ respectively denote the number 
of households of type $i$ in the estimated and generated 
synthetic population. Note that this optimization is simple 
as the number of household type is limited.

### Constituent members selection

Now that a household type has been determined, we can 
detail the members selection process. First a household 
head is draw from the pool of individual `IndPool` defined 
by the estimated individuals distribution `Ind`. Then, 
depending on the household type a partner, children 
and additional adults are also draw if necessary. 
This process is illustrated in Figure 10.1.

```{r, fig.cap="Constituent members selection process", fig.height=6, echo=FALSE}
img <- readPNG("figures/Jojo.png")
grid.raster(img)
```

Some attributes of the members can be directly 
obtained from their household type (for instance 
the gender of the head for an household of the type 
`Isolated Man`). The remaining missing attributes are then:

- either randomly draw according to some known distributions 
(e.g. the household type x head's gender x head's age x mate's age);
- or, if different values are feasible and equally likely, 
retained from the distribution which minimizes $\chi^2$ between generated 
and the estimated distributions.
This is similar to what is done for the household type selection.

After an individual type has been determined, then the 
corresponding member is added to the household being generated:

- if the selected class is still populated in the `IndPool`,
we extract an individual from this class and add it to the household;
- else we find a suitable member by searching in the members of 
the households already generated. This last individual is then replaced 
an appropriate one drawn in `IndPool`.

### End of the household generation process

The household generation process ends after any one of three conditions:
if all households have been 
constructed; if the pool of individual is empty; if the process fails 
to find a member for a household in the previously generated ones.
When the procedure stops, two types of inconsistencies
may remain in the synthetic population: the final number of 
households may be smaller than estimated or the number of individuals estimated
may be less than the known population of the area.

## Cross data: individual and household level information^[This
section is based on research funded by the 
Wallonia Region of Belgium. Timoteo Carletti (UNamur),
Eric Cornélis (UNamur), Philippe Toint (UNamur) and Thierry
Eggericks (UCL Louvain-La-Neuve) were involved in the research.
The academic groups of DEMO from UCL-Louvain-La-Neuve and the
OWS (Observatoire Wallon de la Santé) also provided support.]

In some cases, data on households are included in the
individual datasets. For example, individual-level data
may include variables on sex, age and
the number of cohabitants. This provides cross-tabulated information between
the households and the individuals. Considering the number of cohabitants,
IPF can help to obtain, per zone, inhabitants
described by individual-level variables (such as sex, age and income) and some
household-level information (such as household type and household's size).

To deal with this kind of data, we have two possible alternatives: 
1) aggregation results in independent
individual and household level data or 2) a technique is used
to preserve the full contents of the data. The former solution is simpler and 
requires only the first chapters of the book, but results in
a loss of possible precision.
A second possibility, which preserves all the information in individual and
household-level tables, is explained in this section.

In this situation, we usually proceed in two stages. First, we create
the individuals with all their characteristics.
The second step is to group these individuals into households using
combinatorial optimisation. Each
person must be matched to
one and only one household.

For this process there are two possible methods. One assumes
access to household-level variables in the individual-level
data. The other assumes
no access to additional data concerning the structure of the
households such as the age difference between a married couple.
These options are described below.

### Without additional household's data

When household-level constraints are only
contained in the individual's characteristics, 
the aim is to form households where each individual
is contained in one and only one households.
Each individual must respect,
as well as possible, its household's attributes.
Consider the case where the individual-level variables are
age, gender, municipality, education level, professional status,
size and type of households and link
with the household head (e.g. wife, child). A good grouping is one that 
maximises the number of well-fitted constraints.
The perfect
grouping would be one in which each individual respects its size and
type of household and link with the head. In general, it
is impossible to reach a perfect grouping, since the data are not 
perfect. This can happen, for example, if there are an odd number
of people who need to live in couple, making it impossible
to find a perfect coupling.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.7\textwidth]{figures/HH-CO.png}
\caption{Illustration of the combinatorial problem: How to group individuals into households.}
\end{center}
\end{figure}

As illustrated
in figure 10.2, the individuals can be categorised first by
type of household ("married couples with children"" in this case)
and then by size of households. 
This household type has a size of at least 3
(two parents and at least a child).
Inside this restricted set of households, the next step is to
look at the link that each household has with the household head
("Head 1(W)" in this case)
and again split the pool of individuals per link. It is only after this classification
that we proceed to the random draw, respecting the links.
For example, for the married couples, we first draw randomly a head
and then a partner of the opposite gender (the national register
of Belgium for 2011 doesn't contain homosexual couples). Then, depending
on the size of household to be generated, the right number of children
are also drawn. This process ends when no additional household can be drawn 
and respect the constraints. Figure 10.2 shows that we have
a household with head 1, who is a woman; partner 2, who is a man;
and two children: (with ids 2 and 5).

The main sources of error with this method are incoherence 
of the data and the error caused by the process of the IPF before the grouping. 


### With additional household's data

Without additional data on household structure, the only
possible method is the one described in the previous section. 
However, this allows improbable households, such as 
couples between an individual of 18 years old and
another of 81. For this reason, when we create households,
it is often very useful to take into account
the age distributions. We can consider the ages of a
couple, but also of parents and children.

To do this, we need tables of age differences,
with the same categories as in 
the characteristics of the individuals.


## References
