---
title: "Allocating individuals to households"
layout: default
output:
  pdf_document:
    fig_caption: yes
  word_document: default
bibliography: bibliography.bib
---

```{r, echo=FALSE}
library(png)
library(grid)
```


# Allocating individuals to households

So far the book has explored data on 2 levels: the individual level and
the level of administrative zones. The household is another
fundamental building block of human organisation around which key decision-making,
economic and data-collecting activities are centered.

This chapter explains how to take spatial microdata, of the type
we have generated in the previous chapters, and allocate the resulting
individuals into household units.

As with all spatial microsimulation work, the appropriate method
for household creation
depends on the data available. Data availability scenarios,
in descending order of detail, include:

- Access to a sample of 
households for which you have information about 
each member.
- Access to a separate datasets about individuals and households, stored in independent
data tables that are not linked by household ID.
-  No access to aggregate data relating to households, but access some
individual-level variables related to the household of which they are a part
(e.g. number of people living with, type of households).

This chapter explains methods
for household-level data generation for the latter two cases. The first
possibility, having a sample of households,
is the topic of next chapter (Chapter 11) on the TRESIS method.
In this chapter, we focus on the two cases
where you have no microdata for the households
(one row of data per household). We begin by considering the
case in which data on households and individuals remain completely separate.

## Independent data (individuals and households) {#IndData}

When the individual-level data are
independent from the household-level data, they can rarely be linked.
Data coming from different sources,
sometimes implying different total populations,
can cause this inconsistency.
This section describes the method
proposed by Johan Barthélemy^[This is a contributed chapter by Johan Barthélemy, 
Associate Research Fellow (Agent-Based Modelling), SMART, University of Wollongong]
for dealing with such situations.

The method is to proceed in three steps. 
First, we determine the individual distribution, 
for example by using the package `mipfp`, as
explained before. Second, determine the distribution
of characteristics for the household's data. This can
be done using the same technique as for the individual-level data, considering
the households instead of the individuals in
the previous chapters.

Third, after individual and household level distributions have been 
estimated, the individuals can be allocated
to households. This is done one household 
at a time by first selecting its type before randomly 
drawing its constituent members [@Barthelemy2012]. 

### Household type selection

The household type selection is performed to ensure
the distribution of the generated synthetic households 
is statistically similar to the estimated one, denoted $Hh$. 
This is achieved by choosing the type $hh*$ such that the 
distribution $Hh'$ of the already generated households 
(including the household being built) minimize the 
$\chi^2$ distance between $Hh'$ and $Hh$ i.e:

$$d_{\chi^2}=\sum_{i} \frac{(hh'_i-hh_i)^2}{hh_i^2} $$

where $hh_i$ and $hh_i'$ respectively denote the number 
of households of type $i$ in the estimated and generated 
synthetic population. Note that this optimization is simple 
as the number of household type is limited.

### Constituent members selection

Now that a household type has been determined, we can 
detail the members selection process. First a household 
head is draw from the pool of individual `IndPool` defined 
by the estimated individuals distribution `Ind`. Then, 
depending on the household type a partner, children 
and additional adults are also draw if necessary. 
This process is illustrated in Figure 10.1.

```{r, fig.cap="Constituent members selection process", fig.height=6, echo=FALSE}
img <- readPNG("figures/Jojo.png")
grid.raster(img)
```

Some attributes of the members can be directly 
obtained from their household type (for instance 
the gender of the head for an household of the type 
`Isolated Man`). The remaining missing attributes are then:

- either randomly draw according to some known distributions 
(e.g. the household type x head's gender x head's age x mate's age);
- or, if different values are feasible and equally likely, 
retained from the distribution which minimizes $\chi^2$ between generated 
and the estimated distributions.
This is similar to what is done for the household type selection.

After an individual type has been determined, then the 
corresponding member is added to the household being generated:

- if the selected class is still populated in the `IndPool`,
we extract an individual from this class and add it to the household;
- else we find a suitable member by searching in the members of 
the households already generated. This last individual is then replaced 
an appropriate one drawn in `IndPool`.

### End of the household generation process

The household generation process ends after any one of three conditions:
if all households have been 
constructed; if the pool of individual is empty; if the process fails 
to find a member for a household in the previously generated ones.
When the procedure stops, two types of inconsistencies
may remain in the synthetic population: the final number of 
households may be smaller than estimated or the number of individuals estimated
may be less than the known population of the area.

## Cross data: individual and household level information

^[This whole
section is based on research done by Dumont Morgane (UNamur) and funded by the 
Wallonia Region of Belgium. Timoteo Carletti (UNamur),
Eric Cornélis (UNamur), Philippe Toint (UNamur) and Thierry
Eggericks (UCL Louvain-La-Neuve) were involved in the research.
The academic groups of DEMO from UCL-Louvain-La-Neuve and the
OWS (Observatoire Wallon de la Santé) also provided support.]
In some cases, information about households are included in the
individual datasets.
For example, individual-level data
may include variables on type of household or/and
the number of cohabitants in addition to gender and age. This provides cross-tabulated information between
the households and the individuals. Considering the microdataset,
IPF can help to obtain, per zone, inhabitants
described by individual-level variables (such as sex, age and income) and some
household-level information (such as household type and household's size).

To form the households with this resulting data, we have two possible alternatives.
The first is to aggregate the information concerning the individuals and 
the households independently. By this way, we build two independent 
tables and we can use an algorithm similar to the one in [](#IndData).
The second possibility aims in preserving the full potential of the data.
This means that individuals are joined with the constraint to follow
as well as possible their caracteristics. For example, two people
being head can not live together; if a person has 3 cohabitants,
he needs to be in a household of 4 individuals. 
The former solution is simpler and 
requires only the first chapters of the book. However it results in
a loss of possible precision.
The second possibility, which preserves all the information in individual and
household-level tables, is explained in this section.

With cross data, we usually proceed in two stages. First, we create
the individuals with all their characteristics.
The second step is to group these individuals into households using
combinatorial optimisation. Each
person must be matched to
one and only one household.

For this process there are two possible methods. One assumes
access to household-level variables only in the individual-level
data. The other assumes
access to additional data concerning the structure of the
households such as the age difference between a married couple.
These options are described below. Note that in both 
situations, the aim is to form households where each individual
is contained in one and only one households. Each individual must respect,
as well as possible, its household's attributes.

### Without additional household's data  {#WithoutHHdata}

When household-level constraints are only
contained in the individual's characteristics, 
they are often several possible grouping.

Consider the case of our Belgium study where the individual-level variables are
age, gender, municipality, education level, professional status,
size and type of households and link
with the household head (e.g. wife, child). A good grouping is one that 
maximises the number of well-fitted constraints.
The perfect
grouping would be one in which each individual respects its size,
type of household and link with the head. In general, it
is impossible to reach a perfect grouping, since the data are not 
perfect. Indeed, it can happen, for example, that there are an odd number
of people who need to live in couple, making it impossible
to find a perfect coupling.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.7\textwidth]{figures/HH-CO.png}
\caption{Illustration of the combinatorial problem: How to group individuals into households.}
\end{center}
\end{figure}

As illustrated
in figure 10.2, the individuals can be categorised first by
type of household ("married couples with children"" in this case)
and then by size of households. 
This household type has a size of at least 3
(two parents and at least a child).
Inside this restricted set of households, the next step is to
look at the link that each individual has with the household head
and again split the pool of individuals, per link. It is only after this classification
that we proceed to the random draw, respecting the links.
For example, for the married couples, we first draw randomly a head
and then a partner of the opposite gender (the national register
of Belgium for 2011 doesn't contain homosexual couples). Then, depending
on the size of household to be generated, the right number of children
are also drawn. This process ends when no additional household can be drawn 
and respect the constraints. Figure 10.2 shows that we have
a household with head 1, who is a woman; partner 2, who is a man;
and two children: (with ids 2 and 5).

The main sources of error with this method are incoherence 
of the data and the error caused by the process of the IPF before the grouping. 
Howeber, it implicitly consider that each couple is equiprobable independently 
of their characteristics.


### With additional household's data

Without additional data on household structure, the only
possible method is the one described in [](#WithoutHHdata). 
However, this allows improbable households, such as 
couples between an individual of 18 years old and
another of 81. For this reason, when we create households,
it is often very useful to take into account
the age distributions (when these data are available).
We can consider the ages of a
couple, but also of parents and children.

To do this, we need tables of age differences,
with the same categories as in 
the characteristics of the individuals. To explain 
the process, we develop here the methodology used 
for the creation of the couples. This means that we have 
men and women of different ages and role in the household (head or spouse)
and that we need to form the couples. The random draw executed when not 
having additional data will be improved by considering the
real age distributions. Imagine that a part of the additional data is 
the one in table \ref{fictivageDistrib}.

\begin{table}[h!]
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
Municipality & Age woman & Age man & Count\\
\hline
\hline
Namur & 20-25 & 15-20 & 4\\
\hline
Namur & 20-25 & 20-25 & 25\\
\hline
Namur & 20-25 & 25-30 & 18\\
\hline
Namur & 20-25 & 30-35 & 8\\
\hline
Namur & 20-25 & 35-40 & 2\\
\hline 
... & ... & ... & ...\\
\hline
\end{tabular}
\caption{Example of an age distribution table for the couple without children}
\label{fictivageDistrib}
\end{center}
\end{table}

Note that this is a fictive table, non corresponding to the 
Namur data, just to explain the reasoning. Thanks to this
table, we know that to fit the real population, we will need 25 
couples of man and woman in the same age class 20-25, etc.

However, these data being not perfect, the marginals are incoherent 
with the current synthetic population. For this reason, we will consider 
the new data only as proportions. For our example, 
it means that in the total of women having 20-25 years old 
(57 individuals), $\frac{4}{57}=0.07=7$\% are married with a 
man of age 15-20. With this reasoning, we can have the new table \ref{fictivageDistrib2}, 
with a supplementary column considering the proportions.

\begin{table}[h!]
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
Municipality & Age woman & Age man & Count & Proportion of men married \\
&&&&with a woman of this age\\
\hline
\hline
Namur & 20-25 & 15-20 & 4 & 7\% \\
\hline
Namur & 20-25 & 20-25 & 25 & 43,9\% \\
\hline
Namur & 20-25 & 25-30 & 18 & 31,6\% \\
\hline
Namur & 20-25 & 30-35 & 8 & 14\% \\
\hline
Namur & 20-25 & 35-40 & 2 & 3,5\% \\
\hline 
... & ... & ... & ... & ...\\
\hline
\end{tabular}
\caption{Example of an age distribution table with the proportions per women age}
\label{fictivageDistrib2}
\end{center}
\end{table}

These proportions will be useful in one step of the global process.

The methodology for the male heads is illustrated in figure 
\ref{AlgoCouple}. For
female heads, the process is totally similar. First, 
we separate the set of 
individuals depending on their role and gender. 
This forms male heads and female 
partners to join on one hand; and female heads 
with male partners on the other hand. We consider
each head turn by turn. For each head, we
determine the theoretical distributions of each women ages, 
depending in the age of the head
(thanks to the additional age distribution table). 
Out of this distribution, we 
remove the ages that are no more available in the set of 
possible partners. Indeed, at the end of the process, 
we have only few partners to assign. Thus, 
all ages will not be represented anymore.
Out of this distribution of ages, we calculate the
proportions, whose will be used as probabilities in the random draw.
Then, thanks to this, we draw an age. Finally, knowing
the age of his partner, we choose a wife randomly. This
process is repeated until a set of remaining individuals
is empty, or no remaining partner is the possible age for the
remaining heads.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth]{figures/IllustrationCouples.png}
\caption{Illustration of the algorithm to form the couples}
\label{AlgoCouple}
\end{center}
\end{figure}

This algorithm has been applied to all municipalities in Belgium.
The final result is illustrated in figure \ref{ResultBE}. On this graph, 
each point corresponds to a combination of (age woman x age man) for a 
municipality. Its abscissa is the theoretical count for this category, 
included in the database of the age distributions inside couples. 
Its ordinate is the number of couples in this category in our 
synthetic population. Since the dots are approximately on the line
formed by the points having both coordinates equal, our
simulation worked well.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth]{figures/HHCouplesBelgium.png}
\caption{Illustration of the results for Belgium}
\label{ResultBE}
\end{center}
\end{figure}

Note that we could make a simulated annealing to resolve these kind of problems.
In our case, we have tested it, but it is very long to obtain a result
as good as the one shown above. However, in cases where 
the objectives are different, it is possible that a simulated annealing becomes
better. Indeed, for our purpose, it worked, but only the computational time
was not sufficient. If you would like to fit age distributions, 
diploma distributions, and more complicated cases, the simulated annealing could
be a good option.

## References
