---
title: "Visualising and analysing spatial microdata"
layout: default
bibliography: bibliography.bib
---

```{r, echo=FALSE}
# This is an early draft of a book on spatial microsimulation, for teaching purposes
# Introduction {#Introduction}
library(png)
library(grid)
```

```{r, echo=FALSE}
# TODO:
# Implement what-if scenarios
# Gini indices of income for each area
# Creating tables and summaries per zone with dplyr
```


# Visualising and analysing spatial microdata {#vis}

There is clearly a huge amount of information contained in
spatial microdata. The results created in the CakeMap
example operate at multiple levels, making their analysis
and visualisation problematic. It is challenging enough to
represent a continuous distribution of incomes on in a single
graphic, for example. Trying to do so for every one of hundreds
of geographical zones is almost impossible using standard
methods of visualisation in GIS. Moreover, extracting important
summary statistics is also tricky from spatial microdata
because of the multi-level nature of the data and
the sheer number of rows, zones and
variables to choose from.

This chapter tackles each of these issues. It is the first
chapter in the book that assumes the reader has
access to spatial microdata. (Previous chapters describe the generation
of spatial microdata.) The methods
are therefore equally applicable to synthetic spatial microdata as
for real microdata. With the open data movement, it is possible
that more and more individual-level,
geocoded data is released. After all, there is a strong argument
to suggest that the benefits of releasing such data
far outweigh any
potential privacy issues (especially in the age of smartphones
which multinational companies can use to track our every move).

When real spatial microdatasets
become available, previous work on synthetic spatial microdata
should be easy to adapt. Thus, even if they are not 100%
reliable, synthetic spatial microdatasets can serve as a useful
test case against which multi-level models can be tested. In
addition, by demonstrating the benefits of synthetic spatial
microdata for evidence-based policy, spatial microsimulation
research encourages governments to release more
individual-level information. Clearly, real spatial microdata should only
ever be released in a way that would not 
compromise security, ethics or citizen trust.

This chapter starts simply, with a basic description of how to
visualise geographic data in R using the **ggmap** package. 
Based on the spatial microdata generated from the CakeMap
example, we will progress to see how to convert long
spatial microdatasets into graphical outputs. The final
stages focus on analysis and interpretation of 
spatial microdata. 

## Visualisations {#CakeVis}

Visualisation is an important part of communicating quantitative
data, especially so when the datasets are large and complex so not
conducive to description with tables or words.

Because we have generated spatial data, it is useful to create a map of the
results, to see how it varies from place to place.  The code used to do this
found in 'CakeMapPlot.R'. A vital function within this script is the
`inner_join` function, which depends on the **dplyr** package.


```{r, eval=FALSE}
wardsF <- inner_join(wardsF, wards@data, by = "id")
```

The above line of code by default selects all the data contained in the first
object (`wardsF`) and adds to it new variables from the second object based on
the linking variable.  Also in that script file you will encounter the function
`fortify`, the purpose of which is to convert the spatial data object into a
data frame.  The final map result of `CakeMapPlot.R' is illustrated below.

```{r, fig.cap="CakeMap results: estimated average cake consumption in Leeds", fig.width=5, fig.height=4, echo=FALSE}
library(png)
library(grid)
img <- readPNG("figures/CakeMap-lores.png")
grid.raster(img)
```

## Analysis of spatial microdata {#CakeAnalysis}

Once a spatial microdataset has been generated that we are happy with, we will
probably want to analyse it further.  This means exploring it --- its main
features, variability and links with other datasets. To illustrate this process
we will load an additional dataset and compare it with the estimates of cake
consumption per person  generated in the previous section at the ward level.  

The hypothesis we would like to test is that cake consumption is linked to
deprivation: More deprived people will eat unhealthily and cake is a relatively
cheap 'comfort food'.  Assuming our simulated data is correct ---  a
questionable assumption but lets roll with it for now --- we can explore this at
the ward level thanks to an official
[dataset](http://www.neighbourhood.statistics.gov.uk)
on modelled income from neighbourhood statistics. The following code is taken
from the 'CakeMapPlot.R' script.

Because the income dataset was produced for old ward boundaries (they were
slightly modified for the 2011 census), we cannot merge with the spatial dataset
based on the new zone codes. Instead we rely on the name of the wards. The code
below provides a snapshot of these names and demonstrates how they can be joined
using `inner_join`.

```{r, eval=FALSE}
wards@data <- inner_join(wards@data, imd)
summary(imd$NAME %in% wards$NAME)
##       Mode   FALSE    TRUE    NA's 
##    logical      55      71       0 
```


The above code first joins the two datasets together and then checks the result
by seeing how many matches names there are.  In practice the fit between old
names and new names is quite poor: only 71 out of 124. In a proper analysis we
would have to solve this problem (e.g. via the command `pmatch`, which stands
for partial match).  For the purposes of this exercise we will simply plot
income against simulated cake consumption to gain a feeling what it tells us
about the relationship between cake consumption and wealth.

```{r, fig.cap="The relationship between cake consumption and income according to the synthetic microdata", fig.width=5, fig.height=4, echo=FALSE}
grid.raster(readPNG("figures/incomeCake.png"))
# **Scatterplot** illustrating the relationship between modelled average ward
# income and simulated number of cakes eaten per person per week.
```

The question raised by this finding is: why?  Not why is cake consumption higher
in wealthy areas (this has not been established) but: why has the model resulted
in this correlation?  To explore this question we need to go back and look at
the individual level data. The most relevant constraint variable for income was
class.  When we look at the relationship between social class and cake
consumption in the Dental Health Survey, we find that there is indeed a link:
individuals in the highest three classes (1.1, 1.2, 2) have an average cake
intake of 3.9 cakes per week whereas the three lowest classes have an average
intake of 3.7. This is a relatively modest difference but, when averaging over
large areas, it helps explain the result.

As a bonus exercise, explore the class dependence of cake consumption in the Dental Health Survey.

## Interpretation of results
