# Appendix: Getting up-to-speed with R {#apR}

As mentioned in Chapter 1, R is a general purpose programming
language focussed on data analysis and modelling.  This small tutorial aims to
teach the basics of R, from the perspective of spatial microsimulation research.
It should also be useful to people with existing R skills, to re-affirm their
knowledge base and see how it is applicable to spatial microsimulation.

R's design is built on the idea that everything that exists is an object and everything
that happens is a function. It is a *vectorised*, *object orientated* and
*functional* programming language (Wickham 2014). This means that R
understands vector algebra, all data accessible to R resides in a number of
named objects and functions must be used to modify objects. We will
look at each of these in some code below.

## R understands vector algebra {#vector-alg}

A vector is simply an ordered list of numbers (Beezer 2008).
Imagine two vectors, each consisting of 3 elements:

$$a = (1,2,3); b = (9,8,6) $$

To say that R understands vector algebra is to say that it knows how to
handle vectors in the same way a mathematician does: 

$$a + b = (a_1 + b_1, a_2 + b_2, c_3 + c_3  ) = (10,10,9) $$

This may not seem remarkable, but it is. Most programming
languages are not vectorised, so they would see $a + b$ differently.
In Python, for example, this is the answer we get:^[We can
get the right answer in Python, by typing the following:
`import numpy; a=numpy.array([1,2,3]); b=numpy.array([9,8,6]); a+b`.]

```{r, engine='python', eval=FALSE}
a = [1,2,3]
b = [9,8,6]
print(a + b)
```

`## [1, 2, 3, 9, 8, 6]`

In R, the operation *just works*, intuitively:

```{r}
a <- c(1, 2, 3)
b <- c(9, 8, 6)
a + b
```

This conciseness is clearly very useful in spatial microsimulation, as numeric
variables of the same length are common (e.g. the attributes of individuals in a
zone) and can be acted on with a minimum of effort.

## R is object orientated {#R-object}

In R, everything that exists is an object with a name and a class. This is
useful, because R's functions know automatically how to behave differently on
different objects depending on their class.

To illustrate the point, let's create two objects, each with a different class
and see how the function `summarise` behaves differently, depending on the type.
This behaviour is *polymorphism*  [@Matloff2011]:

```{r}
# Create a character and a vector object
char_obj <- c("red", "blue", "red", "green")
num_obj <- c(1, 4, 2, 532.1)

# Summary of each object
summary(char_obj)
summary(num_obj)

# Summary of a factor object
fac_obj <- factor(char_obj)
summary(fac_obj)
```

In the example above, the output from `summary` for the numeric object `num_obj`
was very different from that of the character vector `char_obj`. Note that
although the same information was contained in `fac_obj` (a factor), the output
from `summary` changes again.

Note that objects can be called almost anything in R with the exceptions of
names beginning with a number or containing operator symbols such as `-`, `^`
and brackets. It is good practice to think about what the purpose of an object
is before naming it: using clear and concise names can save you a huge amount of
time in the long run.


## Subsetting in R {#subsetting}

R has powerful, concise and (over time) intuitive methods for taking subsets of
data. Using the SimpleWorld example we loaded in *Data preparation*,
let's explore the `ind` object in more detail, to see
how we can select the parts of an object we are most interested in. As before,
we need to load the data:

```{r}
ind <- read.csv("data/SimpleWorld/ind.csv") 
```

Now, it is easy from within R to call a single individual (e.g. individual 3)
using the square bracket notation:

```{r}
ind[3,]
```

The above example takes a subset of `ind` all elements present on the 3rd row:
for a 2 dimensional table, anything to the left of the comma refers to rows and
anything to the right refers to columns. Note that `ind[2:3,]` and
`ind[c(3,5),]` also take subsets of the `ind` object: the square brackets can
take *vector* inputs as well as single numbers.

We can also subset by columns: the second dimension. Confusingly, this can be
done in four ways, because `ind` is an R `data.frame`^[This can be ascertained
by typing `class(ind)`. It is useful to know the class of different R objects,
so make good use of the `class()` function.] and a data frame can behave
simultaneously as a list, a matrix and a data frame (only the results of the
first are shown):

```{r}
ind$age # data.frame column name notation I
# ind[, 2] # matrix notation
# ind["age"] # column name notation II
# ind[[2]] # list notation
# ind[2] # numeric data frame notation
```

It is also possible to subset cells by both rows and columns simultaneously.
Let us select query the gender of the 4th individual, as an example
(pay attention to the relative location of the comma inside the square brackets):

```{r}
ind[4, 3] # The attribute of the 4th individual in column 3
```

A commonly used trick in R that helps with the analysis of individual level data
is to subset a data frame based on one or more of its variables. Let's subset
first all females in our dataset and then all females over 50:

```{r}
ind[ind$sex == "f", ]
ind[ind$sex == "f" & ind$age > 50, ]
```

In the above code, R uses relational operators of equality (`==`) and inequality
(`>`) which can be used in combination using the `&` symbol. This works because,
as well as integer numbers, one can also place *boolean* variables into square
brackets: `ind$sex == "f"` returns a binary vector consisting solely of `TRUE`
and `FALSE` values.^[Thus, yet another way to invoke the 2nd column of `ind` is
the following: `ind[c(F, T, F)]`! Here, `T` and `F` are shorthand for "TRUE" and
"FALSE" respectively.] 

## Further R resources {#further}

The above tutorial should provide a sufficient grounding in R for beginners to
understand the practical examples in the book.  However, R is a deep language
and there is much else to learn that will be of benefit to your modelling
skills. There are many excellent books and tutorials that teach the fundamentals
of R for a variety of applications.
The following resources, in ascending order of difficulty,
are highly recommended:

- *Introduction to visualising spatial data in R* (Lovelace and Cheshire 2014)
provides an introductory tutorial on handling spatial data in R, including the
administrative zone data which often form the building blocks of spatial microsimulation
models in R.
- *Introduction to scientific programming and simulation using R*
(Jones et al. 2014) is an
accessible and highly practical course that will form a solid foundation
for a range of modelling applications, including spatial microsimulation.
- *An Introduction to R* (Venables et al. 2014)
is the foundational introductory R manual, written by the
software's core developers and is available on-line for free.
It is terse and covers some advanced topics, but
provides a useful reference on the fundamentals of R as a language.
- *Advanced R* 
(Wickham 2014) (http://www.crcpress.com/product/isbn/9781466586963)
delves into the heart
of the R language. It contains many advanced topics, but the introductory
chapters are straightforward. Browsing some of the pages on
Advanced R's website (http://adv-r.had.co.nz/) and
trying to answer the questions that open each chapter
provides a taste of the book and an excellent
way of testing and improving one's understanding of the R language.

```{r, echo=F}
# There are alternatives to R and in the next section we will consider a few of these.
```
