---
title: "Additional tools and techniques"
layout: default
---

# Additional tools and techniques

```{r, echo=FALSE}

```

## R packages for spatial microsimulation

Most of the code presented so far --- 
with the exception of plotting commands based on **ggplot2**,
data manipulation tools from **dplyr** and some packages offering
niche functionality such as **ipfp** and **RNetLogo** --- has been implemented in R's base functions.
This is deliberate: base R offers robustness of
code and minimisation of installation dependencies.
There are often dozens of ways of doing one thing
in R, and a bewildering number of packages offering to help.
The way that will be understood by the greatest number
of people and work on the maximum number of computers
is often preferable, however, unless there is a
clear advantage to using additional packages. 
The behaviour of contributed packages
(excepting the `r-recommended` [packages](see http://cran.r-project.org/bin/linux/debian/README))
may change unexpectedly,
whereas core R functions are likely to remain stable over many decades to come.

With this caveat in mind, let's begin our tour of
some contributed R that are of use for
spatial microsimulation.

### **GREGWT**

### **sms**

### **multilevel**

## The Flexible Modelling Framework (FMF)

## Allocation of home-work locations

## Spatial interaction modelling

```{r}
## Advanced applications in agent-based modelling
```

