---
title: "Preparing input data"
layout: default
---

# Preparing input data {#DataPrep}

```{r, echo=FALSE}
# Applications of spatial microsimulation - to be completed!

## Updating cross-tabulated census data

## Economic forecasting

## Small area estimation

## Transport modelling

## Dynamic spatial microsimulation

## An input into agent based models
```


```{r, echo=FALSE}
# With the foundations built in the previous chapters now (hopefully) firmly in-place,
# we progress in this chapter to actually *run* a spatial microsimulation model
# This is designed to form the foundation of a spatial microsimulation course.
```


```{r, echo=FALSE}
# This next chapter is where people get their hands dirty for the first time -
# could be the beginning of part 2 if the book's divided into parts. 
```

This chapter focuses on the input datasets needed for spatial microsimulation.
Correctly loading, manipulating and assessing these datasets will be critical to
the performance of your models and the ease of modifying them to include new
inputs.  Fortunately R is an accomplished tool for data cleaning (Kabacoff,
2011) (http://www.manning.com/kabacoff/), as we shall see.  This chapter also
provides the basis for chapter we perform spatial microsimulation.

As with most spatial microsimulation models, this example consists of a
non-geographical individual-level dataset and a series of geographical zones.
The data used in this chapter (and the data for all other chapters) can be
downloaded from the book's GitHub repository (https://github.com/Robinlovelace/spatial-microsim-book). From this page, click on the
'Download ZIP' button to the right of the page and extract the folder into a
sensible place on your computer, such as the Desktop.  From there, you will want
to run R from the project's root directory: open the folder in a file browser
and double click on 'spatial-microsim-book.Rproj'.  This should cause RStudio to
be opened at this location, with all the input data files easily accessible to R
through *relative file paths*.

To ease reproducibility of the analysis when working with real data, it is
recommended that the process begins with a copy of the *raw* dataset on one's
hard disc.  Rather than modifying this file, modified ('cleaned') versions
should be saved as separate files. This ensures that after any mistakes, one can
always recover information that otherwise could have been lost and makes the
project fully reproducible. In this chapter, however, a relatively clean and
very tiny dataset from SimpleWorld is used.  We will see in Chapter
\ref{CakeMap} how to deal with larger and more messy data.  Here the focus is on
the principles.

```{r, echo=FALSE}
# It sounds trivial, but the *precise* origin of the input data
# should be described. Comments in code that loads the data (and resulting publications),
# allows you or others to recall the raw information. # going on a little -> rm
# Show directory structure plot from Gillespie here
```

The process of loading, checking and preparing the input datasets for spatial
microsimulation is generally a linear process, encapsulating the following
stages:

1. Load original data 
2. Remove excess information
3. Re-categorise individual-level data
4. Set variable and value names
5. 'Flatten' individual-level data 

'Stripping down' the datasets so that they only contain the bare essential
information will enable you to focus solely on the data that you are interested
in. This is not covered in this chapter because the input datasets are already
extremely bare and because the process should be obvious.

We start with the individual-level dataset for a reason: this dataset is often
more problematic to format than the constraint variables, so it is worth
becoming acquainted with it at the outset. Of course, it is possible that the
data you have are not suitable for spatial microsimulation because they lack
sufficient constraint variables with shared categories in both individual and
aggregate level tables. We assume that you have already checked this. The
checking process for the datasets used in this chapter is simple: both aggregate
and individual-level tables contain age and sex, so they can by combined. Let us
proceed to load some data saved on our hard disc into R's *environment*, where
it is available in memory.

## Loading input data {#Loading} 

Real-world individual-level data may be provided in a variety of formats
but ultimately needs to be loaded into R as a *data frame* object.

In this case the dataset is loaded from a `.csv` file:

```{r}
# Load the individual-level data
ind <- read.csv("data/SimpleWorld/ind.csv") 
class(ind) # verify the data type of the object
ind # print the individual-level data
```

```{r, echo=FALSE}
### Loading and checking aggregate-level data
```

Constraint data are usually made available one variable at a time,
so these are read in one file at a time:

```{r}
con_age <- read.csv("data/SimpleWorld/age.csv")
con_sex <- read.csv("data/SimpleWorld/sex.csv")
```

We have loaded the aggregate constraints. As with the individual level data, is
worth inspecting each object to ensure that they make sense before continuing.
Taking a look at `age_con`, we can see that this data set consists of 2
variables for 3 zones:

```{r}
con_age
```

This tells us that there 12, 10 and 11 individuals in zones 1, 2 and 3,
respectively, with different proportions of young and old people. Zone 2, for
example, is heavily dominated by older people: there are 8 people over 50 whilst
there are only 2 young people (under 49) in the zone.

Even at this stage there is a potential for errors to be introduced.  A classic
mistake with areal data is that the order in which zones are loaded changes from
one table to the next. The constraint data should therefore come with some kind of *zone
id*, an identifying code that will eventually allow the attribute data to be
combined with polygon shapes in GIS software.

```{r, echo=FALSE}
# Make the constraint data contain an 'id' column, possibly scrambled 
```

If we're sure that the row numbers match between the age and sex tables (we are
sure in this case), the next important test is to check that the total
populations are equal for both sets of variables.  Ideally both the *total*
study area populations and *row totals* should match. If the *row totals* match,
this is a very good sign that not only confirms that the zones are listed in the
same order, but also that each variable is sampling from the same *population
base*. These tests are conducted in the following lines of code:

```{r}
sum(con_age)
sum(con_sex) 

rowSums(con_age)
rowSums(con_sex)
rowSums(con_age) == rowSums(con_sex)
```

The results of the previous operations are encouraging. The total population is
the same for each constraint overall and for each area (row) for both
constraints.  If the total populations between constraint variables do not match
(e.g. because the population bases are different) this is problematic.
Appropriate steps to normalise the errant constraint variables are described in
the CakeMap Chapter (\ref{CakeMap}).

## Subsetting to remove excess information

In the above code, `data.frame` objects containing precisely the information
required for the next stage were loaded.  More often, superfluous information
will need to be removed from the data and subsets taken. It is worth removing
superfluous variables earl, to avoid over-complicating and slowing-down the
analysis.  For example, if `ind` had 100 variables of which only the 1st, 3rd and 4th were of
interest (in that they match the constraint variables), the following command could be used to update the object. Only the relevant variables corresponding to
columns (1, 3 and 4 in this case) are retained: `ind <- ind[, c(1, 3, 4)]`.
Alternatively, `ind$age <- NULL` removes the age variable.

Although `ind` is small and simple it will behave in the same way as a much
larger dataset, providing opportunities for testing subsetting syntax in R.  It
is common, for example, to take a subset of the working *population base*: those
aged 16 and 74 in full-time employment. Methods for doing this are provided in
Section \ref{subsetting}.

## Re-cateorising individual-level variables

Before transforming the individual-level dataset `ind` into a form that can be
compared with the aggregate-level constraints, we must ensure that each dataset
contains the same information. It can be more challenging to re-categorise
individual-level variables than to re-name or combine aggregate-level variables,
so the former should usually be set first.  An obvious difference between the
individual and aggregate versions of the `age` variable is that the former is of
type `integer` whereas the latter is composed of discrete bins: 0 to 49 and 50+.
We can categories the variable into these bins using `cut()`:^[The combination
of curved and square brackets in the output may seem strange but this is in fact
an International Standard - see
http://en.wikipedia.org/wiki/ISO_31-11 for more
information.]

```{r}
# Test binning the age variable
cut(ind$age, breaks = c(0, 49, 120))
```

If we wanted to change these category labels to something more readable,
we can do this by adding another argument to the `cut` function:

```{r}
# Convert age into a categorical variable with user-chosen labels
(ind$age <- cut(ind$age, breaks = c(0, 49, 120),
  labels = c("a0_49", "a50+")))
```

Users should be ware that `cut` results in a vector of class *factor*, which
can cause problems later down the line.

## Matching individual and aggregate level data names

Before combining the newly recategorised individual-level data with the
aggregate constraints, it is useful to for the category labels to match up.
This may seem trivial, but will save time in the long run. Here is the problem:

```{r}
levels(ind$age)
names(con_age)
```

Note that the names are subtly different. To solve this issue, we can
simply change the names of the constraint variable, assuming they
are in the correct order:

```{r}
names(con_age) <- levels(ind$age) # rename aggregate variables
```

With both the age and sex constraint variable names now matching the
category labels of the individual-level data, we can proceed to create a
single constraint object we label `cons`. We do this with `cbind()`:

```{r}
cons <- cbind(con_age, con_sex)
cons[1:2, ] # display the constraints for the first two zones
```

## 'Flattening' the individual level data

We have made steps towards combining the individual and aggregate datasets and
now only need to deal with 2 objects (`ind` and `cons`) which now share
category and variable names.
However, these datasets cannot possibly be compared because they are of different dimensions:

```{r}
dim(ind)
dim(cons)
```

The above code confirms this: we have one individual-level dataset comprising 5
individuals with 3 variables (2 of which are constraint variables and the other an ID) and one
aggregate-level constraint table called `cons`, representing 3 zones
with count data for 4 categories across 2 variables.

The dimensions of at least one of these objects must change
before they can be easily compared. To do this
we 'flatten' the individual-level dataset;
this means increasing its width
to match the constraint data.
This is a two-stage process. First,
`model.matrix()` is used to expand each variable into the number of columns as there are categories in each and assign. 
Knoblauch and Maloney (2012) provide a lengthier description of this
which is available online, for free.

Second, `colSums()` is used to take the sum of each column.^[As we shall see in Section \ref{ipfp},
only the former of these is needed if we use the
**ipfp** package for re-weighting the data, but both are presented to enable
a better understanding of how IPF works.]

```{r}
cat_age <- model.matrix(~ ind$age - 1)
cat_sex <- model.matrix(~ ind$sex - 1)[, c(2, 1)]

 # Combine age and sex category columns into single data frame
(ind_cat <- cbind(cat_age, cat_sex)) # brackets -> print result
```

Note that second call to `model.matrix` is suffixed with `[, c(2, 1)]`.
This is to swap the order of the columns: the column variables are produced
from `model.matrix` is alphabetic, whereas the order in which the variables
have been saved in the constraints object `cons` is `male` then `female`.
Such subtleties can be hard to notice yet completely change one's results
so be warned: the output from `model.matrix` will not always be compatible
with the constraint variables.

To check that the code worked properly,
let's count the number of individuals
represented in the new `ind_cat` variable, using `colSums`:

```{r}
colSums(ind_cat) # view the aggregated version of ind
ind_agg <- colSums(ind_cat) # save the result
```

The sum of both age and sex variables is 5 
(the total number of individuals): it worked! 
Looking at `ind_agg`, it is also clear that the object has the same 'width',
or number of columns,
`cons`. This means that the individual-level data can now be compared with
the aggregate-level data. We can check this by inspecting
each object (e.g. via `View(ind_agg)`). A more rigorous test is to see
if `ind_agg` can be combined with `ind_agg`, using `rbind`:

```{r}
rbind(cons[1,], ind_agg) # test compatibility of ind_agg and cons
```

If no error message is displayed on you computer, the answer is yes.
This shows us a direct comparison between the number of people in each
category of the constraint variables in zone and and in the individual level
dataset overall. Clearly, the fit is not very good, with only 5 individuals
in total existing in `ind_agg` (the total for each constraint) and 12
in zone 1. We can measure the size of this difference using measures of
*goodnes of fit*. A simple measure is total absolute error (TAE), calculated in this
case as `sum(abs(cons[1,] - ind_agg))`: the sum of the positive differences
between cell values in the individual and aggregate level data.

The purpose of the *reweighting* procedure in spatial microsimulation is
to minimise this difference (as measured in TAE above)
by adding high weights to the most representative individuals.
