---
title: "SimpleWorld"
layout: default
---

# SimpleWorld {#SimpleWorld}

To see the link between the methodology introduced later in the
book (see Chapter 5) and the various real-world applications,
this chapter describes the spatial microsimulation method on
an extremely simple example. This SimpleWorld example is analogous to the
'cartoon world' used by professor David MacKay (2009) to simplify the
complexities of sustainable energy. The same principle works here:
"a simple cartoon is the best way to capture the issues" (p. 203, Chapter 27).
By helping us understand likely differences in income and age distributions
in SimpleWorld, the example should also shed light on
of the kinds of situation where spatial microsimulation is useful.

We'll use an imaginary world called SimpleWorld, consisting of only 3 zones that
cover the entirety of SimpleWorld, which is considered to be a 2 dimensional
plane split into 3 zones (see figure 3.1). 

```{r simpleworld,  fig.cap="The SimpleWorld environment, consisting of individuals living in 3 zones.", echo=FALSE, message=FALSE}
# Code to create SimpleWorld
# Builds on this vignette: http://cran.r-project.org/web/packages/sp/vignettes/over.pdf
# library(sp)
# library(ggplot2)
# xpol <- c(-180, -60, -60, -180, -180)
# ypol <- c(-70, -70, 70, 70, -70)
# pol = SpatialPolygons(list(
#   Polygons(list(Polygon(cbind(xpol, ypol))), ID="x1"),
#   Polygons(list(Polygon(cbind(xpol + 120, ypol))), ID="x2"),
#   Polygons(list(Polygon(cbind(xpol + 240, ypol))), ID="x3")
#   ))
# # plot(pol)
# proj4string(pol) <- CRS("+init=epsg:4326")
# pol1 <- fortify(pol)
# 
# theme_space_map <- theme_bw() +
#   theme(
# #     rect = element_blank(),
#     axis.title = element_blank(),
#     axis.text = element_blank(),
#     axis.ticks = element_blank(),
#     panel.grid.major = element_line(size = 3)
#   )
# 
# ggplot(pol1) + geom_path(aes(long, lat, group, fill = group)) +
#     coord_map("ortho", orientation=c(41, -74, 52)) + 
#   theme_space_map
b <- matrix(c(-180, 180, 180, -180, -180, 90, 90, -90, -90, 90), ncol = 2)
library(sp)
l <- Line(coords = b)
l <- Lines(list(l), ID = "a")
l <- SpatialLines(list(l))
library(rgeos)
p <- gPolygonize(l)
plot(p)
l1 <- Line(matrix(c(-60, -60, 90, -90), ncol = 2))
l1 <- Lines(list(l1), "a")
l1 <- SpatialLines(list(l1))
lines(l1)
l2 <- Line(matrix(c(60, 60, 90, -90), ncol = 2))
l2 <- Lines(list(l2), "a")
l2 <- SpatialLines(list(l2))
lines(l2)
zt <- paste0("Zone ", 1:3)
text(seq(-120, 120, length.out = 3), 0, zt)
```

## SimpleWorld data {#SimpleWorldData}

This is a small world, 
containing 12, 10 and 11 individuals of its alien inhabitants
in zones 1 to 3, respectively: a planetary population of 33.
From the SimpleWorld Census, we know
how many young (under 49 space years old) and old (over 50)
residents live in each
zone, as well their genders: male and female.
This information is displayed in the tables below.

Table: Aggregate-level age counts for SimpleWorld.

|zone   | 0-49 yrs| 50 + yrs|
|:--|-----:|-----:|
|1  |     8|     4|
|2  |     2|     8|
|3  |     7|     4|

Table: Aggregate sex counts for SimpleWorld.

|Zone   |  m|  f|
|:--|--:|--:|
|1  |  6|  6|
|2  |  4|  6|
|3  |  3|  8|

```{r fig = "Mercator maps of the zones in SimpleWorld", echo=FALSE, message=FALSE}
con_age <- read.csv("data/SimpleWorld/age.csv")
con_sex <- read.csv("data/SimpleWorld/sex.csv")
cons <- cbind(con_age, con_sex)

# library(knitr)
# kable(con_age, row.names = T)
# kable(con_sex, row.names = T)

pol <- SpatialPolygonsDataFrame(pol, cons, match.ID = F)

# pol@data
pol$p_young <- pol$a0.49 / (pol$a.50. + pol$a0.49) * 100
pol$p_male <- pol$m / (pol$f + pol$m) * 100

pol$id <- c("x1", "x2", "x3")
library(dplyr)
pol1 <- inner_join(pol1, pol@data)
pol1$Name <- paste("Zone", 1:3, sep = " ")
pol1$xpos <- seq(-120, 120, length.out = 3)
pol1$ypos <- 0

# ggplot(pol1) + 
#   geom_polygon(aes(long, lat, group, fill = p_young)) +
#   geom_path(aes(long, lat, group, fill = p_young)) +
#   geom_text(aes(xpos, ypos, label = Name)) +
#   theme_bw() +
#   scale_fill_continuous(low = "black", high = "white", limits = c(0, 100),
#     name = "% Young") +
#   coord_map() 
# 
# ggplot(pol1) + 
#   geom_polygon(aes(long, lat, group, fill = p_male)) +
#   geom_path(aes(long, lat, group, fill = p_male)) +
#   geom_text(aes(xpos, ypos, label = Name)) +
#   theme_bw() +
#   scale_fill_continuous(low = "black", high = "white", limits = c(0, 100),
#     name = "% Male") +
#   coord_map() 
```

These data come from a census, so we can consider that we have here the whole 
population we want to analyse. However, we know only the marginal distributions
of the two variables, but not the contingency table. Indeed, we have per zone
the number of young and less young people, but we do not have the number
of young girls, and so on. Moreover, we need to add an other variable (the 
income). So, we are not satisfied with the database alone, so we desire to
use a spatial microsimulation. But to do this, we need some additional.

Next, imagine a more detailed dataset about only 5 of SimpleWorld's
inhabitants, recorded from a survey. This is in a different
form from the aggregate-level data presented in the above tables.
This *microdata* survey contains one row per individual, in contrast
to the *aggregate constraints*, which have one row per zone.
This individual level data includes exact age
(as opposed to the broad categories used in the zone data), as well as income.

\clearpage

```{r, echo=FALSE}
# ind <- read.csv("data/SimpleWorld/ind.csv")
# ind$income <- round(rnorm(n = nrow(ind), mean = 1000, sd = 100))
# ind$income <- ind$income + 30 * ind$age
# ind$income[ ind$age == "f"] <- ind$income + 1000
# write.csv(ind, "data/SimpleWorld/ind-full.csv", row.names = F)
ind <- read.csv("data/SimpleWorld/ind-full.csv")
# kable(ind)
```

Table: Individual-level survey data from SimpleWorld.

| id| age|sex | income|
|--:|---:|:---|------:|
|  1|  59|m   |   2868|
|  2|  54|m   |   2474|
|  3|  35|m   |   2231|
|  4|  73|f   |   3152|
|  5|  49|f   |   2473|

Note that although the microdataset contains additional information about the
inhabitants of SimpleWorld, it lacks geographical information about where each
inhabitant lives or even which zone they are from.  This is typical of
individual-level survey data.  Spatial microsimulation tackles this issue by
allocating individuals from a non-geographical dataset to geographical zones in
another.

## A weight matrix {#weight-matrix}

The procedures we will learn to use in this book do this by allocating *weights*
to each individual for each zone. The higher the weight for a particular
individual-zone combination, the more representative that individual is of that
zone. This information can be represented as a *weight matrix*, such as the one
shown below.

```{r, echo=FALSE, eval=FALSE}
cat_age <- model.matrix(~ ind$age - 1)
cat_sex <- model.matrix(~ ind$sex - 1)[, c(2, 1)]
ind_cat <- cbind(cat_age, cat_sex) # combine flat representations of the data
ind$age <- cut(ind$age, breaks = c(0, 49, 120), labels = c("a0_49", "a50+"))
weights <- matrix(data = NA, nrow = nrow(ind), ncol = nrow(cons))
cons <- apply(cons, 2, as.numeric)
library(ipfp) # load the ipfp library after: install.packages("ipfp")
apply(cons, MARGIN = 1, FUN =  function(x) ipfp(x, t(ind_cat), x0 = rep(1,nrow(ind))))
```

Table: A 'weight matrix' linking the microdata (rows) to the zones (columns)
of SimpleWorld.

| Individual| Zone 1  | Zone 2  | Zone 3  |
|---:|--------:|--------:|--------:|
|   1|    1.228|    1.725|    0.725|
|   2|    1.228|    1.725|    0.725|
|   3|    3.544|    0.550|    1.550|
|   4|    1.544|    4.550|    2.550|
|   5|    4.456|    1.450|    5.450|

The highest value (5.450) is located, to use R's notation, in cell
`weights[5,3]`, the 5th row and 3rd column in the matrix `weights`. This means
that individual number 5 is considered to be highly representative of Zone 3,
given the input data in SimpleWorld.  This makes sense because there are many
(7) young people and many (8) females in Zone 3, relative to the input
microdataset (which contains only 1 young female). The lowest value (0.550) is
found in cell `[3,2]`. Again this makes sense: individual 3 from the
microdataset is a young male yet there are only 2 young people and 4 males in
zone 2. A special feature of the weight matrix above is that each of the column
sums is equal to the total population in each zone.  We will discover how the
weight matrices are generated in Chapter 5.

## Spatial microdata {#spatial-microdata}

A more useful output from spatial microsimulation is what we refer to as
*spatial microdata*.  This is dataset that contains a single row per individual
(as with the input microdata) but also an additional variable indicating where
each individual lives. The challenge is to ensure that the spatial microdataset
is as representative as possible of the aggregate constraints, while only
sampling from a realistic baseline population. A feasible combination of
individuals sampled from the microdata that represent zone 2 is presented in the
table below. The complete spatial microdataset allocates whole individuals to
each zone, resulting in a more or less realistic insight into the inhabitants of
SimpleWorld. This spatial microdataset is also useful
for the purposes of modelling. We will create this
table at the end of Chapter 4.

| id| zone| age|sex | income|
|--:|----:|---:|:---|------:|
|  1|    2|  59|m   |   2868|
|  2|    2|  54|m   |   2474|
|  4|    2|  73|f   |   3152|
|  4|    2|  73|f   |   3152|
|  4|    2|  73|f   |   3152|
|  4|    2|  73|f   |   3152|
|  5|    2|  49|f   |   2473|
|  4|    2|  73|f   |   3152|
|  5|    2|  49|f   |   2473|
|  2|    2|  54|m   |   2474|

Table: Spatial microdata generated for SimpleWorld zone 2.

The table is a reasonable approximation of the inhabitants of zone 2: older
females dominate in both the aggregate (which contains 8 older people and 6
females) and the simulated spatial microdata (which contains 8 older people and
6 females). Note that in addition the constraint variables, we also have
an estimate the income distribution in SimpleWorld's second zone.

Towards the end of Chapter 4 we will learn how to generate
this table from first principles. The remainder of this section
considers how the outputs of spatial microsimulation, in the context of
SimpleWorld, can be useful before progressing to the practicalities.

```{r, echo=FALSE}
# Add section link here!
```

## SimpleWorld in context {#SimpleWorldContext}

Even though the datasets are tiny in SimpleWorld, we have already generated some
useful output. We can estimate, for example, the average income in each zone.
Furthermore, we could create an estimate of the *distribution* of income in each
area. Although these estimates are unlikely to be very accurate due to the
paucity of data, the methods could be very useful if performed on larger
datasets from the RealWorld (planet Earth).  Finally, the spatial microdata
presented in the above table can be used as an input into an agent-based model
(ABM). Assuming the inhabitants of SimpleWorld are more predictable than those
of RealWorld, the outputs from such a model could be very useful indeed, for
example for predicting future outcomes of current patterns of behaviour.

In addition to clarifying the advantages of spatial microsimulation, the above
example also flags some limitations of the methodology: spatial microsimulation
will only yield useful results if the input microdataset is representative of
the population as a whole, and for each region. If the relationship between age and
sex is markedly different in one zone compared with what we assume to be the
global averages of the input data, for example, our estimates could be way out.
Using such a small sample, one could rightly argue, how could the diversity of
33 inhabitants of SimpleWorld be represented by our simulated spatial microdata?
This question is equally applicable to larger simulations. These issues are
important and will be tackled in a subsequent Section (\ref{svalidation}) on
validation.
