---
title: "Spatial microsimulation for agent-based models"
layout: default
---

```{r, echo=FALSE}
# Resources:
# ABM www.geog.leeds.ac.uk/courses/level3/geog3150/

# Java: http://www.geog.leeds.ac.uk/courses/other/crime/
# Java about half way down ABM "details" page:
# http://www.geog.leeds.ac.uk/courses/other/crime/abm/general-modelling/index.html
```



# Spatial microdata in agent-based models {#ABM}

Spatial microsimulation can be seen as a precursor to, or early
form of, agent-based modelling (ABM). Agent-based models involve three things:

1. a number of discrete agents...
2. with different characteristics
3. interacting.

Note that based on the above definition we
already have two-thirds of the necessary ingredients
of an ABM: the synthetic spatial microdata created in previous chapters can,
with the right tools, skills and understanding, be used almsot directly as an
input into ABM. If your aim is to use spatial microdata as an input into
agent based models, you're more than half way there!

Before describing the transition from the
spatial microdata to ABM based on a simple example (you guessed it --- SimpleWorld),
it is worth mentioning some applications of ABMs built on spatial microdata
and some of the tools that will be useful for the job.

[NetLogo](https://ccl.northwestern.edu/netlogo/) is a
mature and widely used toolkit for agent-based models written in
Java.  The recently published **RNetLogo** package provides an interface between
R and NetLogo, allowing for model runs to be set-up and run directly from within R
(Theile et al. 2014).
Crucially,
this allows the outputs of agent based models to be loaded directly into
your R environment. Using R to run a separate programme may seem overly complicated
("why not simply run the model in NetLogo?" is a valid question) for
very simply models and for setting-up and testing the 
ABM it is recommended to use NetLogo directly, with its
intuitive graphical interface.

```{r,echo=FALSE}
# TODO: Batty reference below on abm design
```


For many ABM applications, however, the R interface is useful. This is because
AMB's are inherently unpredictable due to their 'bottom up' design.
Thus we generally want to study many model runs before drawing conclusions
about how the overall system operates, let alone real world implications.
Because much of the
time taken for agent based modelling is consumed on this sensitivity/parameter space
analysis, running NetLogo
from within R makes sense, as R excells in these areas whereas NetLogo
and other agent-based modelling programs do not
[@Thiele2014]. 

```{r, echo=FALSE}

# 
# Watch this space for more on this!
# 
# https://github.com/Robinlovelace/spatial-microsim-book
```
