---
title: "Spatial microsimulation for agent-based models"
layout: default
---

# Spatial microdata in agent-based models {#ABM}

Spatial microsimulation can be seen as a precursor to, or early
form of, agent-based models (ABMs). Agent-based modelling depends on 1) a number
of discrete agents, 2) with different characteristics, 3) interacting. With the
spatial microsimulation model created in the previous chapters we
already have two thirds of the necessary ingredients
of an ABMs: if your aim is to use spatial microdata as an input into
agent based models, you're more than half way there!

We do not have space in this book to describe the transition from the
spatial microdata we have generated into a full ABM. Suffice to mention some of
the tools that will be useful for the job.

[NetLogo](https://ccl.northwestern.edu/netlogo/) is a
mature and widely used toolkit for agent-based models written in
Java.  The recently published **RNetLogo** package provides an interface between
R and NetLogo, allowing for model runs to be set-up and run directly from within
R ([Thiele, 2014](http://www.jstatsoft.org/v58/i02/paper)). Crucially,
this allows the outputs of agent based models to be loaded directly into
your R environment. Using R to run a separate programme may seem overly complicated
("why not simply run the model in NetLogo?" is a valid question) for
very simply models and for setting-up and testing the 
ABM it is recommended to use NetLogo directly, with its
intuitive graphical interface.

```{r,echo=FALSE}
# TODO: Batty reference below on abm design
```


For many ABM applications, however, the R interface is useful. This is because
AMB's are inherently unpredictable due to their 'bottom up' design.
Thus we generally want to study many model runs before drawing conclusions
about how the overall system operates, let alone real world implications.
Because much of the
time taken for agent based modelling is consumed on this sensitivity/parameter space
analysis, running NetLogo
from within R makes sense

[@Thiele2014]

([Thiele et al., 2014](https://ideas.repec.org/a/jas/jasssj/2013-150-2.html)). 

```{r, echo=FALSE}

# 
# Watch this space for more on this!
# 
# https://github.com/Robinlovelace/spatial-microsim-book
```
