---
title: "Population synthesis with R"
layout: default
---

```{r, echo=FALSE}
# packages to allow 'knitting' the chapter to html
library(png)
library(grid)
```

# Population synthesis with R {#smsimr}

```{r, echo=FALSE}
# TODO:
# Ensure text is compatible with new use of chapter title as 'population synthesis'
# Discuss population synthesis without an input microdataset, with reference to Bathelemy and Toint (2013)
# Consider splitting this section in 2: reweighting vs CO approaches

# Completed:
# Change integerisation section so it's part of IPF (RL)
# Add IPF theory from Cambridge handout.pdf (mostly done)
# update figure with expansion
```

In this chapter we progress from loading and preparing the input data to
running a spatial microsimulation model. We will begin by describing the
IPF procedure as it is the simplest and most widely used method to allocate
individuals to zones and then progress to discuss alternative
*reweighting algorithms* that do the same job in a different way.

The SimpleWorld data, loaded in the previous chapter,
is used. Being small and simple, the example facilitates understanding
the process on a 'human scale' and allows experimentation
without the worry of overloading your computer.
However, the methods apply equally to larger and more complex projects.
Thus practicing the basic principles and methods of spatial microsimulation
in R is the focus of this chapter.
Time spent mastering these basics will make subsequent steps
much easier.

```{r, echo=FALSE}
# How representative each individual is of each zone is determined by their
# *weight* for that zone. If we have `nrow(cons)` zones and `nrow(ind)`
# individuals (3 and 5, respectively, in SimpleWorld) we will create
# 15 weights. Real world datasets (e.g. that presented in chapter xxx)
# could contain 10,000 individuals
# to be allocated to 500 zones, resulting in an unwieldy 5 million element
# weight matrix but we'll stick with the SimpleWorld dataset here for simplicity.
load("cache-data-prep.RData")
```

How representative each individual is of each zone is represented by their
*weight* for that zone. Each weight links and individual to a zone.
The number
of weights is therefore equal to number of zones multiplied
by the number of individuals
in the microdata, that is the number of rows in individual-level and constraint tables
respectively.
In terms of the SimpleWorld data loaded in the previous chapter we have, in
R syntax, `nrow(cons)` zones and `nrow(ind)` individuals.
(Typing those commands with the data loaded should confirm that there are
3 zones and 5 individuals in the input data for the SimpleWorld example).
This means that `nrow(cons) * nrow(ind)` weights will be estimated
(that is $3 * 5 = 15$ in SimpleWorld). The weights must begin with an initial value. We will create a matrix of 
ones so every
individual is seen as equally representative of every
zone:^[Any
initial weight values could be used here: initial weights do not affect the final weights after many iterations. The value of one is
used as a sensible
convention.]

```{r}
# Create the weight matrix of ones
weights <- matrix(data = 1, nrow = nrow(ind), ncol = nrow(cons))
dim(weights) # dimension of weight matrix: 5 rows by 3 columns
```

The weigth matrix links individual-level data to aggregate-level data.
A weight matrix value of 0 in cell `[i,j]`, for example, suggests that
nobody with the characteristics of individual `i` is present in
zone `j`.
During the IPF procedure these weights are iteratively updated until
they *converge* towards a single result: the final weights which create
a representative population for each zone.^[Fienberg (1970) provides
a geometrical proof that IPF converges to a single result, in the absence
of *empty cells*.]

## Weighting algorithms {#weighting}

A wide range of methods can be used to allocate individuals to zones in spatial microsimulation. As with the majority of procedures for statistical 
analysis, there are *deterministic* and *stochastic* methods.
The results of *deterministic* methods, such as IPF, never vary: no random
or probabilistic numbers are used so the resulting weights will be the same
every time. *Stochastic* methods such as *simulated annealing*, on the other hand, use random numbers.

In the literature, the divide between stochastic and deterministic approaches
is usually mapped onto a wider distinction between *reweighting*
and *combinatorial optimisation* methods. Reweighting methods generally calculate
non-integer weights for every individual-zone combination.
Combinatorial optimisation
methods generally work by randomly allocating individuals from the microdata
survey into each zone, one-by-one, and re-calculating the goodness-of-fit after each change.
If the fit between observed and simulated results improves after an individual
has been 'imported' into the zone in question, the individual will stay.
If the fit deteriorates, the individual will be removed from the zone and
the impact of switching a different individual into the zone is tested.

This distinction between reweighting of fractional weights
and combinatorial optimisation algorithms is important: combinatorial
optimisation methods result in whole individuals being allocated to each zones
whereas reweighting strategies result in fractions of individuals being allocated
to each zone. The latter method means that individual $i$ could have a weight of
0.223 (or any other positive real number) for zone $j$. Of course, this sounds
irrational: a quarter of a person cannot possibly exist: either she is in the zone
or she is not!

However, the distinction between combinatorial optimisation and reweighting
approaches to creating spatial microdata is not as clear cut as it may at first
seem. As illustrated figure 5.1, fractional weights generated by reweighting
algorithms such as IPF can be converted into integer weights (via *integerisation*).
Through the process of *expansion*, the integer weight matrix
produced by integerisation can be converted into the final spatial microdata
output stored in 'long' format represented in the right-hand box of figure 5.1.
Thus the combined processes of integerisation
and expansion allow weight matrices to be translated into the same output
format that combinatorial optimisation algorithms produce directly.
In other words fractional weighting is interchangeable with combinatorial
optimisation approaches to population synthesis.

The reverse process is also possible:
synthetic spatial microdata generated by combinatorial
optimisation algorithms can be converted back in to a more compact weight
matrix in a step that we call *compression*. This integer weight has the same
dimensions as the integer matrix generated through integerisation described above.

```{r, echo=FALSE}
# is the function tidyr:@expand the same as expand here? will test.
```

Integerisation, expansion and compression procedures allow fractional
weighting and combinatorial optimisation approaches to population synthesis
to be seen as essentially the same thing.
This equivalence between different methods of population synthesis is
the reason we have labelled this section
*weighting algorithms*: combinatorial optimisation approaches to population synthesis can
be seen as a special case of fractional weighting and vice versa.
Thus all deterministic and stochastic
(or weighting and combinatorial optimisation) approaches to the generation of
spatial microdata can be seen as different methods, algorithms, for allocating
weights to individuals. Individuals representative of a zone will be given a
high weight (which is equivalent to being replicated many times in
combinatorial optimisation). Individuals who are rare in a zone will
be given a low weight (or not appear at all, equivalent to a weight of zero).
Later in this chapter we demonstrate functions to translate between the
'weight matrix' and 'long spatial microdata' formats generated by each approach.

```{r, fig.cap="Schematic of different approaches for the creation of spatial microdata encapsulating stochastic combinatorial optimisation and deterministic reweighting algorithms such as IPF. Note that integerisation and 'compression' steps make the results of the two approaches interchangeable, hence our use of the term 'reweighting algorithm' to cover all methods for generating spatial microdata.", echo=FALSE}
img <- readPNG("figures/co-vs-ipf-schema.png")
grid.raster(img)
```

The concept of weights is critical to understanding how
population synthesis generates spatial microdata. 
To illustrate the point imagine a parallel SimpleWorld, in which we have
no information about the characteristics its inhabitants, only the
total population of each zone. In this case we could only assume that
the distribution of characteristics found in the sample is representative of the distribution of the whole population. Under this scenario, individuals
would be chosen at random from the sample and allocated to zones at random
and the distribution of characteristics of individuals in each zone would be
asymptotically the same as (tending towards) the microdata.

In R, this case can
be achieved using the `sample()` command. We could use this method to
randomly allocate the 5 individuals of the microdata to zone 1 (which has
a population of 12) with the following code:

```{r}
set.seed(1) # set the seed for reproducibility
sel <- sample(x = 5, size = 12, replace = T) # create selection
ind_z1 <- ind_orig[sel, ]
head(ind_z1, 3)
```

Note the use of `set.seed()` in the above code to ensure the results are reproducible. It is important to emphasise that without 'setting the seed' (determining the starting point of the random number sequence) the results will change each time the code is
run. This is because `sample()` (along with other stochastic functions such as `runif()` and `rnorm()`) is probabilistic and its output therefore depends on a random number
generator
(RNG).^[Without
setting the seed, the results will change each time. In addition, changing
the value inside the brackets of `set.seed()`
will result in a different combination of individuals being selected for each new number --- test this
out in your code. This happens because the method relies on *pseudo random numbers* to select values probabilistically and `set.seed()` specifies where the
random number sequence should begin, ensuring reproducibility. We must really trust the random function used. Note that it is impossible for a computer to
choose *entirely* random numbers, so algorithms to generate *pseudo-random numbers* have been developed. See the documentation provided by `?set.seed`
for more information.
] 

The *reweighting* methods consists in adding a weight to each individual for the zone. This method is good if we have 
a representative sample of the zone and the minority of the population is included in it. In contrary, if we have in
the individual level only the majority of the population and for example, we have not an old man still working, this
kind of individual will not appear in the final data. A proposal to avoid this is to use a *genetic algorithm* or something
similar, that will allow to mix some individual to create a new one (mutation). We do not know for the moment
if this solution has already been developed, it is just a proposition and to inform the reader that new process can
still appear in the future.

## Iterative Proportional Fitting

### IPF in theory {#IpfinTh}

The most widely used and mature *deterministic* method to allocate individuals
to zones is iterative proportional fitting (IPF).
IPF is mature, fast and has a long history: it was demonstrated by
Deming and Stephan (1940)
for estimating internal cells based on known marginals.
IPF involves calculating a series of
non-integer weights that represent how representative each individual is of
each zone. This is *reweighting*.

Regardless of implementation method,
IPF can be used to allocate individuals to zones through the calculation
of *maximum likelihood* values for each zone-individual combination represented
in the weight matrix. This is considered as the most probable configuration of 
individuals in these zones. IPF is a method of *entropy maximisation*
(Cleave et al. 1995). 
The *entropy* is the number of configurations of the underlying spatial microdata
that could result in the same marginal counts.
For in-depth treatment of the mathematics and theory underlying IPF,
interested readers are directed towards Fienberg (1979), Cleave et al. (1995) and an excellent recent review
of methods for maximum likelihood estimation (Fienberg and Rinaldo 2007). For the
purposes of this book, we provide a basic overview of the method for the purposes
of spatial microsimulation.

```{r, echo=FALSE}
# TODO (RL): add Lovelace et al (2015) example when published above
```

In spatial microsimulation, IPF is used to allocate individuals to zones.
The subsequent section implements IPF to create *spatial microdata* for
SimpleWorld, using the data loaded in the previous chapter as a basis. 
Overviews of the use of spatial microsimulation method for spatial microsimulation
are provided recent papers by Lovelace and Ballas (2012) and, for in the
context of transport modelling, by Pritchard and Miller (2012).

```{r, echo=FALSE}
# TODO (MD): add reference above
#(Dumont,2014) Jojo thesis and article given by Eric 
```

Such as with the example of SimpleWorld, in each application have a matrix 
`ind` containing the categorical value of each individual. `ind` is a two 
dimensional array (a matrix) in which each row represents an individual and 
each column a variable. The value of the cell `ind(i,j)` is therefore the 
category of the individual `i` for the variable `j`. A second array containing 
the constraining count data `cons` can, for the purpose of explaining the 
theory, be expressed in 3 dimensions, which we will label `cons_t`: 
`cons_t(i,j,k)` is the number of individuals corresponding to the marginal 
for the zone 'i', in the variable 'j' for the category 'k'. For example, 
'i' could be a municipality, 'j' the gender and 'k' the female. 
Element '(i,j,k)' is the total number of woman in this municipality according 
to the constraint data.

The IPF algorithm will proceed zone per zone. For each zone, each individual 
will have a weight of representativity of the zone. The weights matrix 
will then have the dimension 'number of individual x number of zone'. 
'w(i,j,t)' corresponds to the weight of the individual 'i' in the 
zone 'j' (during the step 't'). For the zone 'z', we will adapt the weight matrix to
each constraint 'c'.This matrix is initialized with a full matrix of 1 
and then, for each step 't', the formula can be expressed as:

$$ w(i,z,t+1) = w(i,z,t) * \displaystyle\frac{cons \_ t(z,c,ind(i,c))}{ \displaystyle\sum_{j=1}^{Nbindiv} w(j,z,t) * I(ind(j,c)=ind(i,c))} $$


where the 'I(x)' function is the indicator function which value is 1 
if x is true and 0 otherwise. We can see that 'ind(i,c)' is the category 
of the individual 'i' for the variable 'c'. The denominator corresponds 
to the sum of the actual weights of all individuals having the same category
in this variable as 'i'. We simply redistribue the weights so that the 
data follows the constraint concerning this variable.



### IPF in R {#IpfinR}

In the subsequent examples, we use IPF to allocate individuals to zones
in SimpleWorld, using the data loaded in
the previous chapter as a basis. IPF is mature,
fast and has a long history.
Interested readers are directed towards recent papers (e.g. 
Lovelace and Ballas, 2012;
Pritchard and Miller,2012)
for more detail on the method.

The IPF algorithm can be written in
R from scratch, as illustrated in Lovelace (2014), and as taught in the
smsim-course online
[tutorial](https://github.com/Robinlovelace/spatial-microsim-book).
We will refer to this implementation of IPF as 'IPFinR'.
The code described in this tutorial 'hard-codes' the IPF algorithm
in R and must be adapted to each new application (unlike the generalized 'ipfp' approach,
which works unmodified on any reweighting problem).
IPFinR works by saving the weight
matrix after every constraint for each iteration. We here develop first IPFinR to 
give you an idea of the algorithm. Indeed, 'ipfp' package is more general, but
is like a "black box", so we can use it without being sure of what it performs.

The aim is to obtain the final weight matrix that will represent how well each individual fit each zone in terms of their characteristics. Each row of this matrix is an individual and each column 
is a zone. The algorithm will operate zone per zone. Thus the weight matrix will be filled in
column per column. To help the understanding of this section, we can rename the totals with 
a more intuitive name.

```{r}
# Create intuitiv names for the totals
nzones <- nrow(cons) # number of zones
NbIndiv <- nrow(ind) # number of individuals
NbCatAge <-ncol(con_age) # number of categories of "age"
NbCatSex <-ncol(con_sex) # number of categories of "sex"
```

The earlier step before to perform to an algorithm is to initialize each needed object.
Here, we will have the *weight* matrix and the marginal distribution of individuals in each
zone. Thus first we create an object (`ind_agg0`), in which rows are zones and
columns are the different categories of the variables. Then, we duplicate the
weight matrix to keep in memory each step.

```{r}
# Create initial matrix of categorical counts from ind 
ind_agg0 <- t(apply(cons, 1, function(x) x^0 * ind_agg))
weights1 <- weights2 <- weights # create addition weight objects
```

IPFinR begins with a
couple of nested for loops, one to iterate through each zone
(hence `1:nzones`, which means "from 1 to the number of zones in the
constraint data") and one to iterate through each
category within the constraints (0--49 and 50+ for the first constraint).
Note that this code relies on the `cons` and `ind` objects loaded in the
previous chapter.

```{r, echo=FALSE}
# The long way of doing it 
# # Create initial matrix of categorical counts from ind
# ind_agg0 <- matrix(rep(ind_agg, nrow(cons)), nrow = nrow(cons), byrow = T)
# weights1 <- weights2 <- weights # create addition weight objects for IPFinR
# # Assign values to the previously created weight matrix
# for(j in 1:nrow(cons)){
# for(i in 1:ncol(con_age)){
# weights1[ind_cat[ , i ] == 1, j] <- con_age[j , i] / ind_agg0[j , i]
# }
# }
```


```{r}
# Assign values to the previously created weight matrix 
# to adapt to age constraint
for(j in 1:nzones){
  for(i in 1:NbCatAge){
weights1[ind_cat[, i] == 1, j] <- con_age[j, i] / ind_agg0[j, i]
    }
print(weights1)
  }
```

The above code updates the weight matrix by dividing each cell in
the census constraint (`con_age`) by the equivalent cell
aggregated version of the individual level data.
The weight matrix is critical to the spatial microsimulation procedure because
it describes how representative each individual is of each zone. To see the
weights that have been allocated to individuals to populate zone 2, for example
you would query the second column of the weights: `weights1[, 2]`. Conversely,
to see the weight allocated for individual 3 for each for the 3 zones, you
need to look at the 3rd column of the weight matrix: `weights1[3, ]`. 

Note that we asked R to write the resulting matrix after the completion of each zone.
As said before, the algorithm proceeds zone per zone and each column of the matrix
corresponds to a zone. This explains why the matrix is filled in per column. We can now 
verify that the weights correspond to the application of the theory seen before.
For the first zone, the age constraint was to have 8 people under 50 years old and 
4 over this age. The first individual is a man of 59 years old, so over 50.
To determine the weight of this person inside the zone 1, we multiply the 
actual weight, 1, by a ratio with a numerator corresponding to the
number of person in this category of age for the constraint, here 4,
and a denominator equal to the sum of the weights of the individual 
having the same age category. Here, there are 3 individuals of more 
than 50 years old and all weights are 1 for the moment. The new weight
is $$ 1 * \frac{4}{1+1+1}=1.33333$$.
We can verify the other weights with a similar reasoning. Now that
we explained the whole process under IPF, we can understand the origin
of the name 'Iterative Proportional Fitting'.

Thanks to the weight matrix, we can see that individual 3 
(who's attributes can be viewed by entering `ind[3, ]`)
is young and has a comparatively low weight of 1 for zone two. Intuitively
this makes sense because zone 3 has only 2 young adult inhabitants (see
the result of `cons[2,]`) but 8 older inhabitants. The reweighting stage is
making sense. Note also that the weights generated are fractional;
see [](#sintegerisation) for methods of converting fractional weights into
integer weights to generate a synthetic small-area population ready for
agent-based modelling applications. 

The next step in IPF, however, is to
re-aggregate the results from individual-level data after they have been
reweighted. For the first zone, the weights of each individual are in the
first column of the weight matrix. Moreover, the characteristics of each
individual are inside the matrix `ind_cat`. When multiplying `ind_cat` by 
the first column of `weights1` we obtain a vector, the values of which
correspond to the number of people in each category for zone 1.
To aggregate all individuals this for the
first zone, we just sum the values in each category. The following `for` loop
re-aggregates the individual-level data, with the new weights for each zone:

```{r}
# Create additional ind_agg objects
ind_agg2 <- ind_agg1 <- ind_agg0 * NA

# Assign values to the aggregated data after con 1
for(i in 1:nzones){
  ind_agg1[i, ] <- colSums(ind_cat * weights1[, i])
}
```

Congratulations. Assuming you are running the code on you own computer, you
have just reweighted your first individual-level dataset to a geographical
constraint (the age) and have aggregated the results.
At this early stage it is important to do some preliminary checks
to ensure that the code is working correctly. First, are the resulting
populations for each zone correct? We check this for the first constraint variable
(age) using the following code (test: check the populations of the unweighted
and weighted data for the second constraint --- sex):

```{r}
rowSums(ind_agg1[, 1:2]) # the simulated populations in each zone
rowSums(cons[, 1:2]) # the observed populations in each zone
```

The results of these tests show that the new populations are correct, verifying
the technique. But what about the fit between the observed and simulated
results after constraining by age? We will cover goodness of fit in more
detail in subsequent sections. For now, suffice to know that the simplest
way to test the fit is by using the `cor` function on a 1d representation of the
aggregate-level data:

```{r}
vec <- function(x) as.numeric(as.matrix(x))
cor(vec(ind_agg0), vec(cons))
cor(vec(ind_agg1), vec(cons))
```

The point here is to calculate the correlation between the aggregate actual data
and the constraints. This value is between -1 and 1 and in our case, 
the best fit will be 1, meaning that there is a perfect correlation between our data
and the constraints. Note that as well as creating the correct total population for each zone, the
new weights also lead to much better fit. To see how this has worked, let's
look at the weights generated for zone 1:

```{r}
weights1[, 1]
```

The results mean that individuals 3 and 5 have been allocated a weight of 4
whilst the rest have been given a weight of $4/3$. Note that the total of these
weights is 12, the population of zone 1. Note also that individuals 3 and 5 are
in the younger age group (verify this with `ind$age[c(3,5)]`) which are more
commonly observed in zone 1 than the older age group:

```{r}
cons[1, ]
```

Note there are 8 individuals under 50 years old in zone 1, but only 2 individuals
with this age in the individual-level survey dataset. This explains why the
weights allocated to these individuals is 4: 8 divided by 2 = 4.

So far we have only constrained by age. This results in aggregate-level results
that fit the age constraint but not the sex constraint (figure 5.2). The reason
for this should be obvious: weights are selected such that
the aggregated individual-level data fits the age constraint perfectly, but
no account is taken of the sex constraint. This is why IPF must constrain for
multiple constraint variables.

To constrain by sex, we simply repeat the
nested `for` loop demonstrated above for the sex constraint. This is implemented
in the code block below.

```{r}
for(j in 1:nzones){
  for(i in 1:NbCatSex + NbCatAge){
weights2[ind_cat[, i] == 1, j] <- cons[j , i] / ind_agg1[j, i]
    }
  }
```

Again, the aggregate values need to be calculated in a `for` loop over every
zone. After the first constraint fitting, the weights for zone 1 was:
$$(\frac{4}{3},\frac{4}{3},4,\frac{4}{3},4) $$
We can explain theoretically explain the weights for zone 1 after the
second fitting. For the first individual, its actual weight is $\frac{4}{3}$
and he is a male. In the zone one, the constraint is to have 6 men. The 
three first individuals are men, so the new weight for this person in this
zone is $$weights2[1,1]=\frac{4}{3}*\frac{6}{\frac{4}{3}+\frac{4}{3}+4}=\frac{6}{5}=1.2 $$

With an analogous reasoning, we can find all weights in *weights2*:

```{r}
weights2
```

Note that the final value is calculated by multiplying by
`weights1` *and* `weights2`: the final weight is calculated as the product
of all the weight matrices calculated from all the constraints.

```{r}
for(i in 1:nzones){
ind_agg2[i, ] <- colSums(ind_cat * weights1[, i] * weights2[, i])
}
```

Note that even after constraining by age and sex, there
is still not a perfect fit between observed and simulated cell values
(figure 5.2). The simulated cell counts for the age categories are far from
the observed, whilst the cell counts for the age categories fit perfectly.
On the other hand, after constraining by age *and* sex, the fit is still not
perfect. This time the age categories do not fit perfectly, whilst the sex
categories fit perfectly. Inspect `ind_agg1` and `ind_agg2` and try to explain
why this is. Note that the *overall fit*, combining age and sex categories,
has improved greatly from iteration 1.1 to 1.2 (from $r = 0.63$ to $r = 0.99$).
In iteration 2.1 (in which we constrain by age again) the fit improves again.

So, each time we reweight to a specific constraint, the fit of this constraint 
is perfect, because, as seen in theory, it is a proportional reallocation. 
Then, we repeat for another constraint and the first one can diverge from his
perfect fit. However, when  operating this process several time, we always
refit to the next constraint and we can converge to a unique weight matrix.

These results show that IPF requires multiple *iterations* before converging on
a single weight matrix. It is relatively simple to iterate the procedure
illustrated in this section multiple times, as described in the smsim-course
tutorial. However, for the purposes of this book, we will move on now to consider
implementations of IPF that automate the fitting procedure and iteration process.
The aim is to make spatial microsimulation as easy and accessible as possible.


```{r valit_plot1, echo=FALSE, fig.cap="Fit between observed and simulated values for age and sex categories (column facets) after constraining a first time by age and sex constraints (iterations 1.1 and 1.2, plot rows). The dotted line in each plot represents perfect fit between the simulated and observed cell values. The overall fit in each case would be found by combining the left and right-hand plots. Each symbol correspond to a category and each category has a couple (observed, simulated) for each zone."}
# Make data long it with tidyr
library(tidyr)
library(dplyr) # a package required by tidyr functions
x <- gather(cons, Category, Observed, a0_49:f)
y <- gather(as.data.frame(ind_agg1), cat, Simulated, a0_49:f)
z <- gather(as.data.frame(ind_agg2), cat, Simulated, a0_49:f)
# y$cat <- x$cat # to make categories identical (not needed)
df1 <- cbind(x,y)
df1$Constraint <- "1.1: Age"
df2 <- cbind(x, z)
df2$Constraint <- "1.2: Sex"
df <- rbind(df1, df2)
df$Variable <- "Age"
df$Variable[grep("[mf]", df$Category)] <- "Sex"
library(ggplot2)
qplot(Observed, Simulated, data = df, shape = Category, alpha = 0.5) +
  facet_grid(Constraint ~ Variable) +
  geom_abline(slope = 1, linetype = 3) +
  scale_alpha(guide = F) +
  scale_shape_manual(values = c(1,11,2,12)) +
  theme_bw()
# ggsave("figures/fit-obs-sim-simple-5.png")
# Attempt 2: do it with reshape2
# library(reshape2)
# melt(cons)
```

The advantage of hard-coding the IPF process, as illustrated above, is that
it is helps understand how IPF works and aides diagnosing issues with
the reweighting process as the weight matrix is re-saved after every
constraint and iteration. However, there are more computationally efficient
approaches to IPF. To save computer and researcher time, we 
use in the next sections 
R packages which implement IPF without the user needing to *hard code*
each iteration in R:
**ipfp** and  **mipfp**. We will use each of these methods to generate
fractional weight matrices allocating each individual
to zones. After following this reweighting process with **ipfp**, we will progress to
integerisation: the process of converting the fractional weights into integers. 
This process creates a final contingency table, which is used to generate the
final population. This last step is called the expansion.


```{r, echo=FALSE}
# Possibly more on IPF here. For now, press on
```

### Reweighting with **ipfp** {#ipfp}

IPF runs much faster and with less code using the
**ipfp** package than in pure R. The `ipfp` function runs the IPF algorithm
in the C language, taking aggregate constraints, individual level
data and an initial weight vector (`x0`) as inputs:

```{r}
library(ipfp) # load ipfp library after install.packages("ipfp")
cons <- apply(cons, 2, as.numeric) # to 1d numeric data type
ipfp(cons[1,], t(ind_cat), x0 = rep(1, NbIndiv)) # run IPF
```

It is impressive that the entire IPF process, which takes dozens of lines of
code in pure R can been condensed into two lines: one to
convert the input constraint dataset to `numeric`^[The integer data type fails
because C requires `numeric` data to be converted into its *floating point*
data class.]
and one to perform the IPF operation itself. The whole procedure is hiding
behind the function that is created in C and well optimised. So, it is like a 
magic box where you put your data and that returns the results. This is a good
way to execute the algorithm, but you have to pay attention to well understand 
the format of the input argument to use the function correctly. To be sure, 
type on R *help('ipfp')*.

Note that although
we did not specify how many iterations to run, the above command
ran the default of `maxit = 1000` iterations, despite convergence happening after
10 iterations. This can be seen by specifying the `maxit` and `verbose` arguments
(the latter of which can be referred to lazily as `v`) in `ipfp`, as illustrated below (only the first line of R output is shown):

```{r, eval=FALSE}
ipfp(cons[1,], t(ind_cat), rep(1, NbIndiv), maxit = 20, v = T)
```

```
## iteration 0:   0.141421
## iteration 1:   0.00367328
## iteration 2:   9.54727e-05
## ...
## iteration 9:   4.96507e-16
## iteration 10:  4.96507e-16
```

The `maxit` argument specifies to the function the maximum number of iterations and `verbose` set to True asks
the function to write for each iteration the convergence rate. This corresponds to the distance between
the previous and actual weight matrices. When the two matrix are equal, that means that the algorithm has
converged and the distance will approach 0. Note that when calculating, the computer has to make some approximations
(for example, when calculating the result of $\frac{4}{3}$, the computer cannot save the infinite number of decimals and truncates the number).
Due to this, we never reach a perfect 0, but when we have a very little number, we can consider it enough close to 0.
Usually, we use the precision of the computer that is of order $10^{-16}$ (on R, you can display the precision of the
machine by typing `.Machine$double.eps`). Note that that instead of using `maxit` as stopping condition, 
we can use the `tol` argument, conditional on the distance between two consecutive weight matrices.

Notice also that a *transposed* (via the `t()` function) version of the individual-level
data (`ind_cat`) is used in `ipfp`
to represent the individual-level data, instead of the
`ind_agg` object used in the pure R version. To prevent having to transpose
`ind_cat` every time `ipfp` is called, save the transposed version:

```{r}
ind_catt <- t(ind_cat) # save transposed version of ind_cat
```

Another object that can be saved prior to running `ipfp` on all zones
(the rows of `cons`) is `rep(1, nrow(ind))`, simply a series of ones - one for each individual.
We will call this object `x0` as its argument name representing
the starting point of the weight estimates in `ipfp`:

```{r}
x0 <- rep(1, NbIndiv) # save the initial vector
```

To extend this process to all three zones we can wrap the line beginning
`ipfp(...)` inside a `for` loop, saving the results each time into a
weight variable we created earlier:

```{r}
weights_maxit_2 <- weights # create a copy of the weights object
for(i in 1:ncol(weights)){
  weights_maxit_2[,i] <- ipfp(cons[i,], ind_catt, x0, maxit = 2)
}
```

The above code uses `i` to iterate through the constraints, one row (zone) at
a time, saving the output vector of weights for the individuals into columns
of the weight matrix. To make this process even more concise (albeit
less clear to R beginners), we can use R's internal
`for` loop, `apply`:

```{r}
weights <- apply(cons, MARGIN = 1, FUN = 
    function(x) ipfp(x, ind_catt, x0, maxit = 20))
```

In the above code R iterates through each row
(hence the second argument `MARGIN` being `1`, `MARGIN = 2`
would signify column-wise iteration).
Thus `ipfp` is applied to each zone in turn, as with the `for` loop implementation. 
The speed savings of writing the function 
with different configurations are benchmarked in
'parallel-ipfp.R' in the 'R' folder of the book project directory.
This shows that reducing the maximum iterations of `ipfp` from
the default 1000 to 20 has the greatest performance benefit.^[These
tests also show that any speed gains from using `apply` instead of `for` are negligible, so
whether to use `for` or `apply` can be decided by personal preference.]
To make the code run faster on large datasets, a parallel version of
`apply` called `parApply` can be used. This is also
tested in 'parallel-ipfp.R'.

For your personal applications, take care of using `maxit` alone. Indeed, it is
impossible to predict the number of iterations necessary for all applications.
So, if your argument is to big you will needlessly lose time, but if it is
to small, you will have a result that has not converge! However, using
`tol` alone is also hazardous. Indeed, if you have iteratively the same
matrices, but the approximations on the distance is a number bigger than 
your argument, the algorithm will continue. That's why we use both together and
it stops either if the convergence is achieve nor the maximum number of iterations
is reached. By defaults, `maxit` is 1000 and `tol` is `.Machine$double.eps`. 

```{r, echo=FALSE, eval=FALSE}
# Also discuss what happens when you get a huge dataset, from Stephen's dataset
```

It is important to check that the weights obtained from IPF make sense.
To do this, we multiply the weights of each individual by rows of
the `ind_cat` matrix, for each zone. Again, this can be done using
a for loop, but the apply method is more concise:

```{r}
ind_agg <- t(apply(weights, 2, function(x) colSums(x * ind_cat)))
colnames(ind_agg) <- colnames(cons) # make the column names equal
```

As a preliminary test of fit,
it makes sense to check a sample of the aggregated weighted data
(`ind_agg`) against the same sample of the constraints.
Let's look at the results (one would use a subset of the results, 
e.g. `ind_agg[1:3, 1:5] for the first five values of the first 3
zones for larger constraint tables found in the real world):

```{r}
ind_agg
cons
```

This is a good result: the constraints perfectly match the results
generated using ipf, at least for the sample. To check that this
is due to the `ipfp` algorithm improving the weights with each iteration,
let us analyse the aggregate results generated from the alternative
set of weights, generated with only 2 iterations of IPF:

```{r}
# Update ind_agg values, keeping col names (note '[]')
ind_agg[] <- t(apply(weights_maxit_2, MARGIN = 2, 
  FUN = function(x) colSums(x * ind_cat)))
ind_agg[1:2, 1:4]
```

Clearly the final weights after 2 iterations of IPF represent the constraint
variables well, but do not match perfectly except in the second constraint. This shows the importance of
considering number of iterations in the reweighting stage --- too many iterations
can be wasteful, too few may result in poor results. To reiterate,
20 iterations of IPF are sufficient in most cases for the results
to converge towards their final level of fit. 
More sophisticated ways of evaluating model fit are
presented in Section \ref{svalidation}.
As mentioned at the beginning of this chapter there are alternative methods
for allocating individuals to zones. These are discussed in a subsequent section
[](#alternative-reweighting)
.

For some applications, fractional weight matrices are sufficient for analysis.
Often, however, an individual level resulting dataset is required. To estimate the individual-level
variability in target variables such as income, or for agent-based modelling
applications, a full spatial microdataset, composed of whole individuals is
required. For this, we have to possibly. First, we can consider the weights as
probability and randomly chose the individuals in a distribution corresponding to the weights.
Secondly, we can consider the weights as the number of individual in this category and
the fractional weights must be integerised in a process
known as integerisation (Lovelace and Ballas, 2013). This is the subject of
the next section.

## Reweighting with **mipfp** {#mipfp}

```{r, echo=FALSE}
# Morgane: please make this example work with the SimplWorld data
# DONE 
```


The R package **mipfp** is a more generalized implementation of the IPF algorithm
than **ipfp**, and was designed for population synthesis.
**ipfp** generates a two-dimensional weight matrix based on mutually
exclusive (non cross-tabulated) constraint tables.
This is useful for applications using constraints, which are only marginals and 
which are not cross-tabulated.
**mipfp** is more flexible, allowing multiple cross-tabulations
in the constraint variables, such as age/sex and age/class combinations.

```{r, echo=FALSE}
# we have a new column with, at most, as many categories as the product of the number of categories for the variable age
# and 2 (Male or Female). However, in this case, we need to have the cross table of the age and the sex available to 
# proceed this way.
```

**mipfp** is therefore a multidimensional implementation of 
IPF, which can 
update an initial $N$-dimensional array with respect to given 
target marginal distributions. These, in turn, can be multidimensional. 
In this sense **mipfp** is more advanced than
**ipfp** which solves only the 2 dimensional case.

The main function of **mipfp** is `Ipfp()`, which
fills a $N$-dimensional weight 
matrix based on a range of aggregate-level constraint table options. 
Let's test the package on some example data.
The first step is to load the package into the workspace:

```{r}
library(mipfp) # after install.packages("mipfp")
```

To illustrate the use of `Ipfp`, we will create a fictive example. 
The case of SimpleWorld is here too simple to really illustrate the power 
of this package. The solving of SimpleWorld with `Ipfp` is included in 
the next section containing the comparison between the two packages.

The example case is as follows: to determine the contingency table 
of a population characterized by categorical variables for age (0-17, 18-50, 50+),
gender (male, female) and educational level (level 1 to level 4).
We consider a zone with 50 inhabitants. The classic spatial microsimulation 
problem consists in having all marginal distributions and the cross-tabulated
result (age/gender/education in this case) only for a non-geographical sample. 

We consider the variables in the following 
order: sex (1), age (2) and diploma (3).
Our constraints could for example be:

```{r}
sex <- c(23,27) # the number in each sex category
names(sex) <- c("Male","Female")

age <- c(16,20,14) # the number in each age category
names(age) <- c("Less18","18-50","More50")

diploma <- c(20,18,6,6) # the number in each education category
names(diploma) <- c("Level1","Level2","Level3","Level4")
```

We can verify that the population is equal in each constraint (50 people). Note also
the order in which each category is encoded --- a common source of
error in population synthesis.
For this reason we have labelled each category of each constraint.
The constraints are the target margins and need to be stored as a list.
To tell the algorithm 
which elements of the list correspond to which constraint,
a second list with the description 
of the target must be created. By printing *target* before to run the algorithm, 
we can verify that we have encoded everything well.

```{r}
target <- list (sex, age, diploma)
descript <- list (1, 2, 3)
target
```

Now that all constraint variables have been encoded, let us define 
the initial array to be updated, also referred as the seed or the weight matrix. The dimension
of this matrix must be identical to that of the constraint tables:
$(2 \times3 \times 4)$. Each cell of the array represents 
a combination of the attributes' values, and thus defines a particular 
category of individuals. In our case, we will consider that the weight
matrix contains 0 when the category is not possible and 1 otherwise. In this example we 
will assume that it is impossible for an individual being less than 18 
years old to hold a diploma level higher than 2. 
The corresponding cells are then set to 0, while the cells of the 
feasible categories are set to 1.

```{r}
names <- list (names(sex), names(age), names(diploma))
weight <- array (1, c(2,3,4), dimnames = names)
weight[, c("Less18"), c("Level3","Level4")] <- 0
```

Everything being well defined, we can execute
*Ipfp*. As with the package **ipfp**, we can chose a stoping criterion defined by *tol* and/or a maximum
number of iterations. Finally, setting the argument `print` to `TRUE` 
ensures we will see the evolution of the
of the procedure. After reaching either the maximum number of iterations or convergence 
(whichever comes first), the function will return a list containing the updated 
array, as well as other informations about the convergence of the algorithm. 
Note that if the target margins are not consistent, the input data is then 
normalised by considering probabilities instead of frequencies.

```{r}
result <- Ipfp(weight, descript, target, iter = 50, 
               print = TRUE, tol = 1e-5)
```

We can observe that the stoping criterion becomes always smaller and attains the *tol* after 8 
iterations. The `result`contains the final weight matrix
and some information about the convergence. 
We have a resulting table and we can validate the total number of 50 inhabitants 
in the zone. Note that, thanks to the definitions of names in the array, we can easily 
interpret the result. There are a total 50 people and nobody of less than 18 years old
own a diploma level 3 or 4, as desired.

```{r}
# printing the result
result$x.hat 
# checkin the total number of persons
sum(result$x.hat)
```

The quality of the margins with each constraints is contained in the variable `check.margins`
of the resulting list. In our case, we fit all constaints.

```{r}
# printing the resulting margins
result$check.margins
```

This reasoning works zone per zone and we can generate a 3-dimensional weight matrix. Another
advantage of **mipfp** is that it allows cross-tabulated constraints to be added.
In our example, we could add as target the contingency of the age and the diploma. 
We can define this cross table:

```{r}
# define the cross table
cross <- cbind(c(11,5,0,0), c(3,9,4,4), c(6,4,2,2))
rownames (cross) <- names (diploma)
colnames (cross) <- names(age)

# print the cross table
cross
```

When having several constraints concerning the same variable, we have to insure 
the consistency across these targets (otherwise convergence might not be reached). 
For instance, we can display the margins of `cross` along with `diploma` and `age`
to validate the consistency assumption. There are
both equal, meaning that the constraints are coherent.

```{r}
# check pertinence for diploma 
rowSums(cross)
diploma

# check pertinence for age 
colSums(cross)
age
```

The `target` and `descript` have to be updated to include the cross table.
Pay attention to the order
of the arguments by declaring a contingency table. Then, we can execute
the task and run *Ipfp* again:

```{r}
# defining the new target and associated descript
target <- list(sex, age, diploma, cross)
descript <- list(1, 2, 3, c(3,2))

# running the Ipfp function
result <- Ipfp(weight, descript, target, iter = 50, 
               print = TRUE, tol = 1e-5)
```

The addition of a constraint leave less freedom to the algorithm, implying that the algorithm 
converges faster. By printing the results, we see that all constraints are respected and
we have a 3-dimensional contingency table.

Similarly to the **ipfp** package, the result is in a non-integer array. Consequently 
we can generate a synthetic population either by considering the array cells 
as probabilities for random draws or by applying steps similar to 
the integerisation and expansion steps defined in the `ipfp` Section.

```{r, echo=FALSE}
# TODO: (MD) : put a real reference to the section of integerisation and expansion
```

Able to deal with bigger problems, the package presented here can also manage to 
solve smaller problem such as SimpleWorld. We have simply chosen to develop here a 
different example to show you the vast application of this function. However, 
to compare **ipfp** with **mipfp**, we will focus on this simpler example :
SimpleWorld.


### Comparing **ipfp** with **mipfp** {#Compareipf}

```{r, echo=FALSE}
# We need to establish which is faster (RL)
```

A major difference between the packages **ipfp** and **mipfp** is that the
former is written in C, while the latter is written in R. 
Another is that the *aim* of **mipfp** is to execute IPF for
spatial microsimulation, accepting a wide variety of inputs
(cross-table or marginal 
distributions). 
**ipfp**, by contrast, was created to solve an algebra problem of the form $Ax=b$.
Its aim is to find a vector $x$ such as $Ax=b$, after having predefined the vector
$b$ and the matrix $A$. Thus, the dimensions are fixed : a two-dimensional
matrix and two vectors. 
In addition, **mipfp** is written in R,
so should be easy to understand for R users. **ipfp**, by contrast, is written in pure C: fast but relatively difficult to understand.

As illustrated earlier in the book, **ipfp** gives weights to every
individual in the input microdata, for every zone. 
**mipfp** works differently. Individuals who have
the same caracteristics (in terms of the constraint variables)
appear only once in the input 'seed', their number represented by their
starting weight. **mipfp** thus determines the number of people 
in each unique combination of constraint variable categories.
This makes **mipfp** more *computationally efficient* than **ipfp**.
This efficiency can be explained for SimpleWorld. 
We have constraints about sex and age. What will be defined is the number
of persons in each category so that the contingency table respects the constraints.
In other worlds, we want, for zone 1 of SimpleWorld, to fill in the table such as:


Table: Aim of Mipfp for SimpleWorld.

|sex - age | 0-49 yrs| 50 + yrs| Total|
|:------|-----:|-----:|-----:|
|m      |     ?|     ?|     6|
|f      |     ?|     ?|     6|
|Total  |     8|     4|    12|

From this contingency table, and respecting the known marginal constraints,
the individual dataset must be generated.
For SimpleWorld, input microdata is available. While **ipfp** uses the
individual-level microdata directly, the 'seed' used in **mipfp** must also be
a contingency table. This table is then used to generate the final contingency table
(sex/age) thanks to the theory underlying IPF. 

For the first step, we begin with the aggregation of the microdata.
We have initially a total of 5 individuals and we want to have 12.
During the steps of IPF, we will always consider the current totals
and desired totals.

Table: Initial table of Mipfp for SimpleWorld.

|sex - age | 0-49 yrs| 50 + yrs| Total| Wanted total |
|:------|-----:|-----:|-----:|-----:|
|m      |     1|     2|     3|     6|
|f      |     1|     1|     2|     6|
|Total  |     2|     3|      |      |
|Target total  |     8|     4|      |      |


IPF makes a proportion of the existing table to fit each constraint
turn by turn. First, if we want to fit the age constraint, we need to totalise
8 persons under 50 years old and 4 above (wanted total of age).
Thus, we calculate a proportional 
weight matrix that respects this constraint. The first cell corresponds to 
$Current_Weight*\frac{Wanted_Total}{Current_Total}=1*\frac{8}{2}$. With a similar
reasoning, we can complete the whole matrix.

Table: First step of Mipfp for SimpleWorld - age constraint.

|sex - age | 0-49 yrs| 50 + yrs| Total| Wanted total |
|:------|-----:|-----:|-----:|-----:|
|m      |     4|     2.7|     6.7|     6|
|f      |     4|     1.3|     5.3|     6|
|Total  |     8|     4|      |      |
|Target total  |     8|     4|      |      |

Now, the current matrix fits exactly the age constraint ---
note how this was acheived ($1*\frac{8}{2}$ for the first cell).
However, there are still differences between the current and target sex totals.
IPF proceeds to make the table fit for the sex constraint:

Table: First step of Mipfp for SimpleWorld - sex constraint.

|sex - age | 0-49 yrs| 50 + yrs| Total| Wanted total |
|:------|-----:|-----:|-----:|-----:|
|m      |     3.6|     2.4|     6|     6|
|f      |     4.5|     1.5|     6|     6|
|Total  |     8.1|     3.9|      |      |
|Wanted total  |     8|     4|      |      |

This ends the first iteration of ipfp, illustrated 
in the form of contingency table. **mipfp** 
simply automates the logic, but by watching populations in terms of tables.
When the contingency table is complete,
completely, we know how many
individuals we need in each category. As with the **ipfp** method,
the weights are fractional.
To reach a final individual-level dataset, we can consider the weights as
probabilities
or transform the weight into integers and obtain a final contingency table. 
From this, it is easy to generate the individual dataset.
 
It should be clear from the above that both methods perform the same procedure.
The purpose of this section, beyond demonstrating the equivalence of **ipfp**
and **mipfp** approaches to population synthesis, is to compare the
*performance* of each approach. Using the simple example of SimpleWorld
we first compare the results before assessing the time taken by each package.

```{r, echo=FALSE}
# TODO: (MD) compare with simpleworld
#       (MD) if mipfp is better, justify why we use ipfp with cakemap... Or compare also with cakemap

# TODO (RL): cross reference image of CO vs reweighting approaches
```

The whole process using **ipfp** has been developed in the previous section.
Here, we describe how to solve the same SimpleWorld 
problem with **mipfp**. We proceed zone-by-zone, explaining in
detail the process for zone 1. The constraints for this area are as follows:

```{r}
# SimpleWorld constraints for zone 1
con_age[1,]
con_sex[1,]

# Save the constraints for zone 1
con_age_1 <- as.matrix( con_age[1,] )
con_sex_2 <- data.matrix( con_sex[1, c(2,1)] ) 
```

The inputs of `Ipfp` must be a list. We 
consider the age first and then the sex. Note that
the order of the sex categories must be the same as those of the
seed. Checking the order of the category is sensible as incorrect
ordering here is a common source of error: if the seed has
(Male, Female) and the constraint (Female, Male), 
the results will be wrong. 

```{r}
# Save the target and description for Ipfp
target <- list (con_age_1, con_sex_2)
descript <- list (1, 2)
```

The available sample establishes the seed of the algorithm. This is also known as 
the *initial weight matrix*. This represents the entire individual-level
input microdata (`ind`), converted into the same categories as in the constraints
(50 is still the break point for ages in SimpleWorld).
We also consider income, so that it is stored in the format required by **mipfp**.
(An alternative would be to ignore income here and
add it after, as seen in the expansion section.)

```{r}
# Add income to the categorised individual data
ind_full <- read.csv("data/SimpleWorld/ind-full.csv") # load full ind data
ind_income <- cbind(ind, ind_full[,c("income")])
weight_init <- table(ind_income[, c(2,3,4)])
```

Having the initial weight initialised to the aggregation of the 
individual level data and having the constraints well established, 
we can now execute the `Ipfp` function.

```{r}
# Perform Ipfp on SimpleWorld example
weight_mipfp <- Ipfp( weight_init, descript, target, 
                      print = TRUE, tol = 1e-5)
```

Convergence was acheived after only 5 iterations.
The result, stored in the `weight_mipfp` is a `data.frame` including
the final weight matrix, as well as information about the convergence
of the algorithm. We can check the margins and print the final 
weights. Note that we still have the income into the matrix, but 
comparison, we look at the result for the aggregate
table of age and sex.

```{r}
# Printing the resulting margins
weight_mipfp$check.margins

# Printing the resulting weight matrix for age and sex
apply(weight_mipfp$x.hat, c(1,2), sum)
```

We can see that we reach the convergence after 5 iterations
and that the results are fractional, as with **ipfp**. 
Now that the process is understood for one zone, we can progress to generate
weights for each zone. The aim is to transform the output
into a form similar to the one generated in the section **ipfp** 
to allow an easy comparison.
For each area, the initial weight matrix will be the same.
We store all weights in an array `Mipfp_Tab`, in
which each row represents a zone.

```{r}
# Initialising the result matrix
Names <- list(1:3, colnames(con_sex)[c(2,1)], colnames(con_age))
Mipfp_Tab <- array(data = NA, dim = c(nzones, 2 , 2), 
                   dimnames = Names)

# Loop on the zone and execute Ipfp
for (zone in 1:nzones){
  # Adapt the constraint to the zone
  con_age_1 <- data.matrix(con_age[zone,] )
  con_sex_2 <- data.matrix(con_sex[zone, c(2,1)] ) 
  target <- list(con_age_1, con_sex_2)
  
  # Calculate the weights
  res <- Ipfp(weight_init, descript, target, tol = 1e-5)
  
  # Complete the array of calculated weights
  Mipfp_Tab[zone,,] <- apply(res$x.hat,c(1,2),sum)
}
```

The above code executes IPF for each zone. 
We can print the result for zone 1:2 and observe that it correspond
to the weights previously determined.

```{r}
# Result for zone 1
Mipfp_Tab[1:2,,]
```

Thanks to the possibility of **mipfp** to include cross-tabulated
constraints, we could also consider the zone as a variable. This allows
the execution of one invocation of `Ipfp` (with no `for` loop)
to generate the result for every zone. 

```{r, echo=FALSE}
# (MD) : I am not sure this is very useful here, but
#        it is another alternative.
# RL yes this is very useful: please add an example!
#The code below illustrate this.
```

```{r}
Mipfp_in_one <- array(rep(weight_init, each = nzones), dim = c(nzones, 2, 2, 5))

target <- list(data.matrix(con_age), data.matrix(con_sex[,c(2,1)]))
descript <- list (c(1,2), c(1,3))
res <- Ipfp(Mipfp_in_one, descript, target, tol = 1e-5)
```

Note that the output of the function `Ipfp` can be used 
directly to perform steps such as integerisation and expansion
or as probability. 
However, here we first verify that both algorithms
calculate the same weights. For this we must transform the output of
`Ipfp` match the weight matrix generated by `ipfp` from the **ipfp** package. 
The comparison table will be called 
`weights_mipfp`. To complete its cells,
we will need the number of individuals in the sample in each category.

```{r}
# Initialise the matrix 
weights_mipfp <- matrix(data = NA, nrow = nrow(ind), 
                        ncol = nrow(cons))

# Cross-tabulated contingency table of the microdata, ind
Ind_Tab <- table(ind[,c(2,3)])
```

Now, we can fill this matrix `weights_mipfp`.
For each zone, the weight of the category is 
distributed between the individuals of the microdata
having the right characteristics. This is done thanks to iterations that calculate
the weights per category for the zone and then transform them into weights
per individual, depending on the category of each person.

```{r}
# Loop over the zones to transform the result's structure
for (zone in 1:nzones){
  
  # Transformation into individual weights
  for (i in 1:NbIndiv) {
    # weight of the category
    weight_ind <- Mipfp_Tab[zone,ind[i,c(2)],ind[i,c(3)]]
    
    # number of ind in the category
    sample_ind <- Ind_Tab[ind[i,c(2)],ind[i,c(3)]]
    
    # distribute the weight to the ind of this category
    weights_mipfp[i,zone] <- weight_ind / sample_ind
  }
}
```

The results from **mipfp** are now comparable to those from **ipfp**, 
since we generated weights for each individual in each zone. Having
the same structure, the integerisation and expansion step explained for
**ipfp** could
be followed. However, there is a more direct way to transform the
weights of `Ipfp` into individual data, which is explained in the 
chapter *Population synthesis without input microdata*.

```{r, echo=FALSE}
# This sounds good! (RL)
```

The below code demonstrates that the largest difference is of the order
$10^{-7}$, which is neglible. Thus, we have demonstrated that
both functions generate the same result.

```{r}
# Difference between weights of ipfp and mipfp
abs(weights-weights_mipfp)
```

One package is written in C and the other in R. The current question
is : Which package is more efficient? The answer
will depend on your application. We have seen that both 
algorithms do not consider the same form of input. For **ipfp**
you need constraints and an initial individual level data, coming for example from 
a survey. For **mipfp** the constraints and a contingency table of 
a survey is enough. Thus, in the absence of microdata, 
**mipfp** must be used (described in a subsequent chapter). Moreover, 
their results are in different structures. Weights are created for each 
individual with **ipfp** and weights for the contingency table with **mipfp**.
The results can be transformed from one to the other form if needs be.

### Speed testing

This section compares the computational time of **mipfp** and **ipfp** approaches. 

```{r, eval=FALSE}
# Measure time ipfp
init <- rep(1, NbIndiv)
system.time(ipfp(cons[1,], ind_catt, x0 = init,  tol = 1e-5))
```

The above code measure the time taken to execute the `ipfp` function for the 
zone 1. It takes 0.001 second on a modern computer^[The tests were performed on a
computer with Intel i7-4930K processer running at 3.40GHz.]
The code below make a similar analysis
for the `Ipfp` function and takes 0.002 second on the same computer.
Thus, for this very little example, the package **ipfp** written in C
is faster. The first lines of code below are defined to ensure that
we perform the same calculation. Thus, the times are comparable.

```{r, eval=FALSE}
# Chose correct constraint for mipfp
con_age_1 <- data.matrix( con_age[1,] )
con_sex_2 <- data.matrix( con_sex[1, c(2,1)] ) 
target <- list (con_age_1, con_sex_2)
descript <- list (1, 2)

# Equitable calculus - remove income 
weight_init_no_income <- table( ind[, c(2,3)])

# Measure time mipfp
system.time(Ipfp( weight_init_no_income, descript, target, 
                  tol = 1e-5))
```

An advantage of the **mipfp** package, for problems with a lot 
of input individuals, is in terms of memory. Indeed, **ipfp**
needs a table with as many rows as individuals, whereas **mipfp**
needs a table of a dimension corresponding to the number of 
different categories. For SimpleWorld, the inputs of **ipfp** 
and **mipfp** are respectively:

```{r}
# input for ipfp
ind_cat

# input for mipfp
Ind_Tab
```

Imagine that we have a similar problem, but with a total 
population of 2000 and 500 initial individuals. The dimension 
of `ind_cat` will become 500 rows for 4 columns, where the
one of `Ind_Tab` will not change. 

By testing the time for SimpleWorld when considering having 25,000
individuals (we replicate the data 5000 times) and generating 
12,000,000 individuals (we multiply the constraints by 1000), we obtain
on the same computer that **ipfp** and **mipfp** take, respectively,
0.012 and 0.005 seconds.

```{r, echo=FALSE, eval=FALSE}
#### time tests for bigger databases

ind_cat2 <- ind_cat

for (i in 1:4999){
  ind_cat2<-rbind(ind_cat2,ind_cat)
}

Ind_Tab2 <- 5000 * Ind_Tab

cons2 <- cons * 1000000

# Measure time ipfp
system.time(ipfp(cons2[1,], t(ind_cat2), x0 = rep(1,NbIndiv * 5000),  tol = 1e-5))

# Chose right constraint for mipfp
target <- list (data.matrix(con_age[zone,] * 1000000), data.matrix(con_sex[zone,]  * 1000000)[c(2,1)])

# Equitable calculus - remove income 
weight_init_no_income <- table( ind[, c(2,3)])*5000

# Measure time mipfp
system.time(Ipfp( weight_init_no_income, descript, target, tol = 1e-5))
```

In conclusion to this comparison, **mipfp** is more adaptable
than **ipfp**. Which to use will depend on the structure
of the data available. In terms of computational time,
the comparison demonstrates that **mipfp** was slower
for small problems and faster for bigger problems. 
In all cases, they calculate the same results.

```{r, echo=FALSE, eval=FALSE}
save.image(file = "cache-smsim-in-R.RData")
```

