---
title: "Spatial microsimulation for agent-based models"
layout: default
---

```{r, echo=FALSE}
# Resources:
# ABM www.geog.leeds.ac.uk/courses/level3/geog3150/

# Java: http://www.geog.leeds.ac.uk/courses/other/crime/
# Java about half way down ABM "details" page:
# http://www.geog.leeds.ac.uk/courses/other/crime/abm/general-modelling/index.html
```

# Spatial microdata in agent-based models {#ABM}

Agent-based models (ABMs) of human systems require information about the 
individual people that the agents will represent.
One way to initialise an ABM, is through the use of spatial microsimulation

Agent-based models involve three things:

1. a number of discrete agents, who have
2. different characteristics and which
3. interact with each other and their environment.

ABM is a powerful and extremely flexible approach to modelling.
ABM allows analysis of problems that are highly non-linear and *emergent*:
the dominant processes guiding change can only be seen after the event.
This is
advantageous, allowing a very wide range of problems to be tackled through
a single paradigm: "agent-based modeling can find new, better solutions to many
problems important to our environment, health, and economy" (Grimm and
Railsback, 2011). With this power comes danger: it is easy to get lost in
agent-based modelling and arrive at incorrect conclusions. Using spatial
microsimulation to generate realistic input populations is one way that
ABM can be made to more closely represent the population under study.

Note that based on the above definition we
already have two-thirds of the necessary ingredients
of an ABM: the synthetic spatial microdata created in previous chapters can,
with the right tools, skills and understanding, be used almsot directly as an
input into ABM. If your aim is to use spatial microdata as an input into
agent based models, you're more than half way there!

Before describing the transition from the
spatial microdata to ABM based on a simple example (you guessed it --- SimpleWorld),
it is worth mentioning some applications of ABMs built on spatial microdata
and some of the tools that will be useful for the job.

## ABM applications



## ABM software

A wide range of software is available for ABM. Of these, **NetLogo**
is one of the most
widely used for model prototyping, development and communication 
(Thiele et al. 2012). Alternative open source options include
the Java-based ABM **Repast** (for power users) and **MASON** (which must
be programmed in Java). Due to its excellent documentation, ease of learning
and integration with R, NetLogo is the tool of choice in this book and should
be sufficiently powerful for many applications.

[NetLogo](https://ccl.northwestern.edu/netlogo/) is a
mature and widely used toolkit for agent-based models. It is written in
Java, but uses a derivation of the Logo language to develop models 
(most users will be blissfully unaware of Java).
The recently published **RNetLogo** package provides an interface between
R and NetLogo, allowing for model runs to be set-up and run directly from within R
(Theile 2014). 
This allows the outputs of agent based models to be loaded directly into
your R environment. Using R to run a separate programme may seem overly complicated
for very simply models. For setting-up and testing we recommend using 
ABM, with its intuitive graphical interface. However, there are many advantages
of using NetLogo from within R, the foremost being R's analysis capabilities:
"it is much more time consuming and complicated to analyse ABMs than to
formulate and implement them" (Theile et al. 2012).

```{r, echo=FALSE}
# Note above:
#"For setting-up and testing we recommend using  ABM,"
# Do you mean 'AMB' here? Sentence doesn't make sense to me.
```

```{r,echo=FALSE}
# TODO: Batty reference below on abm design
```

AMB's are inherently unpredictable due to their 'bottom up' design.
Thus we generally want to study many model runs before drawing conclusions
about how the overall system operates, let alone real world implications.
Because much of the
time taken for agent based modelling is consumed on this sensitivity/parameter space
analysis, running NetLogo
from within R makes sense. R excels in these areas whereas NetLogo
and other agent-based modelling programs do not.

```{r, echo=FALSE}

# 
# Watch this space for more on this!
# 
# https://github.com/Robinlovelace/spatial-microsim-book
```

## SimpleWorld in NetLogo

## Allocating attributes to agents

## Cops and robbers in SimpleWorld
