---
title: "What is spatial microsimulation?"
layout: default
bibliography: ~/Documents/smr.bib
---

```{r, echo=FALSE}
# This is an early draft of a book on spatial microsimulation, for teaching purposes
# Introduction {#Introduction}
library(png)
library(grid)
```

# What is spatial microsimulation? {#what-is}

Spatial microsimulation is a technique for generating data statistically as similar as 
possible to aggregate-level observations and with a disaggregate spatial information. Its most common use in the social sciences is for allocating individuals from a survey dataset
to administrative zones, based on shared variables between the areal and
individual level data. Spatial microsimulation can also be used to simulate the values of missing characteristics for individuals in a zone, information that is not otherwise available.
As with many new and infrequently used phrases, this
understanding is not shared by everyone. The meaning of spatial
microsimulation varies depending on context and who you ask. To an
economist, spatial microsimulation is likely to imply
modelling some kind of temporal process such as how individual agents in
different areas respond to changes in prices or policies. To a transport
planner, the term may imply simulating the precise movements of vehicles on
the transport network. To your next door neighbour it may mean you have
started speaking gobbledygook! Hence the need to consider what spatial
microsimulation is, and what it is not, at the outset. However, every case, 
the term involves the creation individual-level data that is grouped by geographic zone via some kind of approximation method.

## Terminology {#terminology}

Delving a little into the etymology and history of the term reveals the
reasons behind this duplicity of meaning and highlights the importance
of terminology. Rarely will you be understood saying
“I use *spatial microsimulation*” in everyday life. Usually it
is important to add context. Below are a few hypothetical situations and
suggested responses.

-   When talking to a colleague, a transport modeller: “spatial
    microsimulation, also known as population synthesis...”

-   Speaking to agent based modellers: “we use spatial microsimulation
    to simulate the characteristics of geo-referenced agents...”

-   Communicating with undergraduates who are unlikely to have come
    across the term or its analogies. “I do spatial microsimulation, a
    way of generating individual-level data for small areas...”

-   Chatting casually in the pub or coffee shop: “I’m using a technique
    called spatial microsimulation to model people...”.

The above examples illustrate potential for
confusion: care needs to be taken to use terminology each audience understands. The transport modeller, for example, may
already know that the term 'population synthesis' means creating an individual-level
dataset of real areas, whereas more basic terms need to be used when communicating
the method to policy makers. All this links back to the importance
of transparency and reproducibility of method discussed in the previous chapter:
avoid implying spatial microsimulation is something it is not.

Faced with uncomprehending stares when describing the method, some may
be tempted to ‘blind them with science’, relying on
sophisticated-sounding jargon, for example by saying: “we use simulated
annealing in our integerised spatial microsimulation model”. Such
wording obscures meaning (how many people in the room will understand
‘integerised’, let alone ‘simulated annealing’) and makes the process
inaccessible. Although much jargon is used in the spatial
microsimulation literature and in this book, care must be taken to ensure
that people understand what you are saying.

This raises the question, why use the term spatial microsimulation at
all, if it is understood by so few people? The answer to this is that
spatial microsimulation, defined clearly at the outset and used
correctly, can concisely describe a technique that would otherwise need
many more words on each use. Try replacing ‘spatial microsimulation’
with ‘a statistical technique to allocate individuals from a survey
dataset to administrative zones’ each time it appears in this book and
the advantages of a simple term should become clear! ‘Population
synthesis’ is perhaps a more accurate term but, transport modelling
aside, the literature already uses ‘spatial microsimulation’. Rather
than create more complexity with *another* piece of jargon, we proceed
with the term favoured by the majority of practitioners.

Why has this situation, in which practitioners of a statistical method
must tread carefully to avoid confusing their audience, come about?
First it’s worth stating that the problem is by no means unique to this
field: imagine the difficulties that Bayesian statisticians must
encounter when speaking of prior and posterior probability distributions
to an uninitiated audience. Let alone describing Gibb’s sampling. Describing complex methods is 
a constant challenge for the scientist. Using jargon when the audience 
is experienced and simplifying when they are not is the art of science communication.

If we look into the terminology "spatial microsimulation", we can intuitively 
detect three pieces of information. First, it is a "simulation", meaning that
we can not reach the real needed data, so we make an approximation helped 
with available data. Second, "micro" reflects the desire to simulate
the data at the individual level. This captures the ability of spatial microsimulation to 'zoom in' from aggregated data to a higher resolution data where all the characteristics of all individuals are simulated.
- Finally, "spatial" indicates that
each individual is allocated to a zone, that depending on the context
can correspond to a postcode, a municipality, a district, a country etc. Usually these zones are geographical.

```{r, echo=FALSE}
# # Commented out for now as it risks confusing the reader
# In each application of spatial microsimulation we refer to a population to be simulated. For example, a population can be a set
# of bulbs, if we analyse the quality of a new production process of bulbs. 
# And their characteristics could be their life duration and their power. 
# Moreover, we need the data in an individual level and the location of each
# of these entities. In our example, we want to know that the first bulb
# has a power of 60Watt, a life duration of 200 hours and was created in 
# the firm of Texas.
```

To avoid confusion regarding the terminology used in this book,
a glossary defining much of the jargon relating to spatial microsimulation
is provided at the end. For now, to
help answer what spatial microsimulation is we will look at 
its applications and then at what it is not.

## Applications {#applications}

Spatial microsimulation has a wide variety of applications and there
are many areas where the technique has been used.
Some of the main areas of application have
been health, economic policy evaluation and transport. Rather than attempt
to provide a comprehensive account of the range of current and possible
applications, this section describes a single study in each area to exemplify
how spatial microsimulation is used.

### Health applications

An excellent example of the potential practical utility of spatial microsimulation
is a study
which estimated the rate of smoking at the
small area level in the city of Leeds UK
(Tomintz et al. 2008).
Smoking is a classic 'target variable' in spatial microsimulation:
it is reported in a number of individual-level surveys but there is surprisingly
little information about how smoking rates vary from place to place.
Thus it is difficult to where to locate services that depend on the rate of smoking.
The synthetic spatial microdata could thus be used to help identify new clinics
to help people stop smoking. (Alternatively, the spatial microdata could be used
by a tobacco chain to help decide where to invest in a new shop, highlighting
the potential misuse of the technique by unscrupulous analysts.) The authors
found that actual anti-smoking clinics were not located optimally. Furthermore,
the results pointed to optimal locations for new clinics, potentially improving
the cost-effectiveness of public health campaigns.

This research has since been 'scaled-up' to estimate smoking rates across the whole
of Austria. The [simSALUD](http://www.simsalud.org/) portal provides users with
access to the resulting spatial microdata and an on-line interface to allow
for the selection of constraint variables and other options to customise the
model for the specific purposes. This portal-based system and the provision
of synthetic spatial microdata to researchers illustrates one possible direction that
spatial microsimulation research could go in, where the synthetic data produced from
a large model is the main output of the research, to be used by others for a variety of
applications.

The example of smoking demonstrates the increase spatial resolution
that spatial microsimulation can bring to bear on under-studied areas in public
health. Where the prevalence of unhealthy activities is closely related 
to socio-demographic variables, a synthetic microdataset can lead to decision
making tools that would be difficult to implement with non-spatial surveys alone.
**Simobesity** is another research project and spatial microsimulation software tool that
estimates the prevalence of obesity at the local levels depending on demographic
constraint variables
(Edwards and Clarke, 2013)
.
Recent evidence has emerged on the impact of
car-dependent urban environments on inactive lifestyles and resulting poor health
(these areas have been labelled 'obesogenic'). In this context, there is
great potential for combining socio-demographic and environmental-geographic
variables in a spatial microsimulation model. Using the same principles
described in Tomintz et al. (2008), the outputs of such a model
could help target local interventions
to tackle physical inactivity, to maximise the benefits of public health funds.

```{r, echo=FALSE}
# TODO: link the reader here to a chapter where smsim models are linked to geo-data
```

### Economic policy evaluation

Microsimulation was originally developed for economic policy evaluation and it   
is still one of most common applications. 'Social impact evaluation', where the
impact of policy changes on different income and socio-demographic groups is
explored, is a classic example of applied microsimulation research. Frequently
these simulations are undertaken by governmental departments
for entire countries and focus on overall shifts in the population rather than
spatial variability in the impacts. The EU funded EUROMOD project
and software package of the same name
is the largest of these 
initiatives. The EUROMOD software is used by government analysts and
research agencies in many countries to estimate the distributional impacts
of policy reforms. In a recent example,
EUROMOD was used to assess
effects of austerity on different income groups in the UK
(De Agostini et al. 2014).
The finding that recent policies have been highly regressive (Figure 2.1)
shows that
microsimulation can provide a new evidence base in policy relevant areas.

```{r vingtile, fig.cap="Output from the EUROMOD economic microsimulation model (Avram et al. 2014). Along the x axes is income group rising to the right. This means, for example, that Latvia (LV) has implemented progressive policies whereas Portugal (PT) has implemented regressive policies. Country acronyms from left to right stand for Estonia (EE), Greece (EL), Spain (ES), Italy (IT), Latvia (LV), Lithuania (LT), Portugal (PT), Romania (RO) and the UK.", fig.width=5, fig.height=5, echo=FALSE}
# A typical output from a microsimulation model for economic policy evaluation, from De Agostini et al. (2014). Each vertical bar represents the relative gains or losses of each income 'vingtile' (there are 20 equally sized groups, by income, in the model) and the black line shows that the austerity budget has been most detrimental to low income groups and most beneficial to the upper-middle classes.
library(png)
library(grid)
img <- readPNG("figures/austerity.png")
grid.raster(img)
```

Spatial microsimulation uses very similar techniques as those employed by
EUROMOD and other economic microsimulation models, including probability-weighted random sampling of individual-level
data and aggregate-level scenario development
(Sutherland and Figari, 2013). The majority of microsimulation
research for economic policy evaluation does not
disagreggate the impacts over space, however. The estimation of variability
at the *local level* is what differentiates spatial microsimulation models
from economic microsimulation models, although the underlying methods are
very similar.

### Transport

Transport modelling is a mature field that increasingly uses individual-level
data as the basis of analysis. Large scale models
such as [MATSim](http://www.matsim.org/) rely on spatial microdata to provide
demand for travel and individual characteristics for origins and destinations. Although the techniques for generating spatial microdata in the transport literature, the process is referred to as *population synthesis*. 

Generally, little attention is paid to this process of
synthetic population generation in transport modelling
because the focus is on movement of individuals rather
than their characteristics. Distributional impacts
are often overlooked in transport models (Lucas 2012)
and there is much potential to integrate spatial microsimulation
with existing transport modelling methods.

An example of the potential uses of spatial microsimulation
in transport models is illustrated in Figure 2.2. This
shows the simulated commuting behaviour of 20 randomly
selected individuals from a large scale spatial microdataset
of Sheffield. Because the constraints used in this model
included socio-demographic variables, each individual
represented in the figure has a rich profile of characteristics
associated with them. This analysis can provide new evidence
about the likely winners and loosers from very specific
interventions such as a new bicycle path or bus route
(Lovelace et al. 2014). As a result of the increased detail
allowed by such methods there is much interest in spatial
microsimulation for transport applications. Figure 2.2 also
illustrates the potential for the output of 
spatial microsimulation to be used as an input into agent-based
models (ABM).

```{r, echo=FALSE}
# r transgeo, fig.cap="An illustration of spatial microdata in transport modelling. 20 people are illustrated on the map as travelling to a range of destinations, specified based on probability-weighted sampling from origin-destination tables (Lovelace et al. 2014).", fig.width=4, fig.height=4, echo=FALSE}
# library(png)
# library(grid)
# img <- readPNG("figures/jtg.png")
# grid.raster(img)
```

## What spatial microsimulation is not {#is-not}

**Spatial microsimulation is not strictly spatial**

The most surprising feature of spatial microsimulation
is that the method is not strictly *spatial*. The only
reason why the method has developed this name (as opposed to 'small
area population synthesis', for example) is that practitioners tend
to use administrative zones, which represent geographical areas, as the
grouping variable. However, any mutually exclusive grouping variable,
such as age band or number of bedrooms in your house, could
be used. Likewise, geographical location can be used as a *constraint variable*.
In most spatial microsimulation models, the spatial variable is a mutually
exclusive grouping, interchangeable with any such group. Of course, the spatial
microdata, maps and analysis that result from spatial microsimulation are spatial,
it's just that there is nothing inherently spatial about the method used to
generate the spatial microdata.

To be more precise, spatial microsimulation is not *inherently spatial*.
Spatial attributes such as the geographic coordinates of the zones
to which individuals have been allocated and home and work
locations can easily be added to the spatial microdata after they have been
generated. It is the use of geographical variables as the grouping variable that is
critical here and which distinguishes spatial microsimulation from other types
of microsimulation.

**Spatial microsimulation is not agent-based modelling (ABM).**

Spatial microsimulation does involve the creation and analysis of individuals,
but it does not imply any interaction between these individuals. For this, an
agent-based model (ABM) is needed. It would be easy to assume that because
the method contains the word 'simulation', it includes modelling of individual
behaviours over time and space. This is not the case.

Spatial microsimulation as an approach to modelling involves
the allocation of individuals to specific zones, what-if scenarios, and, in *dynamic spatial microsimulation*, a time
step to represent the individuals' evolution over time. In *static spatial microsimulation* studies that rely on *what-if* scenarios, time is not 
really included. 

Agent based modelling (ABM), by contrast, implies some kind of interaction between the individual units
and time is an integral part of the process. The aims of the two methods are complementary.
ABM requires an initial population which spatial microsimulation can provide and adds more complexity and interactivity. The rules of the ABM make each agent evolve through time and space 
depending in a bottom-up process. Spatial microsimulation is, by contrast, more of a *top down* process. Spatial microsimulation can be 
closely linked to ABM, however. As described in Chapter \ref{ABM}, the outputs of spatial 
microsimulation form an excellent basis for ABMs.

**Spatial microsimulation does not generate new data**

During spatial microsimulation, apparently new individuals are created and
placed into zones. It would be tempting to think that new information
about the world is somehow
being created. This is not the case: the 'new' individuals are simply repeats
of individuals we already knew about from the individual-level data, albeit
in a different order and in different combinations. If the population of the
study area is greater than the sample size of the input data, many individuals
will have to be 'cloned'. Thus we are not increasing the diversity of the dataset,
simply changing its aggregate-level characteristics.

## Spatial microsimulation in this book {#MSimThisBook}

As we have seen, the term "spatial microsimulation" has a number of meanings incorporating a range of methods and applications. In this book, the focus is on the
most used part of this branch, called reweighting: the process of generating the spatial microdata. From [](#vis) onwards, we also explain what can be done with the spatial microdata that we have generated in the preceding chapters. The technique described in this book will be most useful in contexts where you
have categorical geographically aggregated count data about the entire target population (these are the *contraint variables*, often taken from a national census) and an individual-level survey based on a sample of the target population (the non-spatial *microdata*).

## Assumptions {#assumptions}

As with any simulation technique, spatial microsimulation is based on
assumptions, some of which are unlikely to hold in all cases. This
does not preclude spatial microsimulation in cases where the assumptions
do not hold: "It is far better to foresee even without
certainty than not to foresee at all", as Henri Poincaré put it
(Barthélemy 2014).

```{r, echo=FALSE}
# (Barthélemy, 2014).
```


It is vital, however, that users of spatial microsimulation and
'consumers' of the resulting research understand that the results of spatial
microsimulation are not *real* but a best estimate of the population in a
given area. The danger is that if the assumptions are not understood, incorrect
conclusions will result. It is therefore the duty of researchers using spatial
microsimulation (and other techniques) to clearly state the assumptions on which
the results depend on and the extent to which these assumptions can be expected
to hold in practice. Roughly speaking there are four main assumptions underlying
all spatial microsimulation models:

1. The individual-level microdata are representative of the study area.
2. The target variable is dependent on the constraint variables and their
interactions in a way that is relatively constant over space and time.
3. The relationships between the constraint variables are not spatially dependent.
4. The input microdataset and constraints are sufficiently rich and detailed
to reproduce the full diversity of individuals and areas in the study region.

Obviously the real world is complex and many processes are spatially dependent,
invalidating assumptions 2 and 3. The extent to which the relationships between
variables can be deemed to be constant over space is often
unknown. However, there there are ways of
checking the spatial dependency of
relationships between multiply variables, not least Geographically Weighted Regression
(GWR). 

These limitations should be discussed at the outset of spatial
microsimulation research, with reference to the input data. To see how spatial
microsimulation simplifies the real world, the next chapter describes
a hypothetical scenario where 33 inhabitants of an imaginary land
are simulated and allocated to three zones
based on a microdataset of only 5 individuals and two constraint variables.
