---
title: "Additional tools and techniques"
layout: default
bibliography: bibliography.bib
---

# Additional tools and techniques {#additional}

```{r, echo=FALSE}

```

## R packages for spatial microsimulation

Most of the code presented so far --- 
with the exception of plotting commands based on **ggplot2**,
data manipulation tools from **dplyr** and some packages offering
niche functionality such as **ipfp** and **RNetLogo** --- has been implemented in R's base functions.
This is deliberate: base R offers robustness of
code and minimisation of installation dependencies.
There are often dozens of ways of doing one thing
in R, and a bewildering number of packages offering to help.
The way that will be understood by the greatest number
of people and work on the maximum number of computers
is often preferable, however, unless there is a
clear advantage to using additional packages. 
The behaviour of contributed packages
(excepting the `r-recommended` [packages](see http://cran.r-project.org/bin/linux/debian/README))
may change unexpectedly,
whereas core R functions are likely to remain stable over many decades to come.

With this caveat in mind, let's begin our tour of
some contributed R that are of use for
spatial microsimulation.

### **GREGWT**

### **sms**

### **multilevel**

## The Flexible Modelling Framework (FMF)

## Allocation of home-work locations

A question that arises after spatial microdata have been allocated to
geographical zones is: where exactly in the zone do the people inhabit?
For some applications this may not matter but for others, such as disaster
management and
or transport planning, the precise location of an individual
is important [@Smith2014; @Hanaoka2014].

```{r, echo=FALSE}
# TODO: add reference for above
```

In this section we will see how spatial microdata can be allocated first
to 'urban areas' (to prevent people being placed in the sea, for example)
and then to individual buildings based on freely available Open Street Map
data. This process also makes sense from the perspective of visualisation:
the typical choropleth map outputs of spatial microsimulation models
over-represent low density rural areas and under-represent dense
urban areas in terms of visual 'real-estate'. Plotting attributes only
in the buildings where they occur can help overcome this issue.

```{r, echo=FALSE}
# TODO: add figure from O'Brien's online map
```

## Spatial interaction modelling

In this example we will demonstrate a method for
evaluating the distributional impact of a new pathway
in an urban setting. 

```{r}
## Advanced applications in agent-based modelling
```

